### 2018-04-12 docs/lamachine_clarin2018.tex
starting paper

\documentclass[a4paper,11pt]{article}
\usepackage{CLARIN2018}
% - - - - - - - IMPORTANT - - - - - - - -
% The next three lines allow XeLaTeX, graphics import and hyperlinks, set font and language
\usepackage{xltxtra,polyglossia,graphicx,hyperref}
\setmainfont[Mapping=tex-text]{Times}
\setdefaultlanguage{english}
% If for some reason the above three lines are not compatible with your LaTeX installation,
% comment out the above three instructions and uncomment the following four instead:
%\usepackage{times}
%\usepackage{url}
%\usepackage{latexsym}
%\usepackage[english]{babel}

%\setlength\titlebox{5cm}

% You can expand the titlebox if you need extra space
% to show all the authors. Please do not make the titlebox
% smaller than 5cm (the original size); we will check this
% in the camera-ready version and ask you to change it back.

\usepackage{covington} % if needed, for linguistic examples

\title{LaMachine NLP Software Distribution}

% - - - - - - - IMPORTANT - - - - - - -
% Leave the author information empty until your paper has been accepted

% Uncomment the following line ONLY if you need two author rows
%\setlength\titlebox{80mm}

\author{Maarten van Gompel \\
  Centre for Language and Speech Technology \\
  Radboud University Nijmegen, the Netherlands \\
  {\tt proycon@anaproy.nl} \\ %\And % if needed: this makes a second column
%  Second Author \\
%  Department (optional)\\
%  University Name without city \\
%  City, Country \\
% {\tt email@domain} \\
% \AND % if needed: this makes a second row
%  Third Author \\
%  Department (optional)\\
%  University of City, Country \\
%  {\tt email@domain} \\\And
%  Fourth Author \\
%  Department (optional)\\
%  University Name without city \\
%  City, Country \\
%  {\tt email@domain} \\
}

\date{}

\begin{document}
\maketitle

\begin{abstract}
We introduce LaMachine, a unified Natural Language Processing (NLP) open-source
software distribution to facilitate the installation and deployment of a large
amount of software projects that have been developed in the scope of the
CLARIN-NL project and its current successor CLARIAH. Special attention is paid
to encouragement of good software development practises.
\end{abstract}

\section{Introduction} \label{intro}

Software is a key deliverable and a vital component for research in projects
such as CLARIN. It is software that provides researchers the instruments to
yield for their research, and without it a lot of research would become highly
unfeasible or right-out impossible. It is CLARIN's core mission to make digital
language resources, including software, available to the wider research
community.

Software is a broad but well-ingrained notion in society nowadays, referring to
any form of computer program. This can manifest in various forms, whether it is
an app on a phone, a graphical desktop application, a command line tool,  or a
webpage.

one classification could be
made based on the \emph{interfaces} software offers.

We too often see that the research community develops valuable

### 2018-04-12 bootstrap.sh
no separate web_user/web_group on mac

@@ -1010,7 +1010,7 @@ echo "http_port: 80 #webserver port (for VM or docker)
 mapped_http_port: 8080 #mapped webserver port on host system (for VM or docker)
 services: [ $SERVICES ]  #List of services to provide, if set to [ all ], all possible services from the software categories you install will be provided. You can remove this and list specific services you want to enable. This is especially needed in case of a LaMachine installation that intends to only provide a single service.
 " >> $STAGEDCONFIG
-    if [[ $FLAVOUR == "local" ]]; then
+    if [[ $FLAVOUR == "local" ]] || [[ "$OS" == "mac" ]]; then
         echo "web_user: \"$USERNAME\"" >> $STAGEDCONFIG
         echo "web_group: \"$GROUP\"" >> $STAGEDCONFIG
     else

### 2018-04-12 bootstrap.sh
produce error when homebrew install fails

@@ -683,11 +683,11 @@ for package in ${NEED[@]}; do
                 * ) echo "Please answer yes or no.";;
             esac
         done
-        /usr/bin/ruby -e "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/master/install)"
-        brew tap caskroom/cask
+        /usr/bin/ruby -e "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/master/install)" || fatalerror "Homebrew installation failed!"
+        brew tap caskroom/cask || fatalerror "Failed to install brew-cask, ran: brew tap caskroom/cask"
     elif [ "$package" = "brew-cask" ]; then
         echo "Installing brew-cask"
-        brew tap caskroom/cask
+        brew tap caskroom/cask || fatalerror "Failed to install brew-cask, ran: brew tap caskroom/cask"
     elif [ "$package" = "git" ]; then
         if [ "$OS" = "debian" ]; then
             cmd="sudo apt-get $NONINTERACTIVEFLAGS install git-core"

### 2018-04-12 bootstrap.sh
Extra anaconda check (anaconda refuses to play nice with virtualenv so can't be used for bootstrap)

@@ -108,6 +108,17 @@ if [ ! -z "$CONDA_PREFIX" ]; then
     fatalerror "Inception error: Do not run the LaMachine bootstrap when you are inside an Anaconda environment (run 'source deactivate' first)"
 fi
 
+if which python; then
+    echo "Checking sanity of your Python installation..."
+    python -c "from __future__ import print_function; import sys; print(sys.version)" | grep -i anaconda
+    if [ $? -eq 0 ]; then
+        fatalerror "Conflict error: The default Python on this system is managed by Anaconda, this is incompatible with LaMachine. Ensure the Python version in your \$PATH corresponds to a regular version as supplied with your OS, editing the order of your \$PATH in ~/.bashrc or ~/.bash_profile might be sufficient to solve this without completely uninstalling anaconda"
+    fi
+else
+    fatalerror "No Python found! However, python should be available by default on all supported platforms; please install it yourself through your package manager (and ensure it is in your \$PATH)"
+fi
+echo ""
+
 ####################################################
 #               Platform Detection
 ####################################################

### 2018-04-12 bootstrap.sh
version bump

@@ -20,7 +20,7 @@ boldblue=${bold}$(tput setaf 4) #  blue
 normal=$(tput sgr0)
 
 echo "${bold}=====================================================================${normal}"
-echo "           ,              ${bold}LaMachine v2.1.1${normal} - NLP Software distribution"
+echo "           ,              ${bold}LaMachine v2.1.2${normal} - NLP Software distribution"
 echo "          ~)                     (http://proycon.github.io/LaMachine)"
 echo "           (----í         Language Machines research group"
 echo "            /| |\         Centre of Language and Speech Technology"
@@ -112,7 +112,7 @@ if which python; then
     echo "Checking sanity of your Python installation..."
     python -c "from __future__ import print_function; import sys; print(sys.version)" | grep -i anaconda
     if [ $? -eq 0 ]; then
-        fatalerror "Conflict error: The default Python on this system is managed by Anaconda, this is incompatible with LaMachine. Ensure the Python version in your \$PATH corresponds to a regular version as supplied with your OS, editing the order of your \$PATH in ~/.bashrc or ~/.bash_profile might be sufficient to solve this without completely uninstalling anaconda"
+        fatalerror "Conflict error: The default Python on this system is managed by Anaconda, this is incompatible with LaMachine. Ensure the Python found in your \$PATH corresponds to a regular version as supplied with your OS, editing the order of your \$PATH in ~/.bashrc or ~/.bash_profile should be sufficient to solve this without completely uninstalling anaconda."
     fi
 else
     fatalerror "No Python found! However, python should be available by default on all supported platforms; please install it yourself through your package manager (and ensure it is in your \$PATH)"

### 2018-04-13 bootstrap.sh
added stackoverflow URL for more context

@@ -112,7 +112,7 @@ if which python; then
     echo "Checking sanity of your Python installation..."
     python -c "from __future__ import print_function; import sys; print(sys.version)" | grep -i anaconda
     if [ $? -eq 0 ]; then
-        fatalerror "Conflict error: The default Python on this system is managed by Anaconda, this is incompatible with LaMachine. Ensure the Python found in your \$PATH corresponds to a regular version as supplied with your OS, editing the order of your \$PATH in ~/.bashrc or ~/.bash_profile should be sufficient to solve this without completely uninstalling anaconda."
+        fatalerror "Conflict error: The default Python on this system is managed by Anaconda, this is incompatible with LaMachine. Ensure the Python found in your \$PATH corresponds to a regular version as supplied with your OS, editing the order of your \$PATH in ~/.bashrc or ~/.bash_profile should be sufficient to solve this without completely uninstalling anaconda. See also https://stackoverflow.com/a/37377981/3311445"
     fi
 else
     fatalerror "No Python found! However, python should be available by default on all supported platforms; please install it yourself through your package manager (and ensure it is in your \$PATH)"

### 2018-04-13 helpers/metadater.py
First prototype of metadater tool, with support for conversion from pip and JSON and YAML output #74

@@ -6,85 +6,284 @@
 #https://joinup.ec.europa.eu/svn/adms_foss/adms_sw_v1.00/adms_sw_v1.00.htm
 
 
+import sys
 import argparse
 import subprocess
 import yaml
+import json
+from collections import OrderedDict
 
-properties = {
+def represent_ordereddict(dumper, data):
+    value = []
+
+    for item_key, item_value in data.items():
+        node_key = dumper.represent_data(item_key)
+        node_value = dumper.represent_data(item_value)
+
+        value.append((node_key, node_value))
+
+    return yaml.nodes.MappingNode('tag:yaml.org,2002:map', value)
+
+yaml.add_representer(OrderedDict, represent_ordereddict)
+
+properties = { #first order properties (including collections, i.e. properties that may be specified multiple times!)
     "doap:name": "The name of the software",
     "doap:revision": "The full version number of the software",
     "doap:description": "A brief description of what the software does",
     "doap:homepage": "The homepage of the software",
-    "doap:developer": "The author of the software (can be used multiple times!)",
+    "doap:developer": "The author of the software",
     "dcterms:license": "The software license",
     "lamachine:dependency": "A dependency",
+    "lamachine:interface": "An interface",
+    "admssw:supportsFormat": "Supported data format", #ideally range is dcterms:FileFormat, we settle for plain mimetypes for now (we can also tie this directly to an interface, which is preferred!)
+    "dcterms:created": "Date of creation",
+    "dcterms:modified": "Last modified date",
+    "foaf:logo": "Logo", #just a URL for now
+    "rad:theme": "Theme/topic/domain of the software",
+    "admssw:intendedAudience": "Intended Audience for the software",
+    "rad:keyword": "Keyword for finding the software",
+    "xhv:last": "Latest available version of the software (may be newer than the one actually installed! cf. doap:revision)",
+    "lamachine:destination": "Location where the software is installed on disk", #(we can also tie this directly to an interface, which is preferred!)
+    "schema:operatingSystem": "The operating system on which this software works",
 }
 
+alias = { #just for convenience so common fields work out of the box
+    "version": "doap:revision",
+    "latest": "xhv:last",
+    "author": "doap:developer",
+    "summary": "doap:description",
+    "home-page": "doap:homepage", #used by pip
+    "title": "doap:name",
+    "licence": "dcterms:license",
+    "mimetype": "admssw:supportsFormat",
+    "requires": "lamachine:dependency",
+    "topic": "rad:theme", #used by pip
+}
+
+collections = {
[…]

### 2018-04-13 helpers/metadater.py
Update #74

@@ -32,6 +32,10 @@ properties = { #first order properties (including collections, i.e. properties t
     "doap:description": "A brief description of what the software does",
     "doap:homepage": "The homepage of the software",
     "doap:developer": "The author of the software",
+    "doap:maintainer": "The maintainer of the software",
+    "doap:repository": "Source code repository",
+    "doap:vendor": "Vendor/publisher",
+    "doap:platform": "Platform (non-OS specific), e.g. Python, Java, Firefox, ECMA CLR",
     "dcterms:license": "The software license",
     "lamachine:dependency": "A dependency",
     "lamachine:interface": "An interface",
@@ -45,6 +49,8 @@ properties = { #first order properties (including collections, i.e. properties t
     "xhv:last": "Latest available version of the software (may be newer than the one actually installed! cf. doap:revision)",
     "lamachine:destination": "Location where the software is installed on disk", #(we can also tie this directly to an interface, which is preferred!)
     "schema:operatingSystem": "The operating system on which this software works",
+    "admssw:status": "The development status of the software (e.g. alpha, beta, production, prerelease)",
+    "admssw:programmingLanguage": "Programming language in which the software is written",
 }
 
 alias = { #just for convenience so common fields work out of the box
@@ -68,13 +74,14 @@ collections = {
     "themes": "rad:theme",
     "keywords": "rad:keyword",
     "operatingSystems": "schema:operatingSystem",
+    "programmingLanguages": "admssw:programmingLanguage",
 }
 
 incollection = { v:k for k,v in collections.items() }
 
 dep_properties = {
     "doap:name": "The name of the dependency",
-    "lamachine:externalDependency": "Boolean, external dependencies are dependencies in a different software ecosystem. Internal dependencies are in the same ecosystem (e.g. Python/PyPI, Java/Maven, Perl/CPAN)",
+    "lamachine:externalPlatform": "Boolean, external dependencies are dependencies in a different software ecosystem. Internal dependencies are in the same ecosystem (e.g. Python/PyPI, Java/Maven, Perl/CPAN)",
     "lamachine:minimumVersion": "Minimum version",
     "lamachine:maximumVersion": "Maximum version",
 }
@@ -94,8 +101,9 @@ pip_mapping = { #we only need to cover the ones not already covered by common al
 pip_classifier_mapping = { #we only need to cover the ones not already covered by common aliasses
     "Operating System": "schema:operatingSystem",
[…]

### 2018-04-13 helpers/metadater.py
Further improvements on metadater script #74

@@ -5,12 +5,17 @@
 #Attempting to reuse as much as possible from ADMS.SW and DOAP:
 #https://joinup.ec.europa.eu/svn/adms_foss/adms_sw_v1.00/adms_sw_v1.00.htm
 
+# Maarten van Gompel
+# CLST, Radboud University Nijmegen
+# GPL v3
+
 
 import sys
 import argparse
 import subprocess
 import yaml
 import json
+from itertools import chain
 from collections import OrderedDict
 
 def represent_ordereddict(dumper, data):
@@ -67,17 +72,19 @@ alias = { #just for convenience so common fields work out of the box
 }
 
 collections = {
-    "developers": "doap:developer",
+    "authors": "doap:developer",
+    "developers": "doap:developer", #overrides the previous one which is just an alias
     "dependencies": "lamachine:dependency",
     "interfaces": "lamachine:interface",
     "audiences": "admssw:intendedAudience",
-    "themes": "rad:theme",
+    "topics": "rad:theme",
+    "themes": "rad:theme", #overrides the previous one which is just an alias
     "keywords": "rad:keyword",
     "operatingSystems": "schema:operatingSystem",
     "programmingLanguages": "admssw:programmingLanguage",
 }
-
 incollection = { v:k for k,v in collections.items() }
+
 
 dep_properties = {
     "doap:name": "The name of the dependency",
@@ -107,6 +114,18 @@ pip_classifier_mapping = { #we only need to cover the ones not already covered b
 }
 
 
+def qualify(key):
+    if ':' in key:
+        return key #already qualified
+    for qkey in chain(properties, interface_properties, dep_properties):
+        if qkey.split(':')[1].lower() == key.lower():
+            return qkey
+    for shortkey, qkey in alias.items():
+        if key.lower() == shortkey.lower():
+            return qkey
+    raise KeyError("Unable to qualify key: " + key)
+
+
 class SoftwareMetadata:
     def __init__(self, **kwargs):
         self.data = OrderedDict()
@@ -119,23 +138,21 @@ class SoftwareMetadata:
                     self.add(key, item)
             else:
                 self.add(key, value)
+        for collection, key in collections.items():
+            if collection in d:
+                for item in d[collection]:
+                    if isinstance(item, dict):
+                        self.add(qualify(key), { qualify(k):v for k,v in item.items() })
+                    else:
+                        self.add(qualify(key), item)
 
     def resolvekey(self, key):
-        if key.lower() in alias:
-            key = alias[key.lower()]
[…]

### 2018-04-14 helpers/metadater.py
update

@@ -261,7 +261,10 @@ def parsepip(data, lines):
                 fields = line.split('=')
                 data.add('lamachine:interface',{'lamachine:entrypoint': fields[0].strip(), 'admssw:userInterfaceType': 'cli'})
         else:
-            key, value = line.split(':',1)
+            try:
+                key, value = line.split(':',1)
+            except:
+                continue
             if key == "Author-email":
                 if "doap:developer" in data:
                     data["doap:developer"][0] += " <" + value.strip() + ">"

### 2018-04-14 helpers/metadater.py
adding webservice metadata

@@ -56,6 +56,7 @@ properties = { #first order properties (including collections, i.e. properties t
     "schema:operatingSystem": "The operating system on which this software works",
     "admssw:status": "The development status of the software (e.g. alpha, beta, production, prerelease)",
     "admssw:programmingLanguage": "Programming language in which the software is written",
+    "wdrs:describedby": "Documentation/Specification",
 }
 
 alias = { #just for convenience so common fields work out of the box
@@ -69,6 +70,7 @@ alias = { #just for convenience so common fields work out of the box
     "mimetype": "admssw:supportsFormat",
     "requires": "lamachine:dependency",
     "topic": "rad:theme", #used by pip
+    "specification": "wdrs:describedby",
 }
 
 collections = {
@@ -82,6 +84,7 @@ collections = {
     "keywords": "rad:keyword",
     "operatingSystems": "schema:operatingSystem",
     "programmingLanguages": "admssw:programmingLanguage",
+    "documentation": "wdrs:describedby",
 }
 incollection = { v:k for k,v in collections.items() }
 
@@ -97,6 +100,7 @@ interface_properties = {
     "lamachine:entrypoint": "The name of the executable, module, or an URL (depending on interfaceType)",
     "admssw:userInterfaceType": "User Interface Type", #for now we simply predefine: api (some for of shared library), cli (command line tool),  tui (Text UI), gui (Graphical UI), wui (Web UI), rest (REST webservice), soap (SOAP webservice), xmlrpc (other XMLRPC webservice), ws (other webservice)
     "admssw:supportsFormat": "Supported data format", #ideally range is dcterms:FileFormat, we settle for plain mimetypes for now
+    "wdrs:describedby": "Documentation/Specification",
     "lamachine:destination": "Location where the software is installed on disk, may be more generic or differ from lamachine:entrypoint",
 }
 

### 2018-04-14 helpers/metadater.py
fix

@@ -71,6 +71,7 @@ alias = { #just for convenience so common fields work out of the box
     "requires": "lamachine:dependency",
     "topic": "rad:theme", #used by pip
     "specification": "wdrs:describedby",
+    "interfacetype": "admssw:userInterfaceType"
 }
 
 collections = {

### 2018-04-15 helpers/metadater.py
fix

@@ -99,7 +99,7 @@ dep_properties = {
 
 interface_properties = {
     "lamachine:entrypoint": "The name of the executable, module, or an URL (depending on interfaceType)",
-    "admssw:userInterfaceType": "User Interface Type", #for now we simply predefine: api (some for of shared library), cli (command line tool),  tui (Text UI), gui (Graphical UI), wui (Web UI), rest (REST webservice), soap (SOAP webservice), xmlrpc (other XMLRPC webservice), ws (other webservice)
+    "admssw:userInterfaceType": "User Interface Type, for now we simply predefine: api (some for of shared library), cli (command line tool),  tui (Text UI), gui (Graphical UI), wui (Web UI), rest (REST webservice), soap (SOAP webservice), xmlrpc (other XMLRPC webservice), ws (other webservice)"
     "admssw:supportsFormat": "Supported data format", #ideally range is dcterms:FileFormat, we settle for plain mimetypes for now
     "wdrs:describedby": "Documentation/Specification",
     "lamachine:destination": "Location where the software is installed on disk, may be more generic or differ from lamachine:entrypoint",
@@ -303,6 +303,11 @@ def iterargs(args):
         if shortkey in args and args[shortkey]:
             yield key, args[shortkey]
 
+def propertylabel(key):
+    key = qualify(key)
+    for k, label in itertools.chain(properties.items(), dep_properties.items(), interface_properties.items()):
+        if k == key: return label
+
 def main():
     parser = argparse.ArgumentParser(description="LaMachine Metadater")
     parser.add_argument('--pip', type=str,help="Query through pip, supply the package name", action='store',required=False)
@@ -316,9 +321,9 @@ def main():
             parser.add_argument('--' + shortkey, type=str,help=help + " (" + key +  ")", action='store',required=False)
     for shortkey, key in alias.items():
         if key in incollection:
-            parser.add_argument('--' + shortkey, type=str,nargs='*', help="Alias for --"+ key.split(':')[1] +": " + properties[key], action='store',required=False)
+            parser.add_argument('--' + shortkey, type=str,nargs='*', help="Alias for --"+ key.split(':')[1] +": " + label, action='store',required=False)
         else:
-            parser.add_argument('--' + shortkey, type=str,help="Alias for --"+ key.split(':')[1] +": " + properties[key], action='store',required=False)
+            parser.add_argument('--' + shortkey, type=str,help="Alias for --"+ key.split(':')[1] +": " + label, action='store',required=False)
 
     args = parser.parse_args()
 

### 2018-04-15 helpers/metadater.py
don't just commit stuff without looking at linter output...

@@ -13,10 +13,10 @@
 import sys
 import argparse
 import subprocess
-import yaml
 import json
 from itertools import chain
 from collections import OrderedDict
+import yaml
 
 def represent_ordereddict(dumper, data):
     value = []
@@ -99,7 +99,7 @@ dep_properties = {
 
 interface_properties = {
     "lamachine:entrypoint": "The name of the executable, module, or an URL (depending on interfaceType)",
-    "admssw:userInterfaceType": "User Interface Type, for now we simply predefine: api (some for of shared library), cli (command line tool),  tui (Text UI), gui (Graphical UI), wui (Web UI), rest (REST webservice), soap (SOAP webservice), xmlrpc (other XMLRPC webservice), ws (other webservice)"
+    "admssw:userInterfaceType": "User Interface Type, for now we simply predefine: api (some for of shared library), cli (command line tool),  tui (Text UI), gui (Graphical UI), wui (Web UI), rest (REST webservice), soap (SOAP webservice), xmlrpc (other XMLRPC webservice), ws (other webservice)",
     "admssw:supportsFormat": "Supported data format", #ideally range is dcterms:FileFormat, we settle for plain mimetypes for now
     "wdrs:describedby": "Documentation/Specification",
     "lamachine:destination": "Location where the software is installed on disk, may be more generic or differ from lamachine:entrypoint",
@@ -163,8 +163,7 @@ class SoftwareMetadata:
     def incollection(self, key):
         if key in incollection:
             return incollection[key]
-        else:
-            return False
+        return False
 
     def add(self, key, value):
         key = self.resolvekey(key)
@@ -204,8 +203,7 @@ class SoftwareMetadata:
         collection = self.incollection(key)
         if collection:
             return self.data[collection]
-        else:
-            return self.data[key]
+        return self.data[key]
 
 
     def __items__(self):
@@ -305,25 +303,26 @@ def iterargs(args):
 
 def propertylabel(key):
     key = qualify(key)
-    for k, label in itertools.chain(properties.items(), dep_properties.items(), interface_properties.items()):
+    for k, label in chain(properties.items(), dep_properties.items(), interface_properties.items()):
         if k == key: return label
+    raise KeyError(key)
 
 def main():
     parser = argparse.ArgumentParser(description="LaMachine Metadater")
     parser.add_argument('--pip', type=str,help="Query through pip, supply the package name", action='store',required=False)
[…]

### 2018-04-23 bootstrap.sh
version bump to 2.2

@@ -20,7 +20,7 @@ boldblue=${bold}$(tput setaf 4) #  blue
 normal=$(tput sgr0)
 
 echo "${bold}=====================================================================${normal}"
-echo "           ,              ${bold}LaMachine v2.1.2${normal} - NLP Software distribution"
+echo "           ,              ${bold}LaMachine v2.2.0${normal} - NLP Software distribution" #also change version number in MOTD
 echo "          ~)                     (http://proycon.github.io/LaMachine)"
 echo "           (----í         Language Machines research group"
 echo "            /| |\         Centre of Language and Speech Technology"

### 2018-04-23 bootstrap.sh
Revert "version bump to 2.2"
This reverts commit bbf2c88d3452a47861c90bd4202ee4ee192e4845.

@@ -20,7 +20,7 @@ boldblue=${bold}$(tput setaf 4) #  blue
 normal=$(tput sgr0)
 
 echo "${bold}=====================================================================${normal}"
-echo "           ,              ${bold}LaMachine v2.2.0${normal} - NLP Software distribution" #also change version number in MOTD
+echo "           ,              ${bold}LaMachine v2.1.2${normal} - NLP Software distribution"
 echo "          ~)                     (http://proycon.github.io/LaMachine)"
 echo "           (----í         Language Machines research group"
 echo "            /| |\         Centre of Language and Speech Technology"

### 2018-04-23 bootstrap.sh
backtracking, smaller version bump

@@ -20,7 +20,7 @@ boldblue=${bold}$(tput setaf 4) #  blue
 normal=$(tput sgr0)
 
 echo "${bold}=====================================================================${normal}"
-echo "           ,              ${bold}LaMachine v2.1.2${normal} - NLP Software distribution"
+echo "           ,              ${bold}LaMachine v2.1.3${normal} - NLP Software distribution"
 echo "          ~)                     (http://proycon.github.io/LaMachine)"
 echo "           (----í         Language Machines research group"
 echo "            /| |\         Centre of Language and Speech Technology"

### 2018-04-24 bootstrap.sh
adapting to refactoring of metadata registry + register additionial web entrypoints

@@ -20,7 +20,7 @@ boldblue=${bold}$(tput setaf 4) #  blue
 normal=$(tput sgr0)
 
 echo "${bold}=====================================================================${normal}"
-echo "           ,              ${bold}LaMachine v2.1.3${normal} - NLP Software distribution"
+echo "           ,              ${bold}LaMachine v2.1.4${normal} - NLP Software distribution"
 echo "          ~)                     (http://proycon.github.io/LaMachine)"
 echo "           (----í         Language Machines research group"
 echo "            /| |\         Centre of Language and Speech Technology"

### 2018-04-24 docs/lamachine_extabstract_clarin2018.tex
new attempt at being less verbose

\documentclass[a4paper,11pt]{article}
%\usepackage{CLARIN2018}
% - - - - - - - IMPORTANT - - - - - - - -
% The next three lines allow XeLaTeX, graphics import and hyperlinks, set font and language
%\usepackage{xltxtra,polyglossia,graphicx,hyperref}
%\setmainfont[Mapping=tex-text]{Times}
%\setdefaultlanguage{english}
% If for some reason the above three lines are not compatible with your LaTeX installation,
% comment out the above three instructions and uncomment the following four instead:
%\usepackage{times}
\usepackage{hyperref}
\usepackage{graphicx}
%\usepackage{latexsym}
\usepackage[english]{babel}

%\setlength\titlebox{5cm}

% You can expand the titlebox if you need extra space
% to show all the authors. Please do not make the titlebox
% smaller than 5cm (the original size); we will check this
% in the camera-ready version and ask you to change it back.

\title{LaMachine NLP Software Distribution}

% - - - - - - - IMPORTANT - - - - - - -
% Leave the author information empty until your paper has been accepted

% Uncomment the following line ONLY if you need two author rows
%\setlength\titlebox{80mm}

\author{Maarten van Gompel \\
  Centre for Language and Speech Technology \\
  Radboud University Nijmegen, the Netherlands \\
  {\tt proycon@anaproy.nl} \\ %\And % if needed: this makes a second column
%  Second Author \\
%  Department (optional)\\
%  University Name without city \\
%  City, Country \\
% {\tt email@domain} \\
% \AND % if needed: this makes a second row
%  Third Author \\
%  Department (optional)\\
%  University of City, Country \\
%  {\tt email@domain} \\\And
%  Fourth Author \\
%  Department (optional)\\
%  University Name without city \\
%  City, Country \\
%  {\tt email@domain} \\
}

\date{}

\begin{document}
\maketitle

\begin{abstract}
We introduce LaMachine, a unified Natural Language Processing (NLP) open-source software distribution to facilitate the
installation and deployment of a large amount of software projects that have been developed in the scope of the
CLARIN-NL project and its current successor CLARIAH. Special attention is paid to encouragement of good software
development practises and reuse of established infrastructure in the scientific \& open-source software development
community.
\end{abstract}

\section{Introduction} \label{intro}

Software is a key deliverable and a vital component for research in projects such as CLARIN. It is software that
provides researchers the instruments to yield for their research; without it a lot of research would become highly
unfeasible or right-out impossible. It is CLARIN's core mission to make digital language resources, including software,
available to the wider research community.

We see that NLP software often takes on complex forms such as processing pipelines invoking various individual
components. Add dedicated web-interfaces on top of that and you obtain a suite of interconnected software that is often
non-trivial to install, configure, and deploy. This is where LaMachines comes in.

LaMachine incorporates software providing different types of interfaces\footnote{command line interfaces, programming
interfaces, web-user interfaces, webservice} that typically address different audiences. Whilst we attempt to
accommodate both technical\footnote{data scientists, devops, system administrators, developers} and less-technical
audiences\footnote{end-user researchers, educational settings}, there is a natural bias towards the former as
lower-level interfaces are often a prerequisite to build higher-level interfaces on. Depending on the \emph{flavour} of
LaMachine chosen; it makes a good virtual research environment for a data scientist, whether on a personal computer or on a computing
cluster, a good development environment for a developer and a good deployment method for production servers in for
example CLARIN centres.

We can qualify LaMachine as a \emph{meta distribution} that builds on established practises, technologies and resources.

\section{Architecture}

Being an open-source NLP software distribution, LaMachine is constrained to POSIX-compliant platforms; this primarily
means Linux, but also BSD and Mac OS X, and can be contrasted to mobile platforms (Android/iOS/etc), native Windows/Mac
desktop software, or certain interface types in general such as classical desktop GUI applications. Software that is
incorporated must 1) bear some relevance to NLP, and 2) be under a recognised open-source license, and is, where
necessary, compiled from source.

As briefly stated before, we can more accurately describe LaMachine as a \emph{meta distribution}, as it can be
installed in various contexts. At its core, LaMachine consists of a set of machine-parsable instructions on how to
obtain, install and configure software, these are implemented using Ansible\footnote{\url{https://www.ansible.com}}.
This is notably different from the more classical notion of Linux distributions, which generally provide their own
repositories with (often binary) software packages. LaMachine builds on this already available infrastructure by taking
these repositories as a given already and only needs to know which repositories to use.  Similarly, there are different
programming language-specific ecosystems providing their own repositories, such as the Python Package
Index\footnote{\url{https://pypi.org}} for Python, CRAN\footnote{\url{https://cran.r-project.org/}} for R,
CPAN\footnote{\url{https://www.cpan.org}} for Perl, Maven Central\footnote{\url{https://search.maven.org}} for Java.
LaMachine again simply relies on those to pull and install software from. LaMachine does not fork, archive, or modify the software
in any way. In doing so, we compel participating software projects to adhere to well-established distribution standards
and ensure the software is more sustainable towards the future. Moreover, we ensure that LaMachine never becomes a
prerequisite for the software but merely a courtesy or convenience.

LaMachine provides ample flexibility that allows it to be deployable in different contexts. First of all there is
flexibility with regard to the target platform, where we support several major GNU/Linux distributions (Debian, Ubuntu,
CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as Mac OS X. Second of all there is flexibility with
regard to the form, where we support containerisation through Docker, virtualisation through
Vagrant\footnote{\url{https://vagrant.org}} and VirtualBox\footnote{\url{https://www.virtualbox.org}},
direct remote provisioning through Ansible (for production servers), or an installation that is either global to the
machine or local in a separate directory for a specific user. Pre-built docker containers and virtual machines with a
limited selection of participating software are regularly uploaded to the Docker Hub and Vagrant Cloud, respectively. We
refer to these different context forms as different \emph{flavours}.

Installation of LaMachine begins with a single bootstrap command\footnote{Just run
\verb|bash <(curl -s https://raw.githubusercontent.com/proycon/LaMachine/master/bootstrap.sh)|
in a terminal to get started} that can
interactively query the users for their software preferences, e.g. the flavour of LaMachine and the set of software to
install, which is never static but can be customized by the user. The user may also opt for installing the latest
releases, the more experimental development versions of the software, or a very specific custom versions (to facilitate
scientific reproducibility). The bootstrap procedure detects and installs the necessary prerequisites automatically and
eventually invokes Ansible to perform the bulk of the work.

\end{document}

### 2018-04-24 docs/lamachine_extabstract_clarin2018.tex
paper update

@@
 % comment out the above three instructions and uncomment the following four instead:
-%\usepackage{times}
+\usepackage{times}
 \usepackage{hyperref}
@@
 \author{Maarten van Gompel \\
-  Centre for Language and Speech Technology \\
+  Centre for Language and Speech Technology (CLST) \\
   Radboud University Nijmegen, the Netherlands \\
@@
 We see that NLP software often takes on complex forms such as processing pipelines invoking various individual
-components. Add dedicated web-interfaces on top of that and you obtain a suite of interconnected software that is often
-non-trivial to install, configure, and deploy. This is where LaMachines comes in.
+components, which in turn rely on various dependencies. Add dedicated web-interfaces on top of that and you obtain a
+suite of interconnected software that is often non-trivial to install, configure, and deploy. This is where LaMachines
+comes in.
 
 LaMachine incorporates software providing different types of interfaces\footnote{command line interfaces, programming
-interfaces, web-user interfaces, webservice} that typically address different audiences. Whilst we attempt to
+interfaces, web-user interfaces, webservices} that typically address different audiences. Whilst we attempt to
 accommodate both technical\footnote{data scientists, devops, system administrators, developers} and less-technical
-audiences\footnote{end-user researchers, educational settings}, there is a natural bias towards the former as
+audiences\footnote{the wider researcher community, particularly the Humanities; also educational settings}, there is a natural bias towards the former as
 lower-level interfaces are often a prerequisite to build higher-level interfaces on. Depending on the \emph{flavour} of
@@
 desktop software, or certain interface types in general such as classical desktop GUI applications. Software that is
-incorporated must 1) bear some relevance to NLP, and 2) be under a recognised open-source license, and is, where
-necessary, compiled from source.
+incorporated must 1) bear some relevance to NLP, and 2) be under a recognised open-source license and deposited in a
+public version controlled repository (e.g. Github, Gitlab, Bitbucket).
 
-As briefly stated before, we can more accurately describe LaMachine as a \emph{meta distribution}, as it can be
-installed in various contexts. At its core, LaMachine consists of a set of machine-parsable instructions on how to
-obtain, install and configure software, these are implemented using Ansible\footnote{\url{https://www.ansible.com}}.
-This is notably different from the more classical notion of Linux distributions, which generally provide their own
-repositories with (often binary) software packages. LaMachine builds on this already available infrastructure by taking
-these repositories as a given already and only needs to know which repositories to use.  Similarly, there are different
-programming language-specific ecosystems providing their own repositories, such as the Python Package
-Index\footnote{\url{https://pypi.org}} for Python, CRAN\footnote{\url{https://cran.r-project.org/}} for R,
-CPAN\footnote{\url{https://www.cpan.org}} for Perl, Maven Central\footnote{\url{https://search.maven.org}} for Java.
-LaMachine again simply relies on those to pull and install software from. LaMachine does not fork, archive, or modify the software
-in any way. In doing so, we compel participating software projects to adhere to well-established distribution standards
-and ensure the software is more sustainable towards the future. Moreover, we ensure that LaMachine never becomes a
-prerequisite for the software but merely a courtesy or convenience.
+LaMachine is a \emph{meta distribution} as it can be installed in various contexts. At its core, LaMachine consists of a
+set of machine-parsable instructions on how to obtain, install and configure software, these are implemented using
+Ansible\footnote{\url{https://www.ansible.com}}.  This is notably different from the more classical notion of Linux
+distributions, which generally provide their own repositories with (often binary) software packages. LaMachine builds on
+this already available infrastructure by taking these repositories as a given and only needs to know which
+repositories to use.  Similarly, there are different programming-language-specific ecosystems providing their own
+repositories, such as the Python Package Index\footnote{\url{https://pypi.org}} for Python,
+CRAN\footnote{\url{https://cran.r-project.org/}} for R, CPAN\footnote{\url{https://www.cpan.org}} for Perl, Maven
+Central\footnote{\url{https://search.maven.org}} for Java.  LaMachine again relies on those to pull and install
+software from and never forks, archives, or modifies the software in any way. In doing so, we compel participating
+software projects to adhere to well-established distribution standards and ensure the software is more sustainable
+towards the future. Moreover, we ensure that LaMachine never becomes a prerequisite for the software but merely a
+courtesy or convenience.
 
@@
 CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as Mac OS X. Second of all there is flexibility with
-regard to the form, where we support containerisation through Docker, virtualisation through
+regard to the form, where we support \emph{containerisation} through Docker\footnote{\url{https://www.docker.com}}, virtualisation through
 Vagrant\footnote{\url{https://vagrant.org}} and VirtualBox\footnote{\url{https://www.virtualbox.org}},
@@
 machine or local in a separate directory for a specific user. Pre-built docker containers and virtual machines with a
-limited selection of participating software are regularly uploaded to the Docker Hub and Vagrant Cloud, respectively. We
-refer to these different context forms as different \emph{flavours}.
+limited selection of participating software are regularly uploaded to the Docker Hub and Vagrant Cloud, respectively.
 
-Installation of LaMachine begins with a single bootstrap command\footnote{Just run
-\verb|bash <(curl -s https://raw.githubusercontent.com/proycon/LaMachine/master/bootstrap.sh)|
-in a terminal to get started} that can
-interactively query the users for their software preferences, e.g. the flavour of LaMachine and the set of software to
-install, which is never static but can be customized by the user. The user may also opt for installing the latest
-releases, the more experimental development versions of the software, or a very specific custom versions (to facilitate
-scientific reproducibility). The bootstrap procedure detects and installs the necessary prerequisites automatically and
-eventually invokes Ansible to perform the bulk of the work.
+Installation of LaMachine begins with a single bootstrap command\footnote{See
+\url{https://proycon.github.io/LaMachine}}.  It may interactively query the users for their software preferences, e.g.
+the flavour of LaMachine and the set of software to install, which is never static but can be customized by the user.
+The user may also opt for installing the latest releases, the more experimental development versions of the software, or
+a very specific custom versions (to facilitate scientific reproducibility). The bootstrap procedure detects and installs
+the necessary prerequisites automatically and eventually invokes Ansible to perform the bulk of the work.
+
+After a successful build, the user may interact with LaMachine either through the command line, which offers a standard
+shell and enables access to all lower-level tools and programming languages; or through his or her webbrowser, which
+when enabled offers a simple portal page towards all installed web-capable tools. Amongst these services is also a
+Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
+for e.g. Python and R that has gained great popularity in the scientific community.
+
+The current version of LaMachine comes with some simple data sharing facilities only: care is taken only to provide a
+single shared dataspace between host and VM/container (where applicable), or allow simple upload. Extensive data
+search and management functions are deliberately beyond the scope of LaMachine, and left to more high-level tooling.
+
+With LaMachine we also attempt to harmonise the metadata of all installed software, by converting metadata upstream repositories to
+the CodeMeta standard\footnote{\url{https://codemeta.github.io/}} where possible, or encouraging software developers to provide
+their own. This in turn enables other tools to do proper service discovery, and provenance logging.
+
+\section{Software}
+
+As LaMachine was conceived initially as the primary means of distribution of the software stack developed at CLST,
+Radboud University Nijmegen, it includes a lot of our software. A full list goes beyond the scope of this overview, we
+will merely mention some CLARIN/CLARIAH-funded tools: ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format
+for linguistic Annotation, with assorted tools), FLAT (a web-based linguistic Annotation tool), PICCL (an OCR and
+post-OCR correction pipeline) and CLAM. We already include some relevant software by Groningen University and VU
+Amsterdam. Moreover, LaMachine incorporates a large number of renowned tools by external parties, offering most notably
+a mature Python environment\footnote{scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch, keras, gensim,
+etc...}, but also offering space for e.g. R and Java.
+
+\section{Future}
+
+The very recent release of version 2, which constituted a full rewrite, has opened up LaMachine to outside contribution.
+Contributor documentation has been written, and at this stage, we greatly welcome external participants to join in. The
+main objectives for the future are to providing greater \emph{interoperability} between the included tools through
+better \emph{higher-level interfaces} for the researcher. We see this as a bottom-up process and have now established a
+firm foundation to build upon. LaMachine development presently takes place in the scope of the CLARIAH WP3 Virtual
+Research Environment project, which has much higher ambitions in accommodating the researcher and connectivity of
+data and services. A major part of our future focus will be on interoperability with the higher-level tooling developed there, or in the
+wider CLARIN/CLARIAH commmunity.
 

### 2018-04-24 docs/lamachine_extabstract_clarin2018.tex
paper update

@@
   {\tt proycon@anaproy.nl} \\ %\And % if needed: this makes a second column
+  \url{https://proycon.github.io/LaMachine}
 %  Second Author \\
@@
 With LaMachine we also attempt to harmonise the metadata of all installed software, by converting metadata upstream repositories to
-the CodeMeta standard\footnote{\url{https://codemeta.github.io/}} where possible, or encouraging software developers to provide
+the CodeMeta standard\footnote{\url{https://codemeta.github.io/}} \cite{codemeta,codemetar} where possible, or encouraging software developers to provide
 their own. This in turn enables other tools to do proper service discovery, and provenance logging.
@@
 Radboud University Nijmegen, it includes a lot of our software. A full list goes beyond the scope of this overview, we
-will merely mention some CLARIN/CLARIAH-funded tools: ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format
+will merely mention some CLARIN-NL/CLARIAH-funded tools: ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format
 for linguistic Annotation, with assorted tools), FLAT (a web-based linguistic Annotation tool), PICCL (an OCR and
 post-OCR correction pipeline) and CLAM. We already include some relevant software by Groningen University and VU
-Amsterdam. Moreover, LaMachine incorporates a large number of renowned tools by external parties, offering most notably
-a mature Python environment\footnote{scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch, keras, gensim,
-etc...}, but also offering space for e.g. R and Java.
+Amsterdam as well. Moreover, LaMachine incorporates a large number of renowned tools by external international parties,
+offering most notably a mature Python environment\footnote{scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch,
+keras, gensim, etc...}, but also offering space for e.g. R and Java.
 
-\section{Future}
+\section{Conclusion \& Future work}
 
-The very recent release of version 2, which constituted a full rewrite, has opened up LaMachine to outside contribution.
-Contributor documentation has been written, and at this stage, we greatly welcome external participants to join in. The
-main objectives for the future are to providing greater \emph{interoperability} between the included tools through
-better \emph{higher-level interfaces} for the researcher. We see this as a bottom-up process and have now established a
-firm foundation to build upon. LaMachine development presently takes place in the scope of the CLARIAH WP3 Virtual
-Research Environment project, which has much higher ambitions in accommodating the researcher and connectivity of
-data and services. A major part of our future focus will be on interoperability with the higher-level tooling developed there, or in the
-wider CLARIN/CLARIAH commmunity.
+The very recent release of LaMachine v2, which constituted a full rewrite, has opened up LaMachine to outside
+contribution. Contributor documentation has been written, and at this stage, we greatly welcome external participants
+to join in. Aside from the incorporation of new relevant software, the main objectives for the future are to provide
+greater \emph{interoperability} between the included tools through better \emph{high-level interfaces} for the
+researcher. We see this as a bottom-up process and have now established a firm foundation to build upon. Note that such
+proposed interfaces, including the current portal application in LaMachine, are always considered separate independent
+software projects, which may be deployed by/in/for LaMachine, but also in other contexts. LaMachine remains `just'
+a software distribution at heart.
+
+Development of LaMachine presently takes place in the scope of the CLARIAH WP3 Virtual Research Environment (VRE)
+project\footnote{https://github.com/meertensinstituut/clariah-wp3-vre}, which has much higher ambitions in accommodating
+the researcher and connectivity of data and services, and transcends also those of the CLARIN LRS Switchboard project
+\cite{switchboard}. An important part of our future focus will therefore be on interoperability with the higher-level
+tools emerging from the VRE efforts, but also with other parts of the CLARIN infrastructure; single-sign on
+authentication being a notable example here.
+
+\bibliographystyle{acl}
+\bibliography{lamachine}
 

### 2018-04-24 docs/lamachine_extabstract_clarin2018.tex
paper update

@@
 
-\title{LaMachine NLP Software Distribution}
+\title{LaMachine: A meta-distribution for NLP software}
 
@@
 components, which in turn rely on various dependencies. Add dedicated web-interfaces on top of that and you obtain a
-suite of interconnected software that is often non-trivial to install, configure, and deploy. This is where LaMachines
+suite of interconnected software that is often non-trivial to install, configure, and deploy. This is where LaMachine
 comes in.
@@
 Being an open-source NLP software distribution, LaMachine is constrained to POSIX-compliant platforms; this primarily
-means Linux, but also BSD and Mac OS X, and can be contrasted to mobile platforms (Android/iOS/etc), native Windows/Mac
-desktop software, or certain interface types in general such as classical desktop GUI applications. Software that is
-incorporated must 1) bear some relevance to NLP, and 2) be under a recognised open-source license and deposited in a
-public version controlled repository (e.g. Github, Gitlab, Bitbucket).
+means Linux, but also BSD and, with some restrictions, Mac OS X. This can be put in contrast to mobile platforms
+(Android/iOS/etc), native Windows/Mac desktop software, or certain interface types in general such as classical desktop
+GUI applications. Software that is incorporated must 1) bear some relevance to NLP, and 2) be under a recognised
+open-source license and 3) be deposited in a public version controlled repository\footnote{e.g. Github, Gitlab,
+Bitbucket}.
 
@@
 software projects to adhere to well-established distribution standards and ensure the software is more sustainable
-towards the future. Moreover, we ensure that LaMachine never becomes a prerequisite for the software but merely a
+towards the future \cite{softwarequality}. Moreover, we ensure that LaMachine never becomes a prerequisite for the software but merely a
 courtesy or convenience.
@@
 CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as Mac OS X. Second of all there is flexibility with
-regard to the form, where we support \emph{containerisation} through Docker\footnote{\url{https://www.docker.com}}, virtualisation through
-Vagrant\footnote{\url{https://vagrant.org}} and VirtualBox\footnote{\url{https://www.virtualbox.org}},
+regard to the form, where we support \emph{containerisation} through Docker\footnote{\url{https://www.docker.com}},
+\emph{virtualisation} through Vagrant\footnote{\url{https://vagrant.org}} and VirtualBox\footnote{\url{https://www.virtualbox.org}},
 direct remote provisioning through Ansible (for production servers), or an installation that is either global to the
-machine or local in a separate directory for a specific user. Pre-built docker containers and virtual machines with a
+machine or local in a custom directory for a specific user. Pre-built docker containers and virtual machines with a
 limited selection of participating software are regularly uploaded to the Docker Hub and Vagrant Cloud, respectively.
@@
 project\footnote{https://github.com/meertensinstituut/clariah-wp3-vre}, which has much higher ambitions in accommodating
-the researcher and connectivity of data and services, and transcends also those of the CLARIN LRS Switchboard project
-\cite{switchboard}. An important part of our future focus will therefore be on interoperability with the higher-level
-tools emerging from the VRE efforts, but also with other parts of the CLARIN infrastructure; single-sign on
+the researcher and connectivity of data and services, and transcends also those of the CLARIN LRS Switchboard
+project~\cite{switchboard}. An important part of our future focus will therefore be on interoperability with the
+higher-level tools emerging from the VRE efforts, but also with other parts of the CLARIN infrastructure; single-sign on
 authentication being a notable example here.

### 2018-04-24 docs/lamachine_extabstract_clarin2018.tex
update
first round of fixes (thanks @martinreynaert)

@@
 
-We can qualify LaMachine as a \emph{meta distribution} that builds on established practises, technologies and resources.
-
 \section{Architecture}
@@
 (Android/iOS/etc), native Windows/Mac desktop software, or certain interface types in general such as classical desktop
-GUI applications. Software that is incorporated must 1) bear some relevance to NLP, and 2) be under a recognised
-open-source license and 3) be deposited in a public version controlled repository\footnote{e.g. Github, Gitlab,
-Bitbucket}.
+GUI applications. Software that is incorporated must 1) bear some relevance to NLP, 2) be under a recognised
+open-source license, 3) be deposited in a public version controlled repository\footnote{e.g. Github, Gitlab,
+Bitbucket} and 4) have a release protocol (with semantic versioning) using the proper technology-specific channels.
 
@@
 The user may also opt for installing the latest releases, the more experimental development versions of the software, or
-a very specific custom versions (to facilitate scientific reproducibility). The bootstrap procedure detects and installs
+specific custom versions (to facilitate scientific reproducibility). The bootstrap procedure detects and installs
 the necessary prerequisites automatically and eventually invokes Ansible to perform the bulk of the work.
@@
 
-The current version of LaMachine comes with some simple data sharing facilities only: care is taken only to provide a
+The current version of LaMachine comes with some simple data sharing facilities only: we provide a
 single shared dataspace between host and VM/container (where applicable), or allow simple upload. Extensive data
@@
 
-With LaMachine we also attempt to harmonise the metadata of all installed software, by converting metadata upstream repositories to
-the CodeMeta standard\footnote{\url{https://codemeta.github.io/}} \cite{codemeta,codemetar} where possible, or encouraging software developers to provide
-their own. This in turn enables other tools to do proper service discovery, and provenance logging.
+With LaMachine we also attempt to harmonise the metadata of all installed software, by converting metadata from upstream repositories to
+the CodeMeta standard\footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or encouraging software developers to provide
+their own. This in turn enables other tools to do proper service discovery and provenance logging.
 
@@
 
-As LaMachine was conceived initially as the primary means of distribution of the software stack developed at CLST,
+As LaMachine was conceived initially\footnote{LaMachine exists since May 2015 and has been used extensively ever since by
+numerous users}, as the primary means of distribution of the software stack developed at CLST,
 Radboud University Nijmegen, it includes a lot of our software. A full list goes beyond the scope of this overview, we
 will merely mention some CLARIN-NL/CLARIAH-funded tools: ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format
-for linguistic Annotation, with assorted tools), FLAT (a web-based linguistic Annotation tool), PICCL (an OCR and
+for Linguistic Annotation, with assorted tools), FLAT (a web-based linguistic Annotation tool), PICCL (an OCR and
 post-OCR correction pipeline) and CLAM. We already include some relevant software by Groningen University and VU
@@
 offering most notably a mature Python environment\footnote{scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch,
-keras, gensim, etc...}, but also offering space for e.g. R and Java.
+keras, gensim, etc...}, but also R and Java.
 

### 2018-04-24 docs/lamachine_clarin2018.tex

@@
 \documentclass[a4paper,11pt]{article}
-\usepackage{CLARIN2018}
+%\usepackage{CLARIN2018}
 % - - - - - - - IMPORTANT - - - - - - - -
 % The next three lines allow XeLaTeX, graphics import and hyperlinks, set font and language
-\usepackage{xltxtra,polyglossia,graphicx,hyperref}
-\setmainfont[Mapping=tex-text]{Times}
-\setdefaultlanguage{english}
+%\usepackage{xltxtra,polyglossia,graphicx,hyperref}
+%\setmainfont[Mapping=tex-text]{Times}
+%\setdefaultlanguage{english}
 % If for some reason the above three lines are not compatible with your LaTeX installation,
@@
 %\usepackage{times}
-%\usepackage{url}
+\usepackage{hyperref}
+\usepackage{graphicx}
 %\usepackage{latexsym}
-%\usepackage[english]{babel}
+\usepackage[english]{babel}
 
@@
 % in the camera-ready version and ask you to change it back.
-
-\usepackage{covington} % if needed, for linguistic examples
 
@@
 \begin{abstract}
-We introduce LaMachine, a unified Natural Language Processing (NLP) open-source
-software distribution to facilitate the installation and deployment of a large
-amount of software projects that have been developed in the scope of the
-CLARIN-NL project and its current successor CLARIAH. Special attention is paid
-to encouragement of good software development practises.
+We introduce LaMachine, a unified Natural Language Processing (NLP) open-source software distribution to facilitate the
+installation and deployment of a large amount of software projects that have been developed in the scope of the
+CLARIN-NL project and its current successor CLARIAH. Special attention is paid to encouragement of good software
+development practises and reuse of established infrastructure in the scientific \& open-source software development
+community.
 \end{abstract}
@@
 
-Software is a key deliverable and a vital component for research in projects
-such as CLARIN. It is software that provides researchers the instruments to
-yield for their research, and without it a lot of research would become highly
-unfeasible or right-out impossible. It is CLARIN's core mission to make digital
-language resources, including software, available to the wider research
-community.
+Software is a key deliverable and a vital component for research in projects such as CLARIN. It is software that
+provides researchers the instruments to yield for their research; without it a lot of research would become highly
+unfeasible or right-out impossible. It is CLARIN's core mission to make digital language resources, including software,
+available to the wider research community.
 
-Software is a broad but well-ingrained notion in society nowadays, referring to
-any form of computer program. This can manifest in various forms, whether it is
-an app on a phone, a graphical desktop application, a command line tool,  or a
-webpage.
+Software is a broad but well-ingrained notion in society nowadays, referring to any form of computer program. This can
+manifest in various forms, whether it is an app on a phone, a graphical desktop application, a command line tool, or a
+web-based application. These different \emph{interfaces} generally address different \emph{audiences}; the data
+scientist will feel at home at the command line and in scripting environments and uses these fairly low-level
+interfaces, whereas researchers in the humanities demand higher-level interfaces. There is often a power trade-off
+between lower and higher-level interfaces, with the former providing maximum flexibility at the cost of a steeper
+learning curve, technical ability, and a do-it-yourself mentality. Higher level interfaces, on the other
+hand, expose certain functionality of the software in an easy and accessible way, but in doing so often can not expose
+the full power of the software. The cost trade-off is also apparant in the construction of the interfaces, where
+high-level interfaces are typically far more costly to build.
 
-one classification could be
-made based on the \emph{interfaces} software offers.
+LaMachine attemps to address both audiences and incorporates software providing different types of interfaces. This
+explicitly includes web-based interfaces, both for human end-users as well as machines (i.e. webservices). Development
+of LaMachine proceeds in a bottom-up fashion, which introduces a natural bias towards lower-level interfaces, as those
+are often a prerequisite to build higher-level interfaces on.
 
-We too often see that the research community develops valuable
+We see that NLP software often takes on complex forms such as processing pipelines invoking various individual
+components. Add dedicated web-interfaces on top of that and you obtain a suite of interconnected software that is often
+non-trivial to install. This is precisely where LaMachine shines.
+
+\section{Architecture}
+
+\subsection{Scope}
+
+LaMachine is an open-source NLP software distribution. This implies the following:
+
+\begin{enumerate}
+    \item LaMachine facilitates the installation, distribution and configuration of software. It does not fork, modify
+        or appropriate the participating software in any way, nor does it provide a hosting place or repository for
+        software.
+    \item The software included in LaMachine must under an OSI-approved open-source license and is, where necessary,
+        compiled from source during installation.
+    \item The included software bears at least some relevance to the field of Natural Language Processing.
+    \item LaMachine lives in an open-source ecosystem and therefore builds on POSIX-compliant platforms; this primarily
+        means Linux, as well as BSD and Mac OS X. This by definition excludes certain software for different platforms, such as
+        mobile platforms (Android/iOS/etc), native Windows/Mac desktop applications, or certain interface types in
+        general such as classical desktop GUI applications.
+\end{enumerate}
+
+We can more accurately describe LaMachine as a \emph{meta distribution}: it can be installed in various forms, which we
+will get back to shortly. At its core, LaMachine consists of a set of machine-parsable instructions on how to obtain,
+install and configure software, these are implemented using Ansible\footnote{\url{https://www.ansible.com}}. This is
+notably different from the more classical notion of Linux distributions, which generally provide their own repositories
+with (often binary) software packages. LaMachine builds on this already available infrastructure by taking these
+repositories as a given already and only needs to know which repositories to use.  Similarly, there are different
+programming language-specific ecosystems providing their own repositories, such as the Python Package
+Index\footnote{\url{https://pypi.org}} for Python, CRAN\footnote{\url{https://cran.r-project.org/}} for R,
+CPAN\footnote{\url{https://www.cpan.org}} for Perl, Maven Central\footnote{\url{https://search.maven.org}} for Java.
+LaMachine again simply relies on those to pull and install software from. In doing so, we compel participating software
+projects to adhere to well-established distribution standards and ensure the software is more sustainable towards the
+future.
+
+LaMachine provides ample flexibility that allows it to be deployable in different contexts. First of all there is
+flexibility with regard to the target platform, where we support several major GNU/Linux distributions (Debian, Ubuntu,
+CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as Mac OS X. Second of all there is flexibility with
+regard to the form, where we support containerisation through Docker, virtualisation through Vagrant and VirtualBox,
+direct remote provisioning through Ansible (for production servers), or a installation that is either global to the
+machine or local in a separate directory for a specific user.
+
+\end{document}

### 2018-04-24 docs/lamachine_extabstract_clarin2018.tex

@@
 Development of LaMachine presently takes place in the scope of the CLARIAH WP3 Virtual Research Environment (VRE)
-project\footnote{https://github.com/meertensinstituut/clariah-wp3-vre}, which has much higher ambitions in accommodating
+project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}}, which has much higher ambitions in accommodating
 the researcher and connectivity of data and services, and transcends also those of the CLARIN LRS Switchboard

### 2018-04-25 roles/lamachine-core/templates/lamachine-start-webserver.sh
attempted fix for broken webserver script when there is no systemd (e.g. docker) #80

@@ -6,19 +6,35 @@
 
 {% if locality == "global" and root %}
 #global flavour
+if which systemctl >/dev/null 2>/dev/null; then
+    HAVE_SYSTEMCTL=1
+else
+    HAVE_SYSTEMCTL=0 #If there is no systemd, assume init V and 'service' command, this is relevant also in most docker containers where systemd makes less sense
+fi
 
 #kill and restart uwsgi emperor
 {% if ansible_distribution|lower == "debian" or ansible_distribution|lower == "ubuntu" or ansible_distribution|lower == "linux mint" %}
- sudo systemctl enable uwsgi-emperor
- sudo systemctl start uwsgi-emperor
+if [ $HAVE_SYSTEMCTL -eq 1 ]; then
+     sudo systemctl enable uwsgi-emperor
+     sudo systemctl start uwsgi-emperor
+else
+     sudo service uwsgi-emperor start
+fi
 {% else %}
- sudo uwsgi --ini "{{lm_prefix}}/etc/uwsgi-emperor/emperor.ini" --die-on-term &
+     sudo uwsgi --ini "{{lm_prefix}}/etc/uwsgi-emperor/emperor.ini" --die-on-term &
 {% endif %}
 
 {% if  webservertype == "nginx" %}
- sudo systemctl enable nginx
- sudo systemctl start nginx
+if [ $HAVE_SYSTEMCTL -eq 1 ]; then
+     sudo systemctl enable nginx
+     sudo systemctl start nginx
+else
+     sudo service nginx start
+fi
+{% else %}
+    echo "You are using a non-default webservertype, unable to manage webserver for you...">&2
 {% endif %}
+
 
 
 {% else %}
@@ -27,8 +43,15 @@
 uwsgi --ini "{{lm_prefix}}/etc/uwsgi-emperor/emperor.ini" --die-on-term &
 
 {% if webservertype == "nginx" %}
-nginx -c "{{lm_prefix}}/etc/nginx.conf" -p "{{lm_prefix}}/share/nginx"  -g "pid {{lm_prefix}}/var/run/nginx.pid; worker_processes 2;"
+    nginx -c "{{lm_prefix}}/etc/nginx.conf" -p "{{lm_prefix}}/share/nginx"  -g "pid {{lm_prefix}}/var/run/nginx.pid; worker_processes 2;"
+{% else %}
+    echo "You are using a non-default webservertype, unable to manage webserver for you...">&2
[…]

### 2018-04-25 roles/lamachine-core/templates/lamachine-stop-webserver.sh

@@ -4,16 +4,29 @@
 
 {% if locality == "global" and root %}
 #global flavour
+if which systemctl >/dev/null 2>/dev/null; then
+    HAVE_SYSTEMCTL=1
+else
+    HAVE_SYSTEMCTL=0 #If there is no systemd, assume init V, this is relevant also in most docker containers where systemd makes less sense
+fi
 
 #kill and restart uwsgi emperor
 {% if ansible_distribution|lower == "debian" or ansible_distribution|lower == "ubuntu" or ansible_distribution|lower == "linux mint" %}
- sudo systemctl stop uwsgi-emperor
+if [ $HAVE_SYSTEMCTL -eq 1 ]; then
+     sudo systemctl stop uwsgi-emperor
+else
+     sudo service uwsgi-emperor stop
+fi
 {% else %}
  sudo killall -w uwsgi
 {% endif %}
 
 {% if  webservertype == "nginx" %}
- sudo systemctl stop nginx
+if [ $HAVE_SYSTEMCTL -eq 1 ]; then
+     sudo systemctl stop nginx
+else
+     sudo service nginx stop
+fi
 {% endif %}
 
 

### 2018-04-25 bootstrap.sh
version bump

@@ -20,7 +20,7 @@ boldblue=${bold}$(tput setaf 4) #  blue
 normal=$(tput sgr0)
 
 echo "${bold}=====================================================================${normal}"
-echo "           ,              ${bold}LaMachine v2.1.4${normal} - NLP Software distribution"
+echo "           ,              ${bold}LaMachine v2.1.5${normal} - NLP Software distribution"
 echo "          ~)                     (http://proycon.github.io/LaMachine)"
 echo "           (----í         Language Machines research group"
 echo "            /| |\         Centre of Language and Speech Technology"

### 2018-04-25 bootstrap.sh
set up dev branch again

@@ -499,9 +499,9 @@ fi
 
 if [ -z "$BRANCH" ]; then
     if [[ "$VERSION" == "development" ]]; then
-        BRANCH="master"
+        BRANCH="develop"
     else
-        BRANCH="master"
+        BRANCH="develop"
     fi
 fi
 if [ -z "$GITREPO" ]; then

### 2018-04-27 roles/lamachine-core/templates/lamachine-start-webserver.sh
local nginx fixes

@@ -43,7 +43,7 @@ fi
 uwsgi --ini "{{lm_prefix}}/etc/uwsgi-emperor/emperor.ini" --die-on-term &
 
 {% if webservertype == "nginx" %}
-    nginx -c "{{lm_prefix}}/etc/nginx.conf" -p "{{lm_prefix}}/share/nginx"  -g "pid {{lm_prefix}}/var/run/nginx.pid; worker_processes 2;"
+    nginx -c "{{lm_prefix}}/etc/nginx/nginx.conf" -p "{{lm_prefix}}"  -g "pid {{lm_prefix}}/var/run/nginx.pid;"
 {% else %}
     echo "You are using a non-default webservertype, unable to manage webserver for you...">&2
 {% endif %}

### 2018-04-30 docs/lamachine_extabstract_clarin2018.tex
processed comments by menzo

--- 2018-04-24 docs/lamachine_extabstract_clarin2018.tex
@@
   Centre for Language and Speech Technology (CLST) \\
-  Radboud University Nijmegen, the Netherlands \\
+  Radboud University, Nijmegen, the Netherlands \\
   {\tt proycon@anaproy.nl} \\ %\And % if needed: this makes a second column
@@
 CLARIN-NL project and its current successor CLARIAH. Special attention is paid to encouragement of good software
-development practises and reuse of established infrastructure in the scientific \& open-source software development
+development practises and reuse of established infrastructure in the scientific and open-source software development
 community.
@@
 
-Software is a key deliverable and a vital component for research in projects such as CLARIN. It is software that
-provides researchers the instruments to yield for their research; without it a lot of research would become highly
-unfeasible or right-out impossible. It is CLARIN's core mission to make digital language resources, including software,
-available to the wider research community.
+Software is a key deliverable and a vital component for research in projects such as those under the CLARIN umbrella. It
+is software that provides researchers the instruments to yield for their research; without it a lot of research would
+become highly unfeasible or right-out impossible. It is CLARIN's core mission to make digital language resources,
+including software, available to the wider research community.
 
@@
 
-LaMachine incorporates software providing different types of interfaces\footnote{command line interfaces, programming
-interfaces, web-user interfaces, webservices} that typically address different audiences. Whilst we attempt to
-accommodate both technical\footnote{data scientists, devops, system administrators, developers} and less-technical
-audiences\footnote{the wider researcher community, particularly the Humanities; also educational settings}, there is a natural bias towards the former as
+LaMachine incorporates software providing different types of interfaces\footnote{Command line interfaces, programming
+interfaces, web-user interfaces, webservices.} that typically address different audiences. Whilst we attempt to
+accommodate both technical\footnote{Data scientists, DevOps, system administrators, developers.} and less-technical
+audiences\footnote{The wider researcher community, particularly the Humanities; also educational settings.}, there is a natural bias towards the former as
 lower-level interfaces are often a prerequisite to build higher-level interfaces on. Depending on the \emph{flavour} of
-LaMachine chosen; it makes a good virtual research environment for a data scientist, whether on a personal computer or on a computing
-cluster, a good development environment for a developer and a good deployment method for production servers in for
+LaMachine chosen, it makes a good virtual research environment for a data scientist, whether on a personal computer or on a computing
+cluster, a good development environment for a developer or a good deployment method for production servers in for
 example CLARIN centres.
@@
 Being an open-source NLP software distribution, LaMachine is constrained to POSIX-compliant platforms; this primarily
-means Linux, but also BSD and, with some restrictions, Mac OS X. This can be put in contrast to mobile platforms
-(Android/iOS/etc), native Windows/Mac desktop software, or certain interface types in general such as classical desktop
-GUI applications. Software that is incorporated must 1) bear some relevance to NLP, 2) be under a recognised
-open-source license, 3) be deposited in a public version controlled repository\footnote{e.g. Github, Gitlab,
-Bitbucket} and 4) have a release protocol (with semantic versioning) using the proper technology-specific channels.
+means Linux, but also BSD and, with some restrictions, macOS. This stands in contrast with mobile platforms
+(Android/iOS/etc), native Windows/mac desktop software, or certain interface types in general such as classical desktop
+GUI applications, all of which fall beyond the scope of LaMachine. Software that is incorporated must 1) bear some
+relevance to NLP, 2) be under a recognised open-source license, 3) be deposited in a public version controlled
+repository\footnote{e.g. Github, Gitlab, Bitbucket} and 4) have a release protocol (with semantic versioning) using the
+proper technology-specific channels.
 
 LaMachine is a \emph{meta distribution} as it can be installed in various contexts. At its core, LaMachine consists of a
-set of machine-parsable instructions on how to obtain, install and configure software, these are implemented using
+set of machine-parsable instructions on how to obtain, install and configure software. These are implemented using
 Ansible\footnote{\url{https://www.ansible.com}}.  This is notably different from the more classical notion of Linux
@@
 flexibility with regard to the target platform, where we support several major GNU/Linux distributions (Debian, Ubuntu,
-CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as Mac OS X. Second of all there is flexibility with
+CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as macOS. Second, there is flexibility with
 regard to the form, where we support \emph{containerisation} through Docker\footnote{\url{https://www.docker.com}},
@@
 As LaMachine was conceived initially\footnote{LaMachine exists since May 2015 and has been used extensively ever since by
-numerous users}, as the primary means of distribution of the software stack developed at CLST,
-Radboud University Nijmegen, it includes a lot of our software. A full list goes beyond the scope of this overview, we
+numerous users} as the primary means of distribution of the software stack developed at CLST,
+Radboud University Nijmegen, it includes a lot of our software. A full list goes beyond the scope of this overview; we
 will merely mention some CLARIN-NL/CLARIAH-funded tools: ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format
 for Linguistic Annotation, with assorted tools), FLAT (a web-based linguistic Annotation tool), PICCL (an OCR and
-post-OCR correction pipeline) and CLAM. We already include some relevant software by Groningen University and VU
+post-OCR correction pipeline) and CLAM. We already include some relevant software by the University of Groningen and VU
 Amsterdam as well. Moreover, LaMachine incorporates a large number of renowned tools by external international parties,
@@
 
-The very recent release of LaMachine v2, which constituted a full rewrite, has opened up LaMachine to outside
+The recent release of LaMachine v2, which constituted a full rewrite, has opened up LaMachine to outside
 contribution. Contributor documentation has been written, and at this stage, we greatly welcome external participants
@@
 Development of LaMachine presently takes place in the scope of the CLARIAH WP3 Virtual Research Environment (VRE)
-project\footnote{https://github.com/meertensinstituut/clariah-wp3-vre}, which has much higher ambitions in accommodating
-the researcher and connectivity of data and services, and transcends also those of the CLARIN LRS Switchboard
-project~\cite{switchboard}. An important part of our future focus will therefore be on interoperability with the
-higher-level tools emerging from the VRE efforts, but also with other parts of the CLARIN infrastructure; single-sign on
-authentication being a notable example here.
+project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}}, which has higher ambitions in
+accommodating the researcher and connectivity of data and services, and transcends also those of the CLARIN Language
+Resource Switchboard~\cite{switchboard}. An important part of our future focus will therefore be on interoperability
+with the higher-level tools emerging from the VRE efforts, but also with other parts of the CLARIN infrastructure;
+single-sign on authentication being a notable example here.
 

### 2018-04-30 bootstrap.sh
version bump

@@ -20,7 +20,7 @@ boldblue=${bold}$(tput setaf 4) #  blue
 normal=$(tput sgr0)
 
 echo "${bold}=====================================================================${normal}"
-echo "           ,              ${bold}LaMachine v2.1.5${normal} - NLP Software distribution"
+echo "           ,              ${bold}LaMachine v2.1.6${normal} - NLP Software distribution"
 echo "          ~)                     (http://proycon.github.io/LaMachine)"
 echo "           (----í         Language Machines research group"
 echo "            /| |\         Centre of Language and Speech Technology"

### 2018-05-01 bootstrap.sh
version bump

@@ -20,7 +20,7 @@ boldblue=${bold}$(tput setaf 4) #  blue
 normal=$(tput sgr0)
 
 echo "${bold}=====================================================================${normal}"
-echo "           ,              ${bold}LaMachine v2.1.6${normal} - NLP Software distribution"
+echo "           ,              ${bold}LaMachine v2.1.7${normal} - NLP Software distribution"
 echo "          ~)                     (http://proycon.github.io/LaMachine)"
 echo "           (----í         Language Machines research group"
 echo "            /| |\         Centre of Language and Speech Technology"

### 2018-05-02 bootstrap.sh
setting up master branch

@@ -499,9 +499,9 @@ fi
 
 if [ -z "$BRANCH" ]; then
     if [[ "$VERSION" == "development" ]]; then
-        BRANCH="develop"
+        BRANCH="master"
     else
-        BRANCH="develop"
+        BRANCH="master"
     fi
 fi
 if [ -z "$GITREPO" ]; then

### 2018-05-02 builds/test.py

@@ -49,7 +49,7 @@ def test(build, args):
         #we build local tests in a docker container (clean environment), the dockerfile invokes the bootstrap:
         cwd = os.getcwd()
         os.chdir("context/" + build['context'])
-        cmd = "docker build -t proycon/lamachine:" + build['name'] + " --build-arg NAME=" + build['name'] + " --build-arg VERSION=" + build['version'] + " --build-arg BRANCH=develop --build-arg HOSTNAME=lamachine-" + build['name'] + " . 2>> " + cwd + "/logs/" + buildid(build).replace(':','-') + '.log >&2'
+        cmd = "docker build -t proycon/lamachine:" + build['name'] + " --build-arg NAME=" + build['name'] + " --build-arg VERSION=" + build['version'] + " --build-arg BRANCH=master --build-arg HOSTNAME=lamachine-" + build['name'] + " . 2>> " + cwd + "/logs/" + buildid(build).replace(':','-') + '.log >&2'
         r = os.system(cmd)
         os.chdir(cwd)
     else:

### 2018-07-23 docs/lamachine_extabstract_clarin2018.tex
draft of Clarin paper with Iris as second author and added a case study

@@
 
-\author{Maarten van Gompel \\
+\author{Maarten van Gompel  and Iris Hendrickx\\
   Centre for Language and Speech Technology (CLST) \\
   Radboud University, Nijmegen, the Netherlands \\
-  {\tt proycon@anaproy.nl} \\ %\And % if needed: this makes a second column
+  {\tt proycon@anaproy.nl, i.hendrickx@let.ru.nl} \\ %\And % if needed: this makes a second column
   \url{https://proycon.github.io/LaMachine}
@@
 community.
+We illustrate the usage of LaMachine in an exploratory text mining project at the Dutch Health Inspectorate where LaMachine was applied to create a research environment for automatic text analysis for health care quality monitoring.
 \end{abstract}
@@
 Software is a key deliverable and a vital component for research in projects such as those under the CLARIN umbrella. It
-is software that provides researchers the instruments to yield for their research; without it a lot of research would
-become highly unfeasible or right-out impossible. It is CLARIN's core mission to make digital language resources,
+is software that provides researchers the instruments to yield for their research;
+% without it a lot of research wouldbecome highly unfeasible or right-out impossible. 
+It is CLARIN's core mission to make digital language resources,
 including software, available to the wider research community.
@@
 example CLARIN centres.
+We demonstrate how LaMachine can create a fully functioning and standalone research environment for text mining  and NLP for Dutch texts in a use case project at the Healthcare Inspectorate.
 
@@
 
-Installation of LaMachine begins with a single bootstrap command\footnote{See
-\url{https://proycon.github.io/LaMachine}}.  It may interactively query the users for their software preferences, e.g.
-the flavour of LaMachine and the set of software to install, which is never static but can be customized by the user.
-The user may also opt for installing the latest releases, the more experimental development versions of the software, or
-specific custom versions (to facilitate scientific reproducibility). The bootstrap procedure detects and installs
-the necessary prerequisites automatically and eventually invokes Ansible to perform the bulk of the work.
-
-After a successful build, the user may interact with LaMachine either through the command line, which offers a standard
-shell and enables access to all lower-level tools and programming languages; or through his or her webbrowser, which
-when enabled offers a simple portal page towards all installed web-capable tools. Amongst these services is also a
-Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
-for e.g. Python and R that has gained great popularity in the scientific community.
-
-The current version of LaMachine comes with some simple data sharing facilities only: we provide a
-single shared dataspace between host and VM/container (where applicable), or allow simple upload. Extensive data
-search and management functions are deliberately beyond the scope of LaMachine, and left to more high-level tooling.
+%%iris: ik heb dit stukje geintegreerd in de  case study...
+%Installation of LaMachine begins with a single bootstrap command\footnote{See
+%\url{https://proycon.github.io/LaMachine}}.  It may interactively query the users for their software preferences, e.g.
+%the flavour of LaMachine and the set of software to install, which is never static but can be customized by the user.
+%The user may also opt for installing the latest releases, the more experimental development versions of the software, or
+%specific custom versions (to facilitate scientific reproducibility). The bootstrap procedure detects and installs
+%the necessary prerequisites automatically and eventually invokes Ansible to perform the bulk of the work.
+%
+%After a successful build, the user may interact with LaMachine either through the command line, which offers a standard
+%shell and enables access to all lower-level tools and programming languages; or through his or her webbrowser, which
+%when enabled offers a simple portal page towards all installed web-capable tools. Amongst these services is also a
+%Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
+%for e.g. Python and R that has gained great popularity in the scientific community.
+%
+%The current version of LaMachine comes with some simple data sharing facilities only: we provide a
+%single shared dataspace between host and VM/container (where applicable), or allow simple upload. Extensive data
+%search and management functions are deliberately beyond the scope of LaMachine, and left to more high-level tooling.
 
@@
 
+\section{Case study}\label{sec-case}
+
+We participated the ZonMw project entitled ``Tekstmining in het toezicht: een exploratieve studie naar de automatische verwerking van klachten ingediend bij het Landelijk Meldpunt Zorg"\footnote{project website: \url{https://www.zonmw.nl/nl/onderzoek-resultaten/kwaliteit-van-zorg/programmas/project-detail/effectief-toezicht/tekstmining-in-het-toezicht-een-exploratieve-studie-naar-de-automatische-verwerking-van-klachten-in/}.
+} led by IQhealthcare\footnote{\url{http://www.iqhealthcare.nl/nl/}}, the scientific center for healthcare quality of RadboudUMC.
+This project took place at the Dutch Health Inspectorate and aimed to apply text mining techniques to health care complaints that have been registered at the national contact point for health care (Landelijk Meldpunt Zorg \url{https://www.landelijkmeldpuntzorg.nl})
+We investigated the usefulness of text mining to categorise and cluster complaints, to automatically determine the severity of incoming complaints, to extract patterns  and to identify risk cases.
+This project turned out to be a good test case of the applicability and usefulness of the LaMachine software as a standalone research environment.
+As the complaint data is highly sensitive, this data could not leave the secure servers from the health inspectorate and was stored an environment without internet acces.
+We needed to bring the software to the data via a shared folder.
+
+We created a pre-compiled virtual image of LaMachine\footnote Installation of LaMachine begins with a single bootstrap command, see
+\url{https://proycon.github.io/LaMachine}} and succeeded in running it as a 64-bits Arch Linux-based VM inside another VM with Windows 2012 in which we did have administrative rights but no internet acces.
+In terms of hardware we ran on a machine with 8 cores and  32gb internal memory available.
+%using VirtualBox and Vagrnt on the Windows VM
+Lamachine provided a fully functional research environment and we ran all our expeirments within LaMachine. We interacted with LaMachine both through the command line, which offers a standard
+shell and enables access to all lower-level tools and programming languages; or through the (offline) webbrowser to apply the Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
+for e.g. Python and R that has gained great popularity in the scientific community.
+
+LaMachine comes with some simple data sharing facilities that allowed us to acces the sensitive complaint data via a  single shared dataspace between host and the VM.
+(Note that extensive data search and management functions are deliberately beyond the scope of LaMachine, and left to more high-level tooling.)
+
+We used many of the available tools in LaMachine within this project: Frog for linguistic annotation of the textual content of the complaints, Timbl and the scikit-learn Python package \cite{scikit-learn} for classification, T-scan for feature creation and colibri-core for n-gram analysis.
+
+%het doel van deze studie is om automatische tekstanalysetechnieken (tekstmining) toe te passen op de klachten die bij het Landelijk Meldpunt Zorg binnenkomen.
+Zo hopen we eventuele trends en patronen te vinden die meer inzicht geven in de aard en de relevantie van de klachten.
+De verwachting is om zo potentiële risicogevallen te ontdekken die op basis van de behandeling van individuele klachten nog niet waren gesignaleerd.
+
 \section{Conclusion \& Future work}
@@
 contribution. Contributor documentation has been written, and at this stage, we greatly welcome external participants
-to join in. Aside from the incorporation of new relevant software, the main objectives for the future are to provide
+to join in. Use cases as the example in section \ref{sec-case} contributing to thorough testing and running of LaMachine in less ideal circumstances such as nested VM constructions and offline usage.
+
+Aside from the incorporation of new relevant software, the main objectives for the future are to provide
 greater \emph{interoperability} between the included tools through better \emph{high-level interfaces} for the
@@
 
-\bibliographystyle{acl}
+\section*{Acknowlegdement}
+
+This research was funded by NWO CLARIN-NL, CLARIAH and the ZonMw project {\it Tekstmining in het toezicht: een exploratieve studie naar de automatische verwerking van klachten ingediend bij het Landelijk Meldpunt Zorg}, project number 516004614.
+We thank Tim Voets for his valuable contribution and help in the ZonMw project.
+
+\bibliographystyle{plain}
 \bibliography{lamachine}

### 2018-07-23 docs/lamachine_extabstract_clarin2018.tex
draft of Clarin paper -minor edits

@@
 
-We created a pre-compiled virtual image of LaMachine\footnote Installation of LaMachine begins with a single bootstrap command, see
+We created a pre-compiled virtual image of LaMachine\footnote{ Installation of LaMachine begins with a single bootstrap command, see
 \url{https://proycon.github.io/LaMachine}} and succeeded in running it as a 64-bits Arch Linux-based VM inside another VM with Windows 2012 in which we did have administrative rights but no internet acces.
@@
 %using VirtualBox and Vagrnt on the Windows VM
-Lamachine provided a fully functional research environment and we ran all our expeirments within LaMachine. We interacted with LaMachine both through the command line, which offers a standard
+Lamachine provided a fully functional research environment and we ran all our experiments within LaMachine. We interacted with LaMachine both through the command line, which offers a standard
 shell and enables access to all lower-level tools and programming languages; or through the (offline) webbrowser to apply the Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
@@
 This research was funded by NWO CLARIN-NL, CLARIAH and the ZonMw project {\it Tekstmining in het toezicht: een exploratieve studie naar de automatische verwerking van klachten ingediend bij het Landelijk Meldpunt Zorg}, project number 516004614.
-We thank Tim Voets for his valuable contribution and help in the ZonMw project.
+We thank all project partners: the Dutch Health Inspectorate, IQhealthcare, and Tim Voets for their valuable contributions and help in the ZonMw project.
 

### 2018-07-26 docs/lamachine_extabstract_clarin2018.tex
right stylesheet for CLARIN2018 paper
[paper] minor edits/fixes after iris' edits

@@
 \documentclass[a4paper,11pt]{article}
-%\usepackage{CLARIN2018}
+\usepackage{CLARIN2018}
 % - - - - - - - IMPORTANT - - - - - - - -
@@
 is software that provides researchers the instruments to yield for their research;
-% without it a lot of research wouldbecome highly unfeasible or right-out impossible. 
+% without it a lot of research would become highly unfeasible or right-out impossible.
 It is CLARIN's core mission to make digital language resources,
@@
 
-We participated the ZonMw project entitled ``Tekstmining in het toezicht: een exploratieve studie naar de automatische verwerking van klachten ingediend bij het Landelijk Meldpunt Zorg"\footnote{project website: \url{https://www.zonmw.nl/nl/onderzoek-resultaten/kwaliteit-van-zorg/programmas/project-detail/effectief-toezicht/tekstmining-in-het-toezicht-een-exploratieve-studie-naar-de-automatische-verwerking-van-klachten-in/}.
-} led by IQhealthcare\footnote{\url{http://www.iqhealthcare.nl/nl/}}, the scientific center for healthcare quality of RadboudUMC.
+We participated in the dutch ZonMw project entitled ``Tekstmining in het toezicht: een exploratieve studie naar de
+automatische verwerking van klachten ingediend bij het Landelijk Meldpunt Zorg"\footnote{project website:
+\url{https://www.zonmw.nl/nl/onderzoek-resultaten/kwaliteit-van-zorg/programmas/project-detail/effectief-toezicht/tekstmining-in-het-toezicht-een-exploratieve-studie-naar-de-automatische-verwerking-van-klachten-in/}.
+} led by IQhealthcare\footnote{\url{http://www.iqhealthcare.nl/nl/}}, the scientific centre for healthcare quality of RadboudUMC.
 This project took place at the Dutch Health Inspectorate and aimed to apply text mining techniques to health care complaints that have been registered at the national contact point for health care (Landelijk Meldpunt Zorg \url{https://www.landelijkmeldpuntzorg.nl})
-We investigated the usefulness of text mining to categorise and cluster complaints, to automatically determine the severity of incoming complaints, to extract patterns  and to identify risk cases.
-This project turned out to be a good test case of the applicability and usefulness of the LaMachine software as a standalone research environment.
-As the complaint data is highly sensitive, this data could not leave the secure servers from the health inspectorate and was stored an environment without internet acces.
-We needed to bring the software to the data via a shared folder.
+We investigated the usefulness of text mining to categorise and cluster complaints, to automatically determine the
+severity of incoming complaints, to extract patterns  and to identify risk cases. This project turned out to be a good
+test case of the applicability and usefulness of LaMachine as a standalone research environment.
+As the complaint data is highly sensitive, it could not leave the secure servers from the health inspectorate and was
+stored an environment without internet access. We needed to bring the software to the data via a shared folder.
 
-We created a pre-compiled virtual image of LaMachine\footnote{ Installation of LaMachine begins with a single bootstrap command, see
-\url{https://proycon.github.io/LaMachine}} and succeeded in running it as a 64-bits Arch Linux-based VM inside another VM with Windows 2012 in which we did have administrative rights but no internet acces.
-In terms of hardware we ran on a machine with 8 cores and  32gb internal memory available.
+We used a virtual machine (VM) image of LaMachine\footnote{Installation of LaMachine begins with a single bootstrap command, see
+\url{https://proycon.github.io/LaMachine}} and succeeded in running it as a 64-bits Linux-based VM inside another
+VM with Windows 2012 in which we did have administrative rights but no internet access, i.e. the VM that was offered to
+us for this project. In terms of hardware we ran on a
+machine with 8 cores and  32GB internal memory available.
 %using VirtualBox and Vagrnt on the Windows VM
 Lamachine provided a fully functional research environment and we ran all our experiments within LaMachine. We interacted with LaMachine both through the command line, which offers a standard
-shell and enables access to all lower-level tools and programming languages; or through the (offline) webbrowser to apply the Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
+shell and enables access to all lower-level tools and programming languages; or through the (offline) webbrowser to use the Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
 for e.g. Python and R that has gained great popularity in the scientific community.
 
-LaMachine comes with some simple data sharing facilities that allowed us to acces the sensitive complaint data via a  single shared dataspace between host and the VM.
+LaMachine comes with some simple data sharing facilities that allowed us to access the sensitive complaint data via a single shared dataspace between host and the VM.
 (Note that extensive data search and management functions are deliberately beyond the scope of LaMachine, and left to more high-level tooling.)
@@
 contribution. Contributor documentation has been written, and at this stage, we greatly welcome external participants
-to join in. Use cases as the example in section \ref{sec-case} contributing to thorough testing and running of LaMachine in less ideal circumstances such as nested VM constructions and offline usage.
+to join in. Use cases as the example in section \ref{sec-case} contribute to thorough testing and running of LaMachine in less ideal circumstances such as nested VM constructions and offline usage.
 
@@
 
-\section*{Acknowlegdement}
+\section*{Acknowledgement}
 

### 2018-07-26 docs/lamachine_extabstract_clarin2018.tex
[paper] processed reviewer 1 comments

@@
 
-Being an open-source NLP software distribution, LaMachine is constrained to POSIX-compliant platforms; this primarily
-means Linux, but also BSD and, with some restrictions, macOS. This stands in contrast with mobile platforms
-(Android/iOS/etc), native Windows/mac desktop software, or certain interface types in general such as classical desktop
-GUI applications, all of which fall beyond the scope of LaMachine. Software that is incorporated must 1) bear some
-relevance to NLP, 2) be under a recognised open-source license, 3) be deposited in a public version controlled
-repository\footnote{e.g. Github, Gitlab, Bitbucket} and 4) have a release protocol (with semantic versioning) using the
-proper technology-specific channels.
+Being an open-source NLP software distribution, LaMachine is constrained to Unix-like platforms; this primarily means
+Linux, but also BSD and, with some restrictions, macOS. Cygwin\footnote{A unix environment on Windows} is not tested or
+supported. This constaint stands in contrast with mobile platforms (Android/iOS/etc), native Windows/mac desktop software, or
+certain interface types in general such as classical desktop GUI applications, all of which fall beyond the scope of
+LaMachine. Software that is incorporated must 1) bear some relevance to NLP, 2) be under a recognised open-source
+license, 3) be deposited in a public version controlled repository\footnote{e.g. Github, Gitlab, Bitbucket, provided the
+repository is public} and 4) have
+a release protocol (with semantic versioning) using the proper technology-specific channels.
 

### 2018-07-26 docs/lamachine_extabstract_clarin2018.tex
[paper] minor edits

@@
 Linux, but also BSD and, with some restrictions, macOS. Cygwin\footnote{A unix environment on Windows} is not tested or
-supported. This constaint stands in contrast with mobile platforms (Android/iOS/etc), native Windows/mac desktop software, or
-certain interface types in general such as classical desktop GUI applications, all of which fall beyond the scope of
-LaMachine. Software that is incorporated must 1) bear some relevance to NLP, 2) be under a recognised open-source
-license, 3) be deposited in a public version controlled repository\footnote{e.g. Github, Gitlab, Bitbucket, provided the
-repository is public} and 4) have
-a release protocol (with semantic versioning) using the proper technology-specific channels.
+supported. This constraint stands in contrast with mobile platforms (Android/iOS/etc), native Windows/mac desktop
+software, or certain interface types in general such as classical desktop GUI applications, all of which fall beyond the
+scope of LaMachine. However, virtualisation technology enables deployment on a wider range of platforms, including
+Windows.  All software that is incorporated in LaMachine must 1) bear some relevance to NLP, 2) be under a recognised
+open-source license, 3) be deposited in a public version controlled repository\footnote{e.g. Github, Gitlab, Bitbucket,
+provided the repository is public} and 4) have a release protocol (with semantic versioning) using the proper
+technology-specific channels.
 
 LaMachine is a \emph{meta distribution} as it can be installed in various contexts. At its core, LaMachine consists of a
-set of machine-parsable instructions on how to obtain, install and configure software. These are implemented using
-Ansible\footnote{\url{https://www.ansible.com}}.  This is notably different from the more classical notion of Linux
-distributions, which generally provide their own repositories with (often binary) software packages. LaMachine builds on
-this already available infrastructure by taking these repositories as a given and only needs to know which
-repositories to use.  Similarly, there are different programming-language-specific ecosystems providing their own
-repositories, such as the Python Package Index\footnote{\url{https://pypi.org}} for Python,
+set of machine-parsable instructions on how to obtain, build (e.g. compile from source), install and configure software.
+These are implemented using Ansible\footnote{\url{https://www.ansible.com}}.  This is notably different from the more
+classical notion of Linux distributions, which generally provide their own repositories with (often binary) software
+packages. LaMachine builds on this already established infrastructure by taking these repositories as a given and only
+needs to know which repositories to use.  Similarly, there are different programming-language-specific ecosystems
+providing their own repositories, such as the Python Package Index\footnote{\url{https://pypi.org}} for Python,
 CRAN\footnote{\url{https://cran.r-project.org/}} for R, CPAN\footnote{\url{https://www.cpan.org}} for Perl, Maven
-Central\footnote{\url{https://search.maven.org}} for Java.  LaMachine again relies on those to pull and install
-software from and never forks, archives, or modifies the software in any way. In doing so, we compel participating
-software projects to adhere to well-established distribution standards and ensure the software is more sustainable
-towards the future \cite{softwarequality}. Moreover, we ensure that LaMachine never becomes a prerequisite for the software but merely a
-courtesy or convenience.
+Central\footnote{\url{https://search.maven.org}} for Java.  LaMachine again relies on those to pull and install software
+from and never forks, archives, or modifies the software in any way. In doing so, we compel participating software
+projects to adhere to well-established distribution standards and ensure the software is more sustainable towards the
+future \cite{softwarequality}. Moreover, we ensure that LaMachine never becomes a prerequisite for the software but
+merely a courtesy or convenience.
 
@@
 
-With LaMachine we also attempt to harmonise the metadata of all installed software, by converting metadata from upstream repositories to
+LaMachine we also aims to harmonise the metadata of all installed software, by converting metadata from upstream repositories to
 the CodeMeta standard\footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or encouraging software developers to provide
@@
 
-We used a virtual machine (VM) image of LaMachine\footnote{Installation of LaMachine begins with a single bootstrap command, see
-\url{https://proycon.github.io/LaMachine}} and succeeded in running it as a 64-bits Linux-based VM inside another
+We used a virtual machine (VM) image of LaMachine. Installation of LaMachine always begins with a single bootstrap
+command\footnote{see
+\url{https://proycon.github.io/LaMachine}} and we opted to build a VM and succeeded in running this 64-bits Linux-based VM inside another
 VM with Windows 2012 in which we did have administrative rights but no internet access, i.e. the VM that was offered to
@@
 Lamachine provided a fully functional research environment and we ran all our experiments within LaMachine. We interacted with LaMachine both through the command line, which offers a standard
-shell and enables access to all lower-level tools and programming languages; or through the (offline) webbrowser to use the Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
-for e.g. Python and R that has gained great popularity in the scientific community.
+shell and enables access to all lower-level tools and programming languages; and through the (offline) webbrowser to use the Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
+for e.g. Python and R that has gained great popularity in the scientific community. Moreover, LaMachine also offers a portal
+website with access and overview of all locally installed tools, including web services and web applications.
+%Dit was er nog niet in de use-case formeel gezien
 

### 2018-07-26 docs/lamachine_extabstract_clarin2018.tex
[paper] clarified 'upstream'

@@
 
-LaMachine we also aims to harmonise the metadata of all installed software, by converting metadata from upstream repositories to
+LaMachine also aims to harmonise the metadata of all installed software, by converting metadata from upstream
+repositories\footnote{i.e. repositories that where tool providers depositor their software} to
 the CodeMeta standard\footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or encouraging software developers to provide

### 2018-07-28 docs/lamachine_extabstract_clarin2018.tex
[paper] updates

@@
 direct remote provisioning through Ansible (for production servers), or an installation that is either global to the
-machine or local in a custom directory for a specific user. Pre-built docker containers and virtual machines with a
+machine or local in a custom directory for a specific user. Pre-built docker containers and virtual machine images with a
 limited selection of participating software are regularly uploaded to the Docker Hub and Vagrant Cloud, respectively.
 
-%%iris: ik heb dit stukje geintegreerd in de  case study...
-%Installation of LaMachine begins with a single bootstrap command\footnote{See
-%\url{https://proycon.github.io/LaMachine}}.  It may interactively query the users for their software preferences, e.g.
-%the flavour of LaMachine and the set of software to install, which is never static but can be customized by the user.
-%The user may also opt for installing the latest releases, the more experimental development versions of the software, or
-%specific custom versions (to facilitate scientific reproducibility). The bootstrap procedure detects and installs
-%the necessary prerequisites automatically and eventually invokes Ansible to perform the bulk of the work.
+%%iris: ik heb dit stukje geintegreerd in de  case study... (ik heb de eerste paragraaf toch nog even hier opgenomen)
+Installation of LaMachine begins with a single bootstrap command\footnote{See
+\url{https://proycon.github.io/LaMachine}}.  It may interactively query the users for their software preferences, e.g.
+the flavour of LaMachine and the set of software to install, which is never static but can be customized by the user.
+The user may also opt for installing the latest releases, the more experimental development versions of the software, or
+specific custom versions (to facilitate scientific reproducibility). The bootstrap procedure detects and installs
+the necessary prerequisites automatically and eventually invokes Ansible to perform the bulk of the work.
+
 %
@@
 
-We used a virtual machine (VM) image of LaMachine. Installation of LaMachine always begins with a single bootstrap
-command\footnote{see
-\url{https://proycon.github.io/LaMachine}} and we opted to build a VM and succeeded in running this 64-bits Linux-based VM inside another
-VM with Windows 2012 in which we did have administrative rights but no internet access, i.e. the VM that was offered to
-us for this project. In terms of hardware we ran on a
-machine with 8 cores and  32GB internal memory available.
+We used a virtual machine (VM) image of LaMachine and we ran this 64-bits Linux-based VM inside another
+VM with Windows 2012, provided to us by the health inspectorate for this project, in which we did have administrative rights but no internet access
+In terms of hardware we ran on a machine with 8 cores and  32GB internal memory available.
 %using VirtualBox and Vagrnt on the Windows VM
@@
 website with access and overview of all locally installed tools, including web services and web applications.
-%Dit was er nog niet in de use-case formeel gezien
+%Dit laatste was er nog niet in de use-case formeel gezien, maar wil het toch noemen
 
@@
 
-We used many of the available tools in LaMachine within this project: Frog for linguistic annotation of the textual content of the complaints, Timbl and the scikit-learn Python package \cite{scikit-learn} for classification, T-scan for feature creation and colibri-core for n-gram analysis.
+We used many of the available tools in LaMachine within this project: Frog for linguistic annotation of the textual
+content of the complaints, Timbl and the scikit-learn Python package \cite{scikit-learn} for classification, T-scan for
+feature extraction and colibri-core for n-gram analysis.
 

### 2018-07-29 docs/lamachine_extabstract_clarin2018.tex
[paper] minot update
final fixes (mostly language) before resubmission

@@
 scope of LaMachine. However, virtualisation technology enables deployment on a wider range of platforms, including
-Windows.  All software that is incorporated in LaMachine must 1) bear some relevance to NLP, 2) be under a recognised
+Windows.
+
+All software that is incorporated in LaMachine must 1) bear some relevance to NLP, 2) be under a recognised
 open-source license, 3) be deposited in a public version controlled repository\footnote{e.g. Github, Gitlab, Bitbucket,
@@
 LaMachine also aims to harmonise the metadata of all installed software, by converting metadata from upstream
-repositories\footnote{i.e. repositories that where tool providers depositor their software} to
-the CodeMeta standard\footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or encouraging software developers to provide
-their own. This in turn enables other tools to do proper service discovery and provenance logging.
+repositories\footnote{i.e. the repositories where tool providers deposit their software} to the CodeMeta
+standard\footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or
+encouraging software developers to provide their own. This in turn enables other tools to do proper service discovery
+and provenance logging. Leveraging this metadata, LaMachine comes with a webserver that offers a portal website with
+access and overview of all locally installed tools, including web services and web applications.
 
@@
 
-We participated in the dutch ZonMw project entitled ``Tekstmining in het toezicht: een exploratieve studie naar de
-automatische verwerking van klachten ingediend bij het Landelijk Meldpunt Zorg"\footnote{project website:
+We participated in the dutch ZonMw project titled \emph{``Tekstmining in het toezicht: een exploratieve studie naar de
+    automatische verwerking van klachten ingediend bij het Landelijk Meldpunt Zorg''} \footnote{project website:
 \url{https://www.zonmw.nl/nl/onderzoek-resultaten/kwaliteit-van-zorg/programmas/project-detail/effectief-toezicht/tekstmining-in-het-toezicht-een-exploratieve-studie-naar-de-automatische-verwerking-van-klachten-in/}.
 } led by IQhealthcare\footnote{\url{http://www.iqhealthcare.nl/nl/}}, the scientific centre for healthcare quality of RadboudUMC.
-This project took place at the Dutch Health Inspectorate and aimed to apply text mining techniques to health care complaints that have been registered at the national contact point for health care (Landelijk Meldpunt Zorg \url{https://www.landelijkmeldpuntzorg.nl})
+This project took place at the Dutch Health Inspectorate and aimed to apply text mining techniques to health care
+complaints that have been registered at the national contact point for health care (Landelijk Meldpunt
+Zorg\footnote{\url{https://www.landelijkmeldpuntzorg.nl}})
 We investigated the usefulness of text mining to categorise and cluster complaints, to automatically determine the
@@
 test case of the applicability and usefulness of LaMachine as a standalone research environment.
-As the complaint data is highly sensitive, it could not leave the secure servers from the health inspectorate and was
-stored an environment without internet access. We needed to bring the software to the data via a shared folder.
+As the complaint data is highly sensitive, it could not leave the secure servers of the health inspectorate and was
+stored in an environment without internet access. We needed to bring the software to the data via a shared folder.
 
 We used a virtual machine (VM) image of LaMachine and we ran this 64-bits Linux-based VM inside another
-VM with Windows 2012, provided to us by the health inspectorate for this project, in which we did have administrative rights but no internet access
-In terms of hardware we ran on a machine with 8 cores and  32GB internal memory available.
+VM with Windows 2012, provided to us by the health inspectorate for this project, in which we did have administrative
+rights but no internet access. In terms of hardware we ran on a machine with 8 cores and  32GB internal memory available.
 %using VirtualBox and Vagrnt on the Windows VM
-Lamachine provided a fully functional research environment and we ran all our experiments within LaMachine. We interacted with LaMachine both through the command line, which offers a standard
+LaMachine provided a fully functional research environment and we ran all our experiments within LaMachine. We interacted with LaMachine both through the command line, which offers a standard
 shell and enables access to all lower-level tools and programming languages; and through the (offline) webbrowser to use the Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
-for e.g. Python and R that has gained great popularity in the scientific community. Moreover, LaMachine also offers a portal
-website with access and overview of all locally installed tools, including web services and web applications.
-%Dit laatste was er nog niet in de use-case formeel gezien, maar wil het toch noemen
+for e.g. Python and R that has gained great popularity in the scientific community.
 

### 2018-08-06 roles/lamachine-core/templates/lamachine-activate.sh
force a proper UTF-8 locale for all LaMachine shells (not just webserver environments and the update script as was already the case) #107

@@ -28,6 +28,12 @@ if [[ "{{ ansible_distribution|lower }}" == "macosx" ]]; then
    export CLANG_CXX_LIBRARY="libc++" #needed for python bindings in lamachine-python-install
    export MACOSX_DEPLOYMENT_TARGET="{{ ansible_distribution_version }}"
 fi
+if [[ -z "$LANG" ]] || [[ "$LANG" != *"UTF-8"* ]]; then
+    #force a good UTF-8 locale
+    export LANG=en_US.UTF-8
+    export LANGUAGE=en_US.UTF-8
+    export LC_ALL=en_US.UTF-8
+fi
 if [[ "$LM_LOCALITY" == "local" ]]; then
     export LM_LOCAL_PREFIX={{local_prefix}}
     export LM_PREFIX={{local_prefix}}

### 2018-08-06 bootstrap.sh
version bump

@@ -20,7 +20,7 @@ boldblue=${bold}$(tput setaf 4) #  blue
 normal=$(tput sgr0)
 
 echo "${bold}=====================================================================${normal}"
-echo "           ,              ${bold}LaMachine v2.2.11${normal} - NLP Software distribution" #NOTE FOR DEVELOPER: also change version number in codemeta.json *AND* roles/lamachine-core/defaults/main.yml -> lamachine_version!
+echo "           ,              ${bold}LaMachine v2.2.12${normal} - NLP Software distribution" #NOTE FOR DEVELOPER: also change version number in codemeta.json *AND* roles/lamachine-core/defaults/main.yml -> lamachine_version!
 echo "          ~)                     (http://proycon.github.io/LaMachine)"
 echo "           (----í         Language Machines research group"
 echo "            /| |\         Centre of Language and Speech Technology"

### 2018-08-07 bootstrap.sh
added --nosysupdate parameter to bootstrap (nosysupdate: 1 in config) to skip updating global system packages, enabled by default only for travis

@@ -185,6 +185,7 @@ ANSIBLE_OPTIONS="-v"
 MINIMAL=0
 FORCE=0
 PREFER_DISTRO=0
+NOSYSUPDATE=0
 VMMEM=4096
 VAGRANTBOX="debian/contrib-stretch64" #base distribution for VM
 DOCKERREPO="proycon/lamachine"
@@ -327,6 +328,10 @@ while [[ $# -gt 0 ]]; do
         --force)
         FORCE="$2"
         shift
+        shift
+        ;;
+        --nosysupdate)
+        NOSYSUPDATE=1
         shift
         ;;
         --services)
@@ -1043,6 +1048,9 @@ webservertype: nginx #If set to anything different, the internal webserver will 
 " >> $STAGEDCONFIG
 if [ $FORCE -ne 0 ]; then
     echo "force: $FORCE #Sets the default force parameter for updates, set to 1 to force updates or 2 to explicitly remove all sources and start from scratch on each update. Remove this line entirely if you don't need it or are in doubt" >> $STAGEDCONFIG
+fi
+if [ $NOSYSUPDATE -ne 0 ]; then
+    echo "nosysupdate: $NOSYSUPDATE #Skips updating the global packages provided by the distribution" >> $STAGEDCONFIG
 fi
     if [[ $FLAVOUR == "local" ]] || [[ "$OS" == "mac" ]]; then
         echo "web_user: \"$USERNAME\"" >> $STAGEDCONFIG

### 2018-08-07 docs/lamachine_extabstract_clarin2018.tex

@@
 
-We participated in the dutch ZonMw project titled \emph{``Tekstmining in het toezicht: een exploratieve studie naar de
-    automatische verwerking van klachten ingediend bij het Landelijk Meldpunt Zorg''} \footnote{project website:
+We participated in a small Dutch national project titled \emph{`` Text mining for Inspection: an exploratory study on automatic analysis of health care complaints''} \footnote{project website:
 \url{https://www.zonmw.nl/nl/onderzoek-resultaten/kwaliteit-van-zorg/programmas/project-detail/effectief-toezicht/tekstmining-in-het-toezicht-een-exploratieve-studie-naar-de-automatische-verwerking-van-klachten-in/}.
-} led by IQhealthcare\footnote{\url{http://www.iqhealthcare.nl/nl/}}, the scientific centre for healthcare quality of RadboudUMC.
+} led by IQhealthcare\footnote{\url{http://www.iqhealthcare.nl/nl/}}, the scientific centre for healthcare quality of RadboudUMC hospital.
 This project took place at the Dutch Health Inspectorate and aimed to apply text mining techniques to health care
@@
 We used many of the available tools in LaMachine within this project: Frog for linguistic annotation of the textual
-content of the complaints, Timbl and the scikit-learn Python package \cite{scikit-learn} for classification, T-scan for
-feature extraction and colibri-core for n-gram analysis.
+content of the complaint and the scikit-learn Python package \cite{scikit-learn} for classification, T-scan \cite{tscan} for feature extraction in the form of text characteristics and colibri-core for n-gram analysis.
 
@@
 
-\bibliographystyle{plain}
 \bibliography{lamachine}

### 2018-08-09 bootstrap.sh
setting up dev branch again

@@ -513,9 +513,9 @@ fi
 
 if [ -z "$BRANCH" ]; then
     if [[ "$VERSION" == "development" ]]; then
-        BRANCH="custom"
+        BRANCH="develop"
     else
-        BRANCH="custom"
+        BRANCH="develop"
     fi
 fi
 if [ -z "$GITREPO" ]; then

### 2018-08-09 bootstrap.sh
setting up pipfix branch

@@ -513,9 +513,9 @@ fi
 
 if [ -z "$BRANCH" ]; then
     if [[ "$VERSION" == "development" ]]; then
-        BRANCH="develop"
+        BRANCH="pipfix"
     else
-        BRANCH="develop"
+        BRANCH="pipfix"
     fi
 fi
 if [ -z "$GITREPO" ]; then

### 2018-08-09 builds/test.py

@@ -49,7 +49,7 @@ def test(build, args):
         #we build local tests in a docker container (clean environment), the dockerfile invokes the bootstrap:
         cwd = os.getcwd()
         os.chdir("context/" + build['context'])
-        cmd = "docker build -t proycon/lamachine:" + build['name'] + " --build-arg NAME=" + build['name'] + " --build-arg VERSION=" + build['version'] + " --build-arg BRANCH=develop --build-arg HOSTNAME=lamachine-" + build['name'] + " . 2>> " + cwd + "/logs/" + buildid(build).replace(':','-') + '.log >&2'
+        cmd = "docker build -t proycon/lamachine:" + build['name'] + " --build-arg NAME=" + build['name'] + " --build-arg VERSION=" + build['version'] + " --build-arg BRANCH=pipfix --build-arg HOSTNAME=lamachine-" + build['name'] + " . 2>> " + cwd + "/logs/" + buildid(build).replace(':','-') + '.log >&2'
         r = os.system(cmd)
         os.chdir(cwd)
     else:

### 2018-08-09 bootstrap.sh
explicitly set --prefix with global python installations

@@ -20,7 +20,7 @@ boldblue=${bold}$(tput setaf 4) #  blue
 normal=$(tput sgr0)
 
 echo "${bold}=====================================================================${normal}"
-echo "           ,              ${bold}LaMachine v2.2.12${normal} - NLP Software distribution" #NOTE FOR DEVELOPER: also change version number in codemeta.json *AND* roles/lamachine-core/defaults/main.yml -> lamachine_version!
+echo "           ,              ${bold}LaMachine v2.2.13${normal} - NLP Software distribution" #NOTE FOR DEVELOPER: also change version number in codemeta.json *AND* roles/lamachine-core/defaults/main.yml -> lamachine_version!
 echo "          ~)                     (http://proycon.github.io/LaMachine)"
 echo "           (----í         Language Machines research group"
 echo "            /| |\         Centre of Language and Speech Technology"

### 2018-08-09 roles/lamachine-core/templates/lamachine-activate.sh
add explicit PYTHONPATH for global variant

@@ -48,6 +48,7 @@ if [[ "$LM_LOCALITY" == "local" ]]; then
     fi
 else
     export LM_PREFIX="{{global_prefix}}"
+    export PYTHONPATH="{{global_prefix}}"
     for f in $LM_PREFIX/bin/activate.d/*.sh; do
         if [ ! -z "$f" ]; then
             source $f

### 2018-08-09 roles/lamachine-core/templates/lamachine-activate.sh
fix for setting $PYTHONPATH, also added a $LM_PYTHONVERSION env variable

@@ -38,6 +38,9 @@ if [[ "$LM_LOCALITY" == "local" ]]; then
     export LM_LOCAL_PREFIX={{local_prefix}}
     export LM_PREFIX={{local_prefix}}
     export LM_OLD_PS1="$PS1"
+    {% if python_version is defined %}
+    export LM_PYTHONVERSION="{{python_version}}"
+    {% endif %}
     if [ -d $LM_LOCAL_PREFIX ]; then
         if [[ "$LM_LOCALENV_TYPE" == "conda" ]]; then
             source activate lamachine-{{conf_name}}
@@ -48,7 +51,10 @@ if [[ "$LM_LOCALITY" == "local" ]]; then
     fi
 else
     export LM_PREFIX="{{global_prefix}}"
-    export PYTHONPATH="{{global_prefix}}"
+    {% if python_version is defined %}
+    export PYTHONPATH="{{global_prefix}}/lib/python{{python_version}}/site-packages"
+    export LM_PYTHONVERSION="{{python_version}}"
+    {% endif %}
     for f in $LM_PREFIX/bin/activate.d/*.sh; do
         if [ ! -z "$f" ]; then
             source $f

### 2018-08-09 roles/lamachine-core/templates/lamachine-activate.sh
fix

@@ -39,7 +39,7 @@ if [[ "$LM_LOCALITY" == "local" ]]; then
     export LM_PREFIX={{local_prefix}}
     export LM_OLD_PS1="$PS1"
     {% if python_version is defined %}
-    export LM_PYTHONVERSION="{{python_version}}"
+    export LM_PYTHONVERSION="{{python_version.stdout}}"
     {% endif %}
     if [ -d $LM_LOCAL_PREFIX ]; then
         if [[ "$LM_LOCALENV_TYPE" == "conda" ]]; then
@@ -52,8 +52,8 @@ if [[ "$LM_LOCALITY" == "local" ]]; then
 else
     export LM_PREFIX="{{global_prefix}}"
     {% if python_version is defined %}
-    export PYTHONPATH="{{global_prefix}}/lib/python{{python_version}}/site-packages"
-    export LM_PYTHONVERSION="{{python_version}}"
+    export PYTHONPATH="{{global_prefix}}/lib/python{{python_version.stdout}}/site-packages"
+    export LM_PYTHONVERSION="{{python_version.stdout}}"
     {% endif %}
     for f in $LM_PREFIX/bin/activate.d/*.sh; do
         if [ ! -z "$f" ]; then

### 2018-08-09 roles/lamachine-core/templates/lamachine-activate.sh
add lib64 alternative to PYTHONPATH (used by CentOS for example)

@@ -52,7 +52,7 @@ if [[ "$LM_LOCALITY" == "local" ]]; then
 else
     export LM_PREFIX="{{global_prefix}}"
     {% if python_version is defined %}
-    export PYTHONPATH="{{global_prefix}}/lib/python{{python_version.stdout}}/site-packages"
+    export PYTHONPATH="{{global_prefix}}/lib64/python{{python_version.stdout}}/site-packages:{{global_prefix}}/lib/python{{python_version.stdout}}/site-packages"
     export LM_PYTHONVERSION="{{python_version.stdout}}"
     {% endif %}
     for f in $LM_PREFIX/bin/activate.d/*.sh; do

### 2018-09-10 docs/lamachine_extabstract_clarin2018.tex
[paper] minor fixes

@@
 CLARIN-NL project and its current successor CLARIAH. Special attention is paid to encouragement of good software
-development practises and reuse of established infrastructure in the scientific and open-source software development
+development practices and reuse of established infrastructure in the scientific and open-source software development
 community.
@@
 Linux, but also BSD and, with some restrictions, macOS. Cygwin\footnote{A unix environment on Windows} is not tested or
-supported. This constraint stands in contrast with mobile platforms (Android/iOS/etc), native Windows/mac desktop
-software, or certain interface types in general such as classical desktop GUI applications, all of which fall beyond the
-scope of LaMachine. However, virtualisation technology enables deployment on a wider range of platforms, including
-Windows.
+supported.  However, virtualisation technology enables deployment on a wider range of platforms, including Windows. The
+focus of the LaMachine distribution stands in contrast with mobile platforms (Android/iOS/etc), native Windows/mac
+desktop software, or certain interface types in general such as classical desktop GUI applications or mobile `apps', all
+of which fall beyond our scope.
 
@@
 We used a virtual machine (VM) image of LaMachine and we ran this 64-bits Linux-based VM inside another
-VM with Windows 2012, provided to us by the health inspectorate for this project, in which we did have administrative
+VM with Windows Server 2012, provided to us by the health inspectorate for this project, in which we did have administrative
 rights but no internet access. In terms of hardware we ran on a machine with 8 cores and  32GB internal memory available.

### 2018-09-10 docs/lamachine_extabstract_clarin2018.tex
[paper] updated Software section, less footnotes

@@
 encouraging software developers to provide their own. This in turn enables other tools to do proper service discovery
-and provenance logging. Leveraging this metadata, LaMachine comes with a webserver that offers a portal website with
-access and overview of all locally installed tools, including web services and web applications.
+and provenance logging.
+
+Leveraging this metadata, LaMachine comes with a webserver that offers a portal website with
+access and overview of all installed tools, including web services and web applications. It also comes with a
+Jupyter Lab\footnote{\url{https://jupyter.org/}} environment which provides a web-based Integrated Development
+Environment (IDE) for scripting in Python and R, web-based terminal access, and so-called \emph{notebooks} which mix
+text and code and have gained great popularity in data science community nowadays.
 
@@
 
-As LaMachine was conceived initially\footnote{LaMachine exists since May 2015 and has been used extensively ever since by
-numerous users} as the primary means of distribution of the software stack developed at CLST,
-Radboud University Nijmegen, it includes a lot of our software. A full list goes beyond the scope of this overview; we
-will merely mention some CLARIN-NL/CLARIAH-funded tools: ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format
-for Linguistic Annotation, with assorted tools), FLAT (a web-based linguistic Annotation tool), PICCL (an OCR and
-post-OCR correction pipeline) and CLAM. We already include some relevant software by the University of Groningen and VU
-Amsterdam as well. Moreover, LaMachine incorporates a large number of renowned tools by external international parties,
-offering most notably a mature Python environment\footnote{scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch,
-keras, gensim, etc...}, but also R and Java.
+As LaMachine was conceived initially as the primary means of distribution of the software stack developed at CLST, Radboud University
+Nijmegen, it includes a lot of our software. LaMachine exists since May 2015 and has been used extensively ever since
+by numerous users. A full list of included software goes beyond the scope of this overview; we will merely mention
+some CLARIN-NL/CLARIAH-funded tools: ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format for Linguistic
+Annotation, with assorted tools), FLAT (a web-based linguistic Annotation tool), PICCL (an OCR and post-OCR correction
+pipeline) and CLAM. We already include some relevant software by the University of Groningen (such as Alpino, a
+dependency parser) and VU Amsterdam as well. Moreover, LaMachine incorporates a large number of renowned tools by
+external international parties, offering most notably a mature Python environment with renowned scientific modules such
+as scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch, keras, gensim, and many others, but also R, Java and tools
+such as Stanford CoreNLP.
 
@@
 LaMachine provided a fully functional research environment and we ran all our experiments within LaMachine. We interacted with LaMachine both through the command line, which offers a standard
-shell and enables access to all lower-level tools and programming languages; and through the (offline) webbrowser to use the Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
-for e.g. Python and R that has gained great popularity in the scientific community.
+shell and enables access to all lower-level tools and programming languages; and through the (offline) webbrowser to use
+the Jupyter Notebook environment.
 

### 2018-09-10 docs/lamachine_extabstract_clarin2018.tex
[paper] reworded beginning of software section

@@
 
-As LaMachine was conceived initially as the primary means of distribution of the software stack developed at CLST, Radboud University
-Nijmegen, it includes a lot of our software. LaMachine exists since May 2015 and has been used extensively ever since
-by numerous users. A full list of included software goes beyond the scope of this overview; we will merely mention
+LaMachine exists since May 2015 and has been used extensively ever since by numerous users. It was initiallly conceived
+as the primary means of distribution of the software stack developed at CLST, Radboud University Nijmegen, it includes a
+lot of our software.  A full list of included software goes beyond the scope of this overview; we will merely mention
 some CLARIN-NL/CLARIAH-funded tools: ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format for Linguistic
@@
 external international parties, offering most notably a mature Python environment with renowned scientific modules such
-as scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch, keras, gensim, and many others, but also R, Java and tools
-such as Stanford CoreNLP.
+as scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch, keras, gensim, and many others, but also R, Java and
+tools such as Stanford CoreNLP.
 

### 2018-09-10 docs/lamachine_extabstract_clarin2018.tex
[paper] changed relation with VRE project

@@
 
-Development of LaMachine presently takes place in the scope of the CLARIAH WP3 Virtual Research Environment (VRE)
+Development of LaMachine presently takes place in collaboration with the CLARIAH WP3 Virtual Research Environment (VRE)
 project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}}, which has higher ambitions in

### 2018-09-10 docs/lamachine_extabstract_clarin2018.tex
[paper] further updates

@@
 flexibility with regard to the target platform, where we support several major GNU/Linux distributions (Debian, Ubuntu,
-CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as macOS. Second, there is flexibility with
-regard to the form, where we support \emph{containerisation} through Docker\footnote{\url{https://www.docker.com}},
-\emph{virtualisation} through Vagrant\footnote{\url{https://vagrant.org}} and VirtualBox\footnote{\url{https://www.virtualbox.org}},
-direct remote provisioning through Ansible (for production servers), or an installation that is either global to the
-machine or local in a custom directory for a specific user. Pre-built docker containers and virtual machine images with a
-limited selection of participating software are regularly uploaded to the Docker Hub and Vagrant Cloud, respectively.
+CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as macOS (although with more limitations). Second, there is
+flexibility with regard to the form, where we support \emph{containerisation} through
+Docker\footnote{\url{https://www.docker.com}}, \emph{virtualisation} through Vagrant\footnote{\url{https://vagrant.org}}
+and VirtualBox\footnote{\url{https://www.virtualbox.org}}, direct remote provisioning through Ansible (for production
+servers), or an installation that is either global to the machine or local in a custom directory for a specific user
+(using \texttt{virtualenv}). Pre-built docker containers and virtual machine images with a limited selection of participating software are regularly
+uploaded to the Docker Hub and Vagrant Cloud, respectively. The different flavours all offer a different degree of
+seperation from the host OS, where Virtual Machines are completely virtualised, Docker Containers still share the kernel
+with the host OS, and the machine-specific installation flavour actually compiles against the machine's distribution itself and thus
+offers the least amount of overhead.
 
@@
 LaMachine also aims to harmonise the metadata of all installed software, by converting metadata from upstream
-repositories\footnote{i.e. the repositories where tool providers deposit their software} to the CodeMeta
-standard\footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or
-encouraging software developers to provide their own. This in turn enables other tools to do proper service discovery
-and provenance logging.
+repositories, i.e. the repositories where tool providers deposit their software, to a common standard called CodeMeta
+\footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or
+encouraging software developers to provide their codemeta metadata inside their source code repositories and using
+that directly. This in turn enables other tools to do proper service discovery and provenance logging.
 
@@
 Environment (IDE) for scripting in Python and R, web-based terminal access, and so-called \emph{notebooks} which mix
-text and code and have gained great popularity in data science community nowadays.
+text, code and data output and have gained great popularity in data science community nowadays.
 
@@
 
-LaMachine exists since May 2015 and has been used extensively ever since by numerous users. It was initiallly conceived
-as the primary means of distribution of the software stack developed at CLST, Radboud University Nijmegen, it includes a
-lot of our software.  A full list of included software goes beyond the scope of this overview; we will merely mention
-some CLARIN-NL/CLARIAH-funded tools: ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format for Linguistic
-Annotation, with assorted tools), FLAT (a web-based linguistic Annotation tool), PICCL (an OCR and post-OCR correction
-pipeline) and CLAM. We already include some relevant software by the University of Groningen (such as Alpino, a
-dependency parser) and VU Amsterdam as well. Moreover, LaMachine incorporates a large number of renowned tools by
-external international parties, offering most notably a mature Python environment with renowned scientific modules such
-as scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch, keras, gensim, and many others, but also R, Java and
-tools such as Stanford CoreNLP.
+LaMachine exists since May 2015 and has been used extensively ever since by numerous users, in 2018 version 2 was
+released which was a significant rewrite. LaMachine was initially conceived as the primary means of distribution of the
+software stack developed at CLST, Radboud University Nijmegen. It therefore includes a lot of our software.  A full list
+of included software goes beyond the scope of this overview; we will merely mention some CLARIN-NL/CLARIAH-funded tools:
+ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format for Linguistic Annotation, with assorted tools), FLAT
+(a web-based linguistic Annotation tool), PICCL (an OCR and post-OCR correction pipeline) and CLAM. However, this
+project is not limited to one research group and is open to participation by other software providers, especially those
+also in CLARIAH and the upcoming CLARIAH PLUS project. We already include some relevant software by the University of
+Groningen (such as Alpino, a dependency parser) and VU Amsterdam. Moreover, LaMachine incorporates a large number of
+renowned tools by external international parties, offering most notably a mature Python environment with renowned
+scientific modules such as scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch, keras, gensim, and many others,
+but also R, Java and tools such as Stanford CoreNLP.
 

### 2018-09-10 docs/lamachine_extabstract_clarin2018.tex
[paper] mentioned kaldi

@@
 renowned tools by external international parties, offering most notably a mature Python environment with renowned
-scientific modules such as scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch, keras, gensim, and many others,
-but also R, Java and tools such as Stanford CoreNLP.
+scientific modules such as scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch, keras, gensim, tensorflow, and many others,
+but also R, Java and tools such as Stanford CoreNLP and Kaldi.
 

### 2018-09-10 docs/lamachine_extabstract_clarin2018.tex
[paper] added screenshot of the portal and lab environment, removed some footnotes and references for scikit-learn and tscan to make space

@@
 needs to know which repositories to use.  Similarly, there are different programming-language-specific ecosystems
-providing their own repositories, such as the Python Package Index\footnote{\url{https://pypi.org}} for Python,
-CRAN\footnote{\url{https://cran.r-project.org/}} for R, CPAN\footnote{\url{https://www.cpan.org}} for Perl, Maven
-Central\footnote{\url{https://search.maven.org}} for Java.  LaMachine again relies on those to pull and install software
-from and never forks, archives, or modifies the software in any way. In doing so, we compel participating software
-projects to adhere to well-established distribution standards and ensure the software is more sustainable towards the
-future \cite{softwarequality}. Moreover, we ensure that LaMachine never becomes a prerequisite for the software but
-merely a courtesy or convenience.
+providing their own repositories, such as the Python Package Index for Python, CRAN for R, CPAN for Perl, Maven Central
+for Java.  LaMachine again relies on those to pull and install software from and never forks, archives, or modifies the
+software in any way. In doing so, we compel participating software projects to adhere to well-established distribution
+standards and ensure the software is more sustainable towards the future \cite{softwarequality}. Moreover, we ensure
+that LaMachine never becomes a prerequisite for the software but merely a courtesy or convenience.
 
@@
 flexibility with regard to the form, where we support \emph{containerisation} through
-Docker\footnote{\url{https://www.docker.com}}, \emph{virtualisation} through Vagrant\footnote{\url{https://vagrant.org}}
-and VirtualBox\footnote{\url{https://www.virtualbox.org}}, direct remote provisioning through Ansible (for production
+Docker\footnote{\url{https://www.docker.com}}, \emph{virtualisation} through Vagrant
+and VirtualBox\footnote{\url{https://vagrant.org}, \url{https://www.virtualbox.org}}, direct remote provisioning through Ansible (for production
 servers), or an installation that is either global to the machine or local in a custom directory for a specific user
@@
 text, code and data output and have gained great popularity in data science community nowadays.
+
+\begin{figure}[htb] \begin{center}
+\includegraphics[width=160.0mm]{doublescreenshot.jpg}
+\end{center}
+\caption{Screenshot excerpts of the Portal Web-app inside LaMachine (left) and the Jupyter Lab environment (right)}
+\label{fig:arch}
+\end{figure}
 
@@
 We used many of the available tools in LaMachine within this project: Frog for linguistic annotation of the textual
-content of the complaint and the scikit-learn Python package \cite{scikit-learn} for classification, T-scan \cite{tscan} for feature extraction in the form of text characteristics and colibri-core for n-gram analysis.
+content of the complaint and the scikit-learn Python package for classification, T-scan for feature extraction in the form of text characteristics and colibri-core for n-gram analysis.
 

### 2018-09-10 docs/lamachine_extabstract_clarin2018.tex
removed picture
added architecture schema to paper

@@
 Installation of LaMachine begins with a single bootstrap command\footnote{See
-\url{https://proycon.github.io/LaMachine}}.  It may interactively query the users for their software preferences, e.g.
-the flavour of LaMachine and the set of software to install, which is never static but can be customized by the user.
-The user may also opt for installing the latest releases, the more experimental development versions of the software, or
-specific custom versions (to facilitate scientific reproducibility). The bootstrap procedure detects and installs
-the necessary prerequisites automatically and eventually invokes Ansible to perform the bulk of the work.
+\url{https://proycon.github.io/LaMachine}}.  It can interactively query the users for their software preferences
+\emph{(stored as the host configuration)}, e.g. the flavour of LaMachine, as well as the set of software to install,
+\emph{the installation manifest}. This set is never static but can be customized by the user.  The user may also opt for installing the latest releases, the more
+experimental development versions of the software, or specific custom versions (to facilitate scientific
+reproducibility). The bootstrap procedure detects and installs the necessary prerequisites automatically and eventually
+invokes Ansible to perform the bulk of the work.  Figure~\ref{fig:arch} provides a schematic view.
 
-%
 %After a successful build, the user may interact with LaMachine either through the command line, which offers a standard
@@
 %search and management functions are deliberately beyond the scope of LaMachine, and left to more high-level tooling.
+
+\begin{figure}[htb] \begin{center}
+\includegraphics[width=135.0mm]{architecture.png}
+\end{center}
+\caption{\footnotesize{A schematic representation of the LaMachine architecture}}
+\label{fig:arch}
+\end{figure}
 
@@
 
-\begin{figure}[htb] \begin{center}
-\includegraphics[width=160.0mm]{doublescreenshot.jpg}
-\end{center}
-\caption{Screenshot excerpts of the Portal Web-app inside LaMachine (left) and the Jupyter Lab environment (right)}
-\label{fig:arch}
-\end{figure}
-
 \section{Software}
@@
 project is not limited to one research group and is open to participation by other software providers, especially those
-also in CLARIAH and the upcoming CLARIAH PLUS project. We already include some relevant software by the University of
-Groningen (such as Alpino, a dependency parser) and VU Amsterdam. Moreover, LaMachine incorporates a large number of
-renowned tools by external international parties, offering most notably a mature Python environment with renowned
-scientific modules such as scipy, numpy, scikit-learn, matplotlib, nltk, spacy, pytorch, keras, gensim, tensorflow, and many others,
-but also R, Java and tools such as Stanford CoreNLP and Kaldi.
+also in CLARIAH and the upcoming CLARIAH PLUS project. We already include some relevant software by other CLARIAH
+partners. Moreover, LaMachine incorporates a large number of renowned tools by external international parties, offering
+most notably a mature Python environment with renowned scientific modules such as scipy, numpy, scikit-learn,
+matplotlib, nltk, spacy, pytorch, keras, gensim, tensorflow, and many others, but also R, Java and tools such as
+Stanford CoreNLP and Kaldi.
 
@@
 
-LaMachine comes with some simple data sharing facilities that allowed us to access the sensitive complaint data via a single shared dataspace between host and the VM.
-(Note that extensive data search and management functions are deliberately beyond the scope of LaMachine, and left to more high-level tooling.)
+LaMachine comes with some simple data sharing facilities that allowed us to access the sensitive complaint data via a
+single shared dataspace between host and the VM. Extensive data search and management functions are deliberately beyond the scope of LaMachine, and left to more high-level tooling.
 

### 2018-09-10 docs/lamachine_extabstract_clarin2018.tex
[paper] link shortener

@@
 
-We participated in a small Dutch national project titled \emph{`` Text mining for Inspection: an exploratory study on automatic analysis of health care complaints''} \footnote{project website:
-\url{https://www.zonmw.nl/nl/onderzoek-resultaten/kwaliteit-van-zorg/programmas/project-detail/effectief-toezicht/tekstmining-in-het-toezicht-een-exploratieve-studie-naar-de-automatische-verwerking-van-klachten-in/}.
-} led by IQhealthcare\footnote{\url{http://www.iqhealthcare.nl/nl/}}, the scientific centre for healthcare quality of RadboudUMC hospital.
+We participated in a small Dutch national project titled \emph{``Text mining for Inspection: an exploratory study on
+automatic analysis of health care complaints''} \footnote{\url{https://bit.ly/2N2AICS}} %links to https://www.zonmw.nl/nl/onderzoek-resultaten/kwaliteit-van-zorg/program/project-detail/effectief-toezicht/tekstmining-in-het-toezicht-een-exploratieve-studie-naar-de-automatische-verwerking-van-klachten-in/}.
+led by IQhealthcare\footnote{\url{http://www.iqhealthcare.nl/nl/}}, the scientific centre for healthcare quality of RadboudUMC hospital.
 This project took place at the Dutch Health Inspectorate and aimed to apply text mining techniques to health care

### 2019-01-14 bootstrap.sh
minor update

@@ -954,10 +954,9 @@ for package in ${NEED[@]}; do
     elif [ "$package" = "virtualenv" ]; then
         if [ "$OS" = "debian" ]; then
             cmd="sudo apt-get $NONINTERACTIVEFLAGS install python-virtualenv"
-            if [ "$DISTRIB_RELEASE" != "14.04" ]; then #except on old ubuntu
+            if [[ "$DISTRIB_RELEASE" != "14.04" ]]; then #except on old ubuntu
                 cmd="$cmd virtualenv"
             fi
-
         elif [ "$OS" = "redhat" ]; then
             cmd="sudo yum  $NONINTERACTIVEFLAGS install python-virtualenv"
         elif [ "$OS" = "arch" ]; then

### 2019-01-14 docs/lamachine_clarin2018.tex
install virtualenv (and starting paper)

--- 2018-04-12 docs/lamachine_clarin2018.tex
@@
 % The next three lines allow XeLaTeX, graphics import and hyperlinks, set font and language
-\usepackage{xltxtra,polyglossia,graphicx,hyperref}
-\setmainfont[Mapping=tex-text]{Times}
-\setdefaultlanguage{english}
+%\usepackage{xltxtra,polyglossia,graphicx,hyperref}
+%\setmainfont[Mapping=tex-text]{Times}
+%\setdefaultlanguage{english}
 % If for some reason the above three lines are not compatible with your LaTeX installation,
 % comment out the above three instructions and uncomment the following four instead:
-%\usepackage{times}
-%\usepackage{url}
+\usepackage{times}
+\usepackage{hyperref}
+\usepackage{graphicx}
 %\usepackage{latexsym}
-%\usepackage[english]{babel}
+\usepackage[english]{babel}
 
@@
 
-\usepackage{covington} % if needed, for linguistic examples
-
-\title{LaMachine NLP Software Distribution}
+\title{LaMachine: A meta-distribution for NLP software}
 
@@
 
-\author{Maarten van Gompel \\
-  Centre for Language and Speech Technology \\
-  Radboud University Nijmegen, the Netherlands \\
-  {\tt proycon@anaproy.nl} \\ %\And % if needed: this makes a second column
+\author{Maarten van Gompel  and Iris Hendrickx\\
+  Centre for Language and Speech Technology (CLST) \\
+  Radboud University, Nijmegen, the Netherlands \\
+  {\tt proycon@anaproy.nl, i.hendrickx@let.ru.nl} \\ %\And % if needed: this makes a second column
+  \url{https://proycon.github.io/LaMachine}
 %  Second Author \\
@@
 \begin{abstract}
-We introduce LaMachine, a unified Natural Language Processing (NLP) open-source
-software distribution to facilitate the installation and deployment of a large
-amount of software projects that have been developed in the scope of the
-CLARIN-NL project and its current successor CLARIAH. Special attention is paid
-to encouragement of good software development practises.
+We introduce LaMachine, a unified Natural Language Processing (NLP) open-source software distribution to facilitate the
+installation and deployment of a large amount of software projects that have been developed in the scope of the
+CLARIN-NL project and its current successor CLARIAH. Special attention is paid to encouragement of good software
+development practices and reuse of established infrastructure in the scientific and open-source software development
+community.
+We illustrate the usage of LaMachine in an exploratory text mining project at the Dutch Health Inspectorate where LaMachine was applied to create a research environment for automatic text analysis for health care quality monitoring.
 \end{abstract}
@@
 
-Software is a key deliverable and a vital component for research in projects
-such as CLARIN. It is software that provides researchers the instruments to
-yield for their research, and without it a lot of research would become highly
-unfeasible or right-out impossible. It is CLARIN's core mission to make digital
-language resources, including software, available to the wider research
-community.
+Software is a key deliverable and a vital component for research in projects such as those under the CLARIN umbrella. It
+is software that provides researchers the instruments to yield for their research; It is CLARIN's core mission to make
+digital language resources, including software, available to the wider research community.
 
-Software is a broad but well-ingrained notion in society nowadays, referring to
-any form of computer program. This can manifest in various forms, whether it is
-an app on a phone, a graphical desktop application, a command line tool,  or a
-webpage.
+We see that NLP software often takes on complex forms such as processing pipelines invoking various individual
+components, which in turn rely on various dependencies. Add dedicated web-interfaces on top of that and you obtain a
+suite of interconnected software that is often non-trivial to install, configure, and deploy. This is where LaMachine
+comes in.
 
-one classification could be
-made based on the \emph{interfaces} software offers.
+Software is a broad but well-ingrained notion in society nowadays, referring to any form of computer program. This can
+manifest in various forms, whether it is an app on a phone, a graphical desktop application, a command line tool, or a
+web-based application. These different \emph{interfaces} generally address different \emph{audiences}; the data
+scientist will feel at home at the command line and in scripting environments and uses these fairly low-level
+interfaces, whereas researchers in the humanities demand higher-level interfaces. There is often a power trade-off
+between lower and higher-level interfaces, with the former providing maximum flexibility at the cost of a steeper
+learning curve, technical ability, and a do-it-yourself mentality. Higher level interfaces, on the other
+hand, expose certain functionality of the software in an easy and accessible way, but in doing so often can not expose
+the full power of the software. The cost trade-off is also apparant in the construction of the interfaces, where
+high-level interfaces are typically far more costly to build.
 
-We too often see that the research community develops valuable
+LaMachine incorporates software providing different types of interfaces\footnote{Command line interfaces, programming
+interfaces, web-user interfaces, webservices.} which, as seen above, typically address different audiences. Whilst we
+attempt to accommodate both technical\footnote{Data scientists, DevOps, system administrators, developers.} and
+less-technical audiences\footnote{The wider researcher community, particularly the Humanities; also educational
+settings.}, there is a natural bias towards the former as lower-level interfaces are often a prerequisite to build
+higher-level interfaces on. Depending on the \emph{flavour} of LaMachine chosen, it makes a good virtual research
+environment for a data scientist, whether on a personal computer or on a computing cluster, a good development
+environment for a developer or a good deployment method for production servers in for example CLARIN centres.  We
+demonstrate how LaMachine can create a fully functioning and standalone research environment for text mining  and NLP
+for Dutch texts in a use case project at the Healthcare Inspectorate.
+
+\section{Architecture}
+
+LaMachine is an open-source NLP software distribution. This implies the following in our case:
+
+\begin{enumerate}
+    \item LaMachine facilitates the installation, distribution and configuration of software. It does not fork, modify
+        or appropriate the participating software in any way, nor does it provide a hosting place or repository for
+        software.
+    \item The software included in LaMachine must under a recognised (i.e. OSI-approved) open-source license.
+    \item The source
+        code must be hosted in a public version controlled repository\footnote{e.g. Github, Gitlab, Bitbucket, provided
+        the respository is public}
+    \item There must be build process to compile the source, if applicable, and install the program
+    \item There must be some kind of release protocol (with semantic versioning) using the proper
+technology-specific channels.
+    \item All software that is incorporated in LaMachine must bear at least some relevance to the field of Natural Language Processing.
+    \item LaMachine lives in an open-source ecosystem and therefore builds on Unix-like platforms; this primarily
+        means Linux, as well as BSD and, with some restrictions, macOS. This by definition excludes certain software for different platforms, such as
+        mobile platforms (Android/iOS/etc), native Windows/mac desktop applications, or certain interface types in
+        general such as classical desktop GUI applications or mobile `apps', all of which fall beyond our scope.
+        Cygwin\footnote{A unix environment on Windows} is not tested or
+        supported.  However, virtualisation technology enables deployment on a wider range of platforms, including Windows.
+\end{enumerate}
+
+LaMachine is a \emph{meta distribution} as it can be installed in various contexts. At its core, LaMachine consists of a
+set of machine-parsable instructions on how to obtain, build (e.g. compile from source), install and configure software.
+These are implemented using Ansible\footnote{\url{https://www.ansible.com}}.  This is notably different from the more
+classical notion of Linux distributions, which generally provide their own repositories with (often binary) software
+packages. LaMachine builds on this already established infrastructure by taking these repositories as a given and only
+needs to know which repositories to use.  Similarly, there are different programming-language-specific ecosystems
+providing their own repositories, such as the Python Package Index for Python, CRAN for R, CPAN for Perl, Maven Central
+for Java.  LaMachine again relies on those to pull and install software from and never forks, archives, or modifies the
+software in any way. In doing so, we compel participating software projects to adhere to well-established distribution
+standards and ensure the software is more sustainable towards the future \cite{softwarequality}. Moreover, we ensure
+that LaMachine never becomes a prerequisite for the software but merely a courtesy or convenience.
+
+LaMachine provides ample flexibility that allows it to be deployable in different contexts. First of all there is
+flexibility with regard to the target platform, where we support several major GNU/Linux distributions (Debian, Ubuntu,
+CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as macOS (although with more limitations). Second, there is
+flexibility with regard to the form, where we support \emph{containerisation} through
+Docker\footnote{\url{https://www.docker.com}}, \emph{virtualisation} through Vagrant
+and VirtualBox\footnote{\url{https://vagrant.org}, \url{https://www.virtualbox.org}}, direct remote provisioning through Ansible (for production
+servers), or an installation that is either global to the machine or local in a custom directory for a specific user
+(using \texttt{virtualenv}). Pre-built docker containers and virtual machine images with a limited selection of participating software are regularly
+uploaded to the Docker Hub and Vagrant Cloud, respectively. The different flavours all offer a different degree of
+seperation from the host OS, where Virtual Machines are completely virtualised, Docker Containers still share the kernel
+with the host OS, and the machine-specific installation flavour actually compiles against the machine's distribution itself and thus
+offers the least amount of overhead.
+
+%%iris: ik heb dit stukje geintegreerd in de  case study... (ik heb de eerste paragraaf toch nog even hier opgenomen)
+Installation of LaMachine begins with a single bootstrap command\footnote{See
+\url{https://proycon.github.io/LaMachine}}.  It can interactively query the users for their software preferences
+\emph{(stored as the host configuration)}, e.g. the flavour of LaMachine, as well as the set of software to install,
+\emph{the installation manifest}. This set is never static but can be customized by the user.  The user may also opt for installing the latest releases, the more
+experimental development versions of the software, or specific custom versions (to facilitate scientific
+reproducibility). The bootstrap procedure detects and installs the necessary prerequisites automatically and eventually
+invokes Ansible to perform the bulk of the work.  Figure~\ref{fig:arch} provides a schematic view.
+
+%After a successful build, the user may interact with LaMachine either through the command line, which offers a standard
+%shell and enables access to all lower-level tools and programming languages; or through his or her webbrowser, which
+%when enabled offers a simple portal page towards all installed web-capable tools. Amongst these services is also a
+%Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
+%for e.g. Python and R that has gained great popularity in the scientific community.
+%
+%The current version of LaMachine comes with some simple data sharing facilities only: we provide a
+%single shared dataspace between host and VM/container (where applicable), or allow simple upload. Extensive data
+%search and management functions are deliberately beyond the scope of LaMachine, and left to more high-level tooling.
+
+\begin{figure}[htb] \begin{center}
+\includegraphics[width=135.0mm]{architecture.png}
+\end{center}
+\caption{\footnotesize{A schematic representation of the LaMachine architecture}}
+\label{fig:arch}
+\end{figure}
+
+LaMachine also aims to harmonise the metadata of all installed software, by converting metadata from upstream
+repositories, i.e. the repositories where tool providers deposit their software, to a common standard called CodeMeta
+\footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or
+encouraging software developers to provide their codemeta metadata inside their source code repositories and using
+that directly. This in turn enables other tools to do proper service discovery and provenance logging.
+
+Leveraging this metadata, LaMachine comes with a webserver that offers a portal website with
+access and overview of all installed tools, including web services and web applications. It also comes with a
+Jupyter Lab\footnote{\url{https://jupyter.org/}} environment which provides a web-based Integrated Development
+Environment (IDE) for scripting in Python and R, web-based terminal access, and so-called \emph{notebooks} which mix
+text, code and data output and have gained great popularity in data science community nowadays.
+
+\section{Software}
+
+LaMachine exists since May 2015 and has been used extensively ever since by numerous users, in 2018 version 2 was
+released which was a significant rewrite. LaMachine was initially conceived as the primary means of distribution of the
+software stack developed at CLST, Radboud University Nijmegen. It therefore includes a lot of our software.  A full list
+of included software goes beyond the scope of this overview; we will merely mention some CLARIN-NL/CLARIAH-funded tools:
+ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format for Linguistic Annotation, with assorted tools), FLAT
+(a web-based linguistic Annotation tool), PICCL (an OCR and post-OCR correction pipeline) and CLAM. However, this
+project is not limited to one research group and is open to participation by other software providers, especially those
+also in CLARIAH and the upcoming CLARIAH PLUS project. We already include some relevant software by other CLARIAH
+partners. Moreover, LaMachine incorporates a large number of renowned tools by external international parties, offering
+most notably a mature Python environment with renowned scientific modules such as scipy, numpy, scikit-learn,
+matplotlib, nltk, spacy, pytorch, keras, gensim, tensorflow, and many others, but also R, Java and tools such as
+Stanford CoreNLP and Kaldi.
+
+\section{Case study}\label{sec-case}
+
+We participated in a small Dutch national project titled \emph{``Text mining for Inspection: an exploratory study on
+automatic analysis of health care complaints''} \footnote{\url{https://bit.ly/2N2AICS}} %links to https://www.zonmw.nl/nl/onderzoek-resultaten/kwaliteit-van-zorg/program/project-detail/effectief-toezicht/tekstmining-in-het-toezicht-een-exploratieve-studie-naar-de-automatische-verwerking-van-klachten-in/}.
+led by IQhealthcare\footnote{\url{http://www.iqhealthcare.nl/nl/}}, the scientific centre for healthcare quality of RadboudUMC hospital.
+This project took place at the Dutch Health Inspectorate and aimed to apply text mining techniques to health care
+complaints that have been registered at the national contact point for health care (Landelijk Meldpunt
+Zorg\footnote{\url{https://www.landelijkmeldpuntzorg.nl}})
+We investigated the usefulness of text mining to categorise and cluster complaints, to automatically determine the
+severity of incoming complaints, to extract patterns  and to identify risk cases. This project turned out to be a good
+test case of the applicability and usefulness of LaMachine as a standalone research environment.
+As the complaint data is highly sensitive, it could not leave the secure servers of the health inspectorate and was
+stored in an environment without internet access. We needed to bring the software to the data via a shared folder.
+
+We used a virtual machine (VM) image of LaMachine and we ran this 64-bits Linux-based VM inside another
+VM with Windows Server 2012, provided to us by the health inspectorate for this project, in which we did have administrative
+rights but no internet access. In terms of hardware we ran on a machine with 8 cores and  32GB internal memory available.
+%using VirtualBox and Vagrnt on the Windows VM
+LaMachine provided a fully functional research environment and we ran all our experiments within LaMachine. We interacted with LaMachine both through the command line, which offers a standard
+shell and enables access to all lower-level tools and programming languages; and through the (offline) webbrowser to use
+the Jupyter Notebook environment.
+
+LaMachine comes with some simple data sharing facilities that allowed us to access the sensitive complaint data via a
+single shared dataspace between host and the VM. Extensive data search and management functions are deliberately beyond the scope of LaMachine, and left to more high-level tooling.
+
+We used many of the available tools in LaMachine within this project: Frog for linguistic annotation of the textual
+content of the complaint and the scikit-learn Python package for classification, T-scan for feature extraction in the form of text characteristics and colibri-core for n-gram analysis.
+
+%het doel van deze studie is om automatische tekstanalysetechnieken (tekstmining) toe te passen op de klachten die bij het Landelijk Meldpunt Zorg binnenkomen.
+Zo hopen we eventuele trends en patronen te vinden die meer inzicht geven in de aard en de relevantie van de klachten.
+De verwachting is om zo potentiële risicogevallen te ontdekken die op basis van de behandeling van individuele klachten nog niet waren gesignaleerd.
+
+\section{Conclusion \& Future work}
+
+The recent release of LaMachine v2, which constituted a full rewrite, has opened up LaMachine to outside
+contribution. Contributor documentation has been written, and at this stage, we greatly welcome external participants
+to join in. Use cases as the example in section \ref{sec-case} contribute to thorough testing and running of LaMachine in less ideal circumstances such as nested VM constructions and offline usage.
+
+Aside from the incorporation of new relevant software, the main objectives for the future are to provide
+greater \emph{interoperability} between the included tools through better \emph{high-level interfaces} for the
+researcher. We see this as a bottom-up process and have now established a firm foundation to build upon. Note that such
+proposed interfaces, including the current portal application in LaMachine, are always considered separate independent
+software projects, which may be deployed by/in/for LaMachine, but also in other contexts. LaMachine remains `just'
+a software distribution at heart.
+
+Development of LaMachine presently takes place in collaboration with the CLARIAH WP3 Virtual Research Environment (VRE)
+project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}}, which has higher ambitions in
+accommodating the researcher and connectivity of data and services, and transcends also those of the CLARIN Language
+Resource Switchboard~\cite{switchboard}. An important part of our future focus will therefore be on interoperability
+with the higher-level tools emerging from the VRE efforts, but also with other parts of the CLARIN infrastructure;
+single-sign on authentication being a notable example here.
+
+\section*{Acknowledgement}
+
+This research was funded by NWO CLARIN-NL, CLARIAH and the ZonMw project {\it Tekstmining in het toezicht: een exploratieve studie naar de automatische verwerking van klachten ingediend bij het Landelijk Meldpunt Zorg}, project number 516004614.
+We thank all project partners: the Dutch Health Inspectorate, IQhealthcare, and Tim Voets for their valuable contributions and help in the ZonMw project.
+
+\bibliography{lamachine}
+
+\end{document}

### 2019-01-14 docs/lamachine_clarin2018.tex
paper update

@@
 needs to know which repositories to use.  Similarly, there are different programming-language-specific ecosystems
-providing their own repositories, such as the Python Package Index for Python, CRAN for R, CPAN for Perl, Maven Central
-for Java.  LaMachine again relies on those to pull and install software from and never forks, archives, or modifies the
-software in any way. In doing so, we compel participating software projects to adhere to well-established distribution
-standards and ensure the software is more sustainable towards the future \cite{softwarequality}. Moreover, we ensure
-that LaMachine never becomes a prerequisite for the software but merely a courtesy or convenience.
+providing their own repositories, such as the Python Package Index\footnote{\url{https://pypi.org}} for Python, CRAN\footnote{\url{https://cran.r-project.org/}} for R, CPAN\footnote{\url{https://www.cpan.org}} for Perl, Maven Central\footnote{\url{https://search.maven.org}} for Java.
+As said, LaMachine relies on those to pull and install software from and never forks, archives, or modifies the software in any way.
+In doing so, we compel participating software projects to adhere to well-established distribution standards and ensure the software is more sustainable towards the future \cite{softwarequality}.
+Moreover, we ensure that LaMachine never becomes a prerequisite for the software but merely a courtesy or convenience.
 
 LaMachine provides ample flexibility that allows it to be deployable in different contexts. First of all there is
-flexibility with regard to the target platform, where we support several major GNU/Linux distributions (Debian, Ubuntu,
+flexibility regarding the installation form, which we call \emph{flavours}:
+
+\begin{itemize}
+    \item Local installation - This installs the bulk of LaMachine in a separate local virtual environment (a seperate
+        directory) that has to be explicitly activated to be used. This also allows for multiple different installations
+        of LaMachine on the same host system (e.g. for different users or with different software configurations).
+        This local installation still actively relies on various global dependencies that are available through the
+        package manager of your distribution.
+    \item Global installation - This flavour is used for a host that is fully dedicated to LaMachine. Everything will be
+        installed globally so there is only one installation possible, multiple users will all make use of the same
+        installation.
+    \item Docker container - Installs LaMachine in a container. Containerisation seperates the entire run-time
+        environment
+\end{itemize}
+
+flexibility with regard to the target platform, where we support several major GNU/Linux distributions:
+
+\begin{enumerate}
+    \item Debian
+    \item Ubuntu
+    \item Arch Linux
+    \item CentOS and RedHat Enterprise Linux
+    \item Fedora
+\end{enumerate}
+
+In addition, we have limited support for macOS and the homebrew package manager\footnote{Alternatives such as macports are not
+supported}.
+
+(Debian, Ubuntu,
 CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as macOS (although with more limitations). Second, there is

### 2019-01-14 docs/lamachine_clarin2018.tex
work on the Lamachine paper

@@
 \documentclass[a4paper,11pt]{article}
-\usepackage{CLARIN2018}
+\usepackage{etendu-cis}
+%\usepackage{CLARIN2018}
 % - - - - - - - IMPORTANT - - - - - - - -
@@
 
+\subsection{Flavours}
+
 LaMachine provides ample flexibility that allows it to be deployable in different contexts. First of all there is
@@
 
-\begin{itemize}
-    \item Local installation - This installs the bulk of LaMachine in a separate local virtual environment (a seperate
+\begin{enumerate}
+    \item \textbf{Local installation} - This installs the bulk of LaMachine in a separate local virtual environment (a seperate
         directory) that has to be explicitly activated to be used. This also allows for multiple different installations
@@
         package manager of your distribution.
-    \item Global installation - This flavour is used for a host that is fully dedicated to LaMachine. Everything will be
+    \item \textbf{Global installation} - This flavour is used for a host that is fully dedicated to LaMachine. Everything will be
         installed globally so there is only one installation possible, multiple users will all make use of the same
         installation.
-    \item Docker container - Installs LaMachine in a container. Containerisation seperates the entire run-time
-        environment
-\end{itemize}
+    \item \textbf{Docker container} - Installs LaMachine in a container. Containerisation separates the entire run-time
+        environment from the host system, only the Linux kernel is shared. It is a lighter option than full
+        virtualisation.
+    \item \textbf{Singularity container} - Another form of containerisation with more focus on security\footnote{reduced
+        privilege escalation}, making it more suitable than Docker for deployment on computing clusters.
+    \item \textbf{Virtual Machine} - Installs LaMachine as a virtual machine, i.e. through full virtualisation. This allows
+        deployment of LaMachine on hosts which would otherwise not support it, such as Windows. Virtualisation for
+        LaMachine is achieved through Vagrant and VirtualBox\footnote{\url{https://vagrant.org}, \url{https://www.virtualbox.org}}.
+    \item \textbf{Remote provisioning} - Installs LaMachine on a remote dedicated server. This option is most suited for
+        hosting centres.
+\end{enumerate}
 
-flexibility with regard to the target platform, where we support several major GNU/Linux distributions:
+The different flavours all offer a different degree of separation from the host OS, where Virtual Machines are
+completely virtualised, containers still share the kernel with the host OS, and the two native installation flavours,
+local \& global, actually compile against the machine's distribution itself and thus offers the least amount of
+overhead.
 
-\begin{enumerate}
+The two native options, local and global installation, support a variety of major GNU/Linux distributions:
+
+\begin{itemize}
     \item Debian
@@
     \item Arch Linux
-    \item CentOS and RedHat Enterprise Linux
+    \item CentOS
     \item Fedora
+\end{itemize}
+
+Distributions that are derivates of these \emph{may} also work, such as RedHat Enterprise Linux (CentOS), Linux Mint (Ubuntu),
+and Manjaro (Arch Linux).
+
+In addition, we have limited support for macOS (the local installation flavour) and the homebrew package
+manager\footnote{Alternatives such as macports are not supported}, the support is limited because not all participating
+software supports macOS.
+
+For the containerisation and virtualisation solutions, the default distribution we supply is Debian\footnote{The latest
+stable release}. It is, however, still possible to build your own container or virtual machine based on any of the other supported
+distributions. In fact, containers, virtual machines and remote provisioning can all be considered special wrapped forms of
+the global installation flavour.
+
+Pre-built docker containers and virtual machine images with a limited selection of participating software are uploaded
+to the Docker Hub\footnote{\url{https://hub.docker.com/r/proycon/lamachine/}} and Vagrant
+Cloud\footnote{\url{https://app.vagrantup.com/proycon/boxes/lamachine}}, respectively, for each LaMachine release.
+
+\subsection{Versions}
+
+LaMachine offers three distinct \emph{versions}, regardless of the flavour:
+
+\begin{enumerate}
+    \item \textbf{stable} - This is the default and recommended for most situations. It installs the latest stable
+        versions of all included software.
+    \item \textbf{development} - This installs the latest development versions of the included software.
+In practise, this usually means that software is pulled directly from the version controlled repository and compiled and installed from source.
+Due to the experimental nature, the development version of
+        LaMachine may at times break and not install successfully.
+    \item \textbf{custom} - This installs custom versions of all included software, i.e. the user explicitly
+        specifies which versions to install for each software package.  Such a version list can for instance be exported
+        from another LaMachine installation, and then allows to rebuild a similar environment from scratch, providing a
+        limited level of reproducibility. We say limited, because packages
+        provided by the underlying distribution are not a part of this scheme.
 \end{enumerate}
 
-In addition, we have limited support for macOS and the homebrew package manager\footnote{Alternatives such as macports are not
-supported}.
+The nomenclature is admittedly a bit confusing, but the notion of version discussed in the section refers to the
+versions of the various software packages inside LaMachine. It should not be
+confused with the actual \emph{version release number} of Lamachine as a whole, which is the version number assigned to
+the collective of installation scripts LaMachine provides, and which marks LaMachine releases.
 
-(Debian, Ubuntu,
-CentOS, RedHat Enterprise Linux, Fedora, Arch Linux), as well as macOS (although with more limitations). Second, there is
-flexibility with regard to the form, where we support \emph{containerisation} through
-Docker\footnote{\url{https://www.docker.com}}, \emph{virtualisation} through Vagrant
-and VirtualBox\footnote{\url{https://vagrant.org}, \url{https://www.virtualbox.org}}, direct remote provisioning through Ansible (for production
-servers), or an installation that is either global to the machine or local in a custom directory for a specific user
-(using \texttt{virtualenv}). Pre-built docker containers and virtual machine images with a limited selection of participating software are regularly
-uploaded to the Docker Hub and Vagrant Cloud, respectively. The different flavours all offer a different degree of
-seperation from the host OS, where Virtual Machines are completely virtualised, Docker Containers still share the kernel
-with the host OS, and the machine-specific installation flavour actually compiles against the machine's distribution itself and thus
-offers the least amount of overhead.
+\subsection{Bootstrapping}
 
-%%iris: ik heb dit stukje geintegreerd in de  case study... (ik heb de eerste paragraaf toch nog even hier opgenomen)
-Installation of LaMachine begins with a single bootstrap command\footnote{See
-\url{https://proycon.github.io/LaMachine}}.  It can interactively query the users for their software preferences
-\emph{(stored as the host configuration)}, e.g. the flavour of LaMachine, as well as the set of software to install,
-\emph{the installation manifest}. This set is never static but can be customized by the user.  The user may also opt for installing the latest releases, the more
-experimental development versions of the software, or specific custom versions (to facilitate scientific
-reproducibility). The bootstrap procedure detects and installs the necessary prerequisites automatically and eventually
-invokes Ansible to perform the bulk of the work.  Figure~\ref{fig:arch} provides a schematic view.
+Installation of LaMachine begins with a single bootstrap command\footnote{See \url{https://proycon.github.io/LaMachine}}
+executed on the command line.  It can interactively query the users for their software preferences \emph{(stored as the
+host configuration)}, e.g. the flavour of LaMachine, as well as the set of software to install, \emph{the installation
+manifest}. This set is never static but can be customized by the user. The bootstrap procedure detects and installs the
+necessary prerequisites automatically and eventually invokes Ansible to perform the bulk of the work, unless a
+pre-published container or VM is selected.  Figure~\ref{fig:arch} provides a schematic view of the LaMachine
+architecture.
 
@@
 
-LaMachine also aims to harmonise the metadata of all installed software, by converting metadata from upstream
+Once LaMachine is installed, in any of its flavours, you can update it from inside by running
+\textttt{lamachine-update}. This updates all of the software managed by LaMachine.
+
+\subsection{Metadata}
+
+LaMachine aims to harmonise the metadata of all installed software, by converting metadata from upstream
 repositories, i.e. the repositories where tool providers deposit their software, to a common standard called CodeMeta
@@
 that directly. This in turn enables other tools to do proper service discovery and provenance logging.
+
+\section{Software}
+
+LaMachine exists since May 2015 and has been used extensively ever since by numerous users, in 2018 version 2 was
+released which was a significant rewrite, powered by Ansible.
+
+In LaMachine, software is grouped into various ``packages'', each package\footnote{For those familiar with
+Ansible, a package in LaMachine is an Ansible role} groups one or multiple programs that have some kind of relation. The
+installation manifest lists all packages that will be installed, at the user's discretion.
+
+LaMachine was initially conceived as the primary means of distribution of the software stack developed at the Language
+Machines Research Group and the Centre of
+Language and Speech Technology, Radboud University Nijmegen. The majority of this sofware was either fully or partially developed under the auspices of CLARIN-NL or successor
+CLARIAH. Some software by other CLARIN-NL/CLARIAH partners is also included.
+
+\begin{itemize}
+ \item by the Language Machines Research Group and the Centre of Language and Speech Technology, Radboud University Nijmegen
+\footnote{Links: \url{https://languagemachines.github.io/timbl}, \url{https://languagemachines.github.io/mbt},
+    \url{https://languagemachines.github.io/ucto}, \url{https://languagemachines.github.io/frog},
+    \url{http://ilk.uvt.nl/wopr}, \url{https://proycon.github.io/folia}, \url{https://proycon.github.io/clam},
+\url{https://github.com/proycon/flat}, \url{https://proycon.anaproy.nl/pynlpl},
+\url{https://proycon.github.io/colibri-core/},
+\url{https://github.com/proycon/gecco},
+\url{https://github.com/proycon/valkuil-gecco},
+\url{https://github.com/LanguageMachines/PICCL}, \url{https://githu.com/LanguageMachines/ticcltools},
+\url{https://github.com/proycon/labirinto},
+\url{https://github.com/proycon/oersetter-webservice}
+}
+ \begin{itemize}
+     \item \textbf{Timbl} - A memory-based machine learning
+         toolkit, and \textbf{Mbt}, a memory-based tagger based on timbl.
+     \item \textbf{Ucto} - A tokeniser
+     \item \textbf{Frog} - Frog is an integration of various memory-based natural language processing (NLP) modules
+         developed for Dutch. It can do Part-of-Speech tagging, lemmatisation, named entity recogniton, shallow parsing,
+         dependency parsing and morphological analysis. Also included in LaMachine;
+         \textbf{Toad}, Trainer Of All Data, training tools for Frog.
+     \item \textbf{Wopr}- Memory-based Word Predictor.
+     \item \textbf{CLAM} - Quickly build RESTful webservices, powers many webservices offered by LaMachine.
+     \item \textbf{FoLiA} - Format for Linguistic Annotation, with tools and libraries in/for Python and $C++$.
+     \item \textbf{FLAT} - FoLiA Linguistic Annotation Tool: a web-based linguistic annotation tool.
+     \item \textbf{PyNLPl} - Python Natural Language Processing Library.
+     \item \textbf{Colibri Core} -- Colibri core is an NLP tool as well as a C++ and Python library for working with
+         basic linguistic constructions such as n-grams and skipgrams (i.e patterns with one or more gaps, either of
+         fixed or dynamic size) in a quick and memory-efficient way. -
+     \item \textbf{Gecco} - Generic Environment for Context-Aware Correction of Orthography - and \textbf{Valkuil.net},
+         an instantiation thereof for Dutch.
+     \item \textbf{PICCL} - PICCL: A set of workflows for corpus building through OCR, post-correction (through \textbf{TICCL}) and Natural Language Processing.
+     \item \textbf{Labirinto}- A web-based portal listing all
+         available tools in LaMachine, an ideal starting point for LaMachine.
+     \item \textbf{Oersetter} - A Frisian-Dutch Machine Translation system in collaboration with the Fryske Akademy.
+ \end{itemize}
+ \item by the University of Groningen
+ \begin{itemize}
+     \item \textbf{Alpino}\footnote{\url{http://www.let.rug.nl/vannoord/alp/Alpino/}} - a dependency parser and tagger for Dutch
+ \end{itemize}
+ \item by the Vrije Universiteit Amsterdam
+ \begin{itemize}
+     \item \textbf{KafNafParserPy}\footnote{\url{https://github.com/cltl/KafNafParserPy}} - A python module to parse NAF files
+ \end{itemize}
+ \item by Utrecht University
+ \begin{itemize}
+     \item  \textbf{T-scan}\footnote{\url{https://github.com/proycon/tscan}} - T-scan is a Dutch text analytics tool for readability prediction (initially developed at TiCC, Tilburg University, and in collaboration with Radboud University, Nijmegen)
+ \end{itemize}
+ \item by Meertens Instituut
+ \begin{itemize}
+     \item \textbf{Python Course for the Humanities}\footnote{http://www.karsdorp.io/python-course/} - Interactive
+         tutorial and introduction into programming with Python for the humanities
+ \end{itemize}
+\end{itemize}
+
+In addition to the above listed specific software, LaMachine also incorporates a large number of renowned tools by external international parties, offering
+most notably a mature Python environment with renowned scientific modules. The following list is not
+exhaustive: \footnote{\url{https://jupyterlab.readthedocs.io/en/stable/}, \url{https://pytorch.org}, \url{http://www.nltk.org},
+    \url{https://spacy.io},  \url{http://www.nextflow.io}, \url{https://stanfordnlp.github.io/CoreNLP/},  \url{https://github.com/tesseract-ocr/tesseract}, \url{https://tensorflow.org},
+\url{http://kaldi-asr.org}, \url{http://www.statmt.org/moses}}
+
+\begin{itemize}
+    \item \textbf{Python}: Numpy, Scipy, Matplotlib, Scikit-Learn, IPython, Jupyter, ...
+    \begin{itemize}
+        \item \textbf{Jupyter Lab} The successor of the popular Jupyter Notebooks, offers notebooks, a web-based IDE, terminals. An ideal entry point to get started with LaMachine and all it contains!
+        \item \textbf{PyTorch} - Deep-learning library for Python
+        \item \textbf{NLTK} - Natural Language Toolkit for Python
+        \item \textbf{Spacy} - Industrial-Strength NLP in Python
+    \end{itemize}
+    \item R
+    \item Java
+    \begin{itemize}
+        \item NextFlow - A system and language for writing parallel and scalable pipelines in a portable manner.
+        \item Stanford CoreNLP](https://stanfordnlp.github.io/CoreNLP/) - Various types of linguistic enrichment
+    \end{itemize}
+    \item \textbf{Tesseract} - Open Source Optical Character Recognition (OCR)
+    \item \textbf{Tensorflow} - Open-source machine learning framework
+    \item \textbf{Kaldi} - Speech Recognition Framework (ASR)
+    \item \textbf{Moses} - Statistical Machine Translation system
+\end{itemize}
+
+%However, this
+%project is not limited to one research group and is open to participation by other software providers, especially those
+%also in CLARIAH and the upcoming CLARIAH PLUS project. We already include some relevant software by other CLARIAH
+%partners.
+
+\section{Web Interface}
 
@@
 text, code and data output and have gained great popularity in data science community nowadays.
-
-\section{Software}
-
-LaMachine exists since May 2015 and has been used extensively ever since by numerous users, in 2018 version 2 was
-released which was a significant rewrite. LaMachine was initially conceived as the primary means of distribution of the
-software stack developed at CLST, Radboud University Nijmegen. It therefore includes a lot of our software.  A full list
-of included software goes beyond the scope of this overview; we will merely mention some CLARIN-NL/CLARIAH-funded tools:
-ucto (a tokeniser), Frog (an NLP suite for Dutch), FoLiA (Format for Linguistic Annotation, with assorted tools), FLAT
-(a web-based linguistic Annotation tool), PICCL (an OCR and post-OCR correction pipeline) and CLAM. However, this
-project is not limited to one research group and is open to participation by other software providers, especially those
-also in CLARIAH and the upcoming CLARIAH PLUS project. We already include some relevant software by other CLARIAH
-partners. Moreover, LaMachine incorporates a large number of renowned tools by external international parties, offering
-most notably a mature Python environment with renowned scientific modules such as scipy, numpy, scikit-learn,
-matplotlib, nltk, spacy, pytorch, keras, gensim, tensorflow, and many others, but also R, Java and tools such as
-Stanford CoreNLP and Kaldi.
 

### 2019-01-16 docs/lamachine_clarin2018.tex
work on paper

@@
 \documentclass[a4paper,11pt]{article}
-\usepackage{etendu-cis}
-%\usepackage{CLARIN2018}
+\usepackage{CLARIN2018}
 % - - - - - - - IMPORTANT - - - - - - - -
@@
 Once LaMachine is installed, in any of its flavours, you can update it from inside by running
-\textttt{lamachine-update}. This updates all of the software managed by LaMachine.
+\texttt{lamachine-update}. This updates all of the software managed by LaMachine.
 
@@
 \begin{itemize}
- \item by the Language Machines Research Group and the Centre of Language and Speech Technology, Radboud University Nijmegen
-\footnote{Links: \url{https://languagemachines.github.io/timbl}, \url{https://languagemachines.github.io/mbt},
+ \item by the Language Machines Research Group and the Centre of Language and Speech Technology, Radboud University
+     Nijmegen\footnote{Links: \url{https://languagemachines.github.io/timbl}, \url{https://languagemachines.github.io/mbt},
     \url{https://languagemachines.github.io/ucto}, \url{https://languagemachines.github.io/frog},
@@
  \begin{itemize}
-     \item \textbf{Python Course for the Humanities}\footnote{http://www.karsdorp.io/python-course/} - Interactive
+     \item \textbf{Python Course for the Humanities}\footnote{\url{http://www.karsdorp.io/python-course/}} - Interactive
          tutorial and introduction into programming with Python for the humanities
@@
 
-Leveraging this metadata, LaMachine comes with a webserver that offers a portal website with
-access and overview of all installed tools, including web services and web applications. It also comes with a
+LaMachine comes with a webserver that offers a portal website which provides an overview of all installed software in
+LaMachine, and acts as a point of access for web services andweb applications. This portal, as shown in
+Figure~\ref{fig:portal}, leverages the metadata
+registry compiled in each LaMachine installation. In situations where web interfaces are not needed or desired, the
+user has the ability to opt-out.
+
+\begin{figure}[htb] \begin{center}
+\includegraphics[width=135.0mm]{screenshot_portal.jpg}
+\end{center}
+\caption{\footnotesize{A screenshot of the LaMachine portal, powered by Labirinto:
+\url{https://github.com/proycon/labirinto}}}
+\label{fig:portal}
+\end{figure}
+
+LaMachine also comes with  a
 Jupyter Lab\footnote{\url{https://jupyter.org/}} environment which provides a web-based Integrated Development
 Environment (IDE) for scripting in Python and R, web-based terminal access, and so-called \emph{notebooks} which mix
-text, code and data output and have gained great popularity in data science community nowadays.
+text, code and data output and have gained great popularity in data science community nowadays. This is shown in
+Figure~\ref{fig:lab}.
+
+\begin{figure}[htb] \begin{center}
+\includegraphics[width=135.0mm]{screenshot_lab.jpg}
+\end{center}
+\caption{\footnotesize{A screenshot of the Jupyter Lab Environment}}
+\label{fig:lab}
+\end{figure}
 

### 2019-01-16 docs/lamachine_clarin2018.tex
paper update

@@
 
-In addition, we have limited support for macOS (the local installation flavour) and the homebrew package
-manager\footnote{Alternatives such as macports are not supported}, the support is limited because not all participating
-software supports macOS.
+In addition, we have limited support for macOS (the local installation flavour), powered in part by the homebrew package
+manager\footnote{Alternatives such as macports are not supported}. Support for macOS is limited because not all participating
+software supports it.
 
@@
 Environment (IDE) for scripting in Python and R, web-based terminal access, and so-called \emph{notebooks} which mix
-text, code and data output and have gained great popularity in data science community nowadays. This is shown in
+text, code and data output and have gained great popularity in data science community. This is shown in
 Figure~\ref{fig:lab}.

### 2019-01-17 docs/lamachine_clarin2018.tex
advancing the paper

@@
 \subsection{Metadata}
+\label{sec:metadata}
 
-LaMachine aims to harmonise the metadata of all installed software, by converting metadata from upstream
-repositories, i.e. the repositories where tool providers deposit their software, to a common standard called CodeMeta
+LaMachine aims to harmonise the metadata of all installed software, by converting metadata from upstream repositories,
+i.e. the repositories where tool providers deposit their software, to a common yet simple standard called CodeMeta
 \footnote{\url{https://codemeta.github.io/}, described in JSON-LD} \cite{codemeta,codemetar} where possible, or
-encouraging software developers to provide their codemeta metadata inside their source code repositories and using
-that directly. This in turn enables other tools to do proper service discovery and provenance logging.
+encouraging software developers to provide their codemeta metadata inside their source code repositories and using that
+directly. This in turn enables other tools to do proper service discovery and provenance logging.
 
@@
 
-LaMachine exists since May 2015 and has been used extensively ever since by numerous users, in 2018 version 2 was
+LaMachine exists since May 2015 and has been used extensively ever since by numerous users, in early 2018 version 2 was
 released which was a significant rewrite, powered by Ansible.
@@
 Ansible, a package in LaMachine is an Ansible role} groups one or multiple programs that have some kind of relation. The
-installation manifest lists all packages that will be installed, at the user's discretion.
+installation manifest lists all packages that will be installed, at the user's discretion. After the initial
+installation, the user can always add more packages using \texttt{lamachine-add}.
 
@@
 Language and Speech Technology, Radboud University Nijmegen. The majority of this sofware was either fully or partially developed under the auspices of CLARIN-NL or successor
-CLARIAH. Some software by other CLARIN-NL/CLARIAH partners is also included.
+CLARIAH. Some software by other CLARIN-NL/CLARIAH partners is also included. This project is not limited to one research
+group and is explicitly open to participation by other software providers, especially those
+also in CLARIAH.
 
@@
 
-%However, this
-%project is not limited to one research group and is open to participation by other software providers, especially those
-%also in CLARIAH and the upcoming CLARIAH PLUS project. We already include some relevant software by other CLARIAH
-%partners.
+\section{Interfaces and Audiences}
 
-\section{Web Interface}
+We already addressed the need for different interfaces for different audiences in Section~\ref{intro}. The challenge we
+face, and for which LaMachine offers a solution, is one of software maintainability and distribution. How do we maintain
+and distribute software that is often highly complex and consists of multiple interconnected components, considering
+that this software is used differently by different audiences? A key aspect here is the reusability and accessibility of individual
+software components. The philosophy we subscribe to encourages the development and distribution of software in a layered or modular fashion,
+allowing each building block to serve as the foundation for another more high-level interface, without sacrificing the usability of the
+foundation as such. This is in line with the UNIX philosophy of developing tools that do \emph{"one thing only, and do
+it well"} and is contrasted with monolithic software solutions that are limiting because they either provide only high-level interfaces but do not
+expose their inner components to build upon, or they are as a swiss-army knife full of needless but inseparable
+components.
 
-LaMachine comes with a webserver that offers a portal website which provides an overview of all installed software in
-LaMachine, and acts as a point of access for web services andweb applications. This portal, as shown in
-Figure~\ref{fig:portal}, leverages the metadata
-registry compiled in each LaMachine installation. In situations where web interfaces are not needed or desired, the
-user has the ability to opt-out.
+\subsection{Low-level Interfaces}
+
+In any of its flavours, LaMachine offers low-level shell access, i.e. a command-line interface accessed through a
+terminal. In flavours that are separated from the host system over the network, this is accomplished over \texttt{ssh}.
+At that point, as shown in Figure~\ref{fig:venv}, the user has the liberty to do whatever she wants and can invoke any
+of the tools that offer a text interface, including text editors such as vim, emacs or nano, version control systems
+such as git and interpreters such as Python or R (or compilers for that matter). The latter in turn allow users to use
+one of the many specialised programming libraries included in LaMachine to build their own tools. All this makes
+LaMachine ideally suited as a development environment.
+
+\begin{figure}[htb] \begin{center}
+\includegraphics[width=135.0mm]{screenshot_venv_activate.jpg}
+\end{center}
+\caption{\footnotesize{Activation of a local LaMachine environment on the command-line}}
+\label{fig:venv}
+\end{figure}
+
+\subsection{High-level Interfaces: Web-based access}
+
+LaMachine comes with a webserver\footnote{In situations where web interfaces are not needed or desired, the
+user has the ability to opt-out of this}. This webserver serves various web-capable tools that are incorporated in LaMachine. One
+of these tools is a portal website that provides an overview of all installed software in
+LaMachine, and acts as a point of access for web services and web applications. This portal, as shown in
+Figure~\ref{fig:portal}, leverages the metadata registry compiled in each LaMachine installation (see
+section~\ref{sec:metadata}).
 
@@
 LaMachine also comes with a
-Jupyter Lab\footnote{\url{https://jupyter.org/}} environment which provides a web-based Integrated Development
+Jupyter Lab\footnote{\url{https://jupyter.org/}} installation which provides a web-based Integrated Development
 Environment (IDE) for scripting in Python and R, web-based terminal access, and so-called \emph{notebooks} which mix
 text, code and data output and have gained great popularity in data science community. This is shown in
-Figure~\ref{fig:lab}.
+Figure~\ref{fig:lab}. This type of virtual laboratory provides a powerful interface that can form an alternative to
+regular shell access and may have a larger appeal for those parts of the audience that do not feel completely
+comfortable in the terminal.
 
@@
 \end{figure}
+
+LaMachine provides various CLAM-based webservices. CLAM is a long-time CLARIN project that allows developers to turn
+their command-line NLP tool into a RESTful webservice with relatively little effort \cite{CLAMPAPER}, and which in
+addition also provides a generic web-interface for human end-users. The portal website in LaMachine links to these. This
+again address an audience that does not require a high degree of technical expertise, but is suitable for a wider
+audience of researchers, including those in the Humanities.
+
+\section{Comparisons and alternatives}
+\label{sec:comparisons}
+
+How does LaMachine compare to alternatives? To answer this question we must first look at what may be considered
+alternatives to LaMachine. At this point, it is important to emphasise once more that LaMachine is offered as convenience.
+LaMachine always seeks to use installation channels that the user can also access directly. For instance, is the user
+only interested in a single Python package that LaMachine offers? Then she can just as well install it directly using
+\texttt{pip install}. Is the user only interested in a single package that underlying Linux distribution (or macOS' homebrew)
+already provides? Then, by all means, she should go for that.
+
+LaMachine starts to fill a need when the user is interested in complex software, such as complex NLP pipelines with
+multiple components, the individual installation (sometimes involving compilation from source) and configuration of
+which would be not be trivial. Moreover, LaMachine shines when the user wants higher-level interfaces on top of that,
+something she can use right out of the box. It also fulfills a role in case the user simply does not know what she wants
+or needs yet, and needs an environment where she can explore a variety of pre-installed tools and experiment.
+
+Comparisons could be drawn between LaMachine as a kind of Linux distribution and other Linux distributions. But that
+would not be appropriate as we characterise LaMachine more as a meta-distribution which simply builds on the foundation
+of several existing major Linux distributions. As a meta-distribution that runs in a variety of flavours and on a verity
+of platforms, it has a great deal of flexibility that normal distributions don't have. At the same time, the scope of
+LaMachine is more narrowly defined than that of a generic Linux distribution, including even specialised Linux
+distributions that focus on scientific computing, such as Fedora
+Scientific\footnote{\url{https://labs.fedoraproject.org/en/scientific/}}.
+
+The best comparison can be drawn with Anaconda\footnote{\url{https://www.anaconda.com/}}, which we could also qualify as
+a meta-distribution under our definition, and which enjoys great popularity in the data science community. Anaconda has
+a focus on Python and R. It is much wider in scope than LaMachine, which has a strong NLP and CLARIN/CLARIAH focus. When
+it comes to Python, there is considerable overlap in the modules that LaMachine and Anaconda offer. Unlike LaMachine,
+Anaconda does not focus on providing different flavours such as a VM or a Docker Container\footnote{This is not a
+technical limitation but simply a matter of different objectives, one could easily create a VM or Docker Container and
+install Anaconda in it.}.
+
+At the start of the development cycle for LaMachine v2, we investigated whether Anaconda and its ecosystem would provide
+a sufficient foundation for us to build upon. We decided this was not the case for a number of reasons: Anaconda
+introduces more overhead than we desired for our purposes and conflicted with certain technology choices we made. We
+wanted our native flavours (local environment and global environment) to be as close to the underlying distribution as possible and to reuse existing
+technologies and packages from the distribution (or from homebrew for macOS), such as compilers (e.g. gcc) and
+interpreters (e.g. python).  Anaconda, on the other hand, choses to provide its own packages and builds on those. It
+does so using its own package manager and package repositories. This would require repackaging certain software for the
+anaconda ecosystem, especially handling of non-Python software was not trivial.  Whereas LaMachine consists of a series
+of Ansible scripts for the installation and configuration of software, Anaconda has its own package manager (conda)
+that draw from its own package repositories (conda-forge).
+
+For the distribution and deployment of complex software setups such as NLP pipelines, containers are a common solution
+nowadays. The Docker flavour of LaMachine provides something similar, but rather than providing a single static Docker recipe
+that builds a single kind of container, LaMachine offers a high degree of flexibility for the construction of different
+containers, i.e. containing different software of having different configurations, based on the user's needs. It is
+quite feasible to use instantiate a variety of LaMachine containers and use a container orchestration system such
+Kubernetes or Docker Swarm for automated deployment, scaling and management thereof. Such orchestration, however, is
+beyond the scope of LaMachine itself, which aims to provide a singular environment, i.e. everything installed in a
+LaMachine instance shares the same userspace and duplication is strongly minimised.
 

### 2019-01-17 docs/lamachine_clarin2018.tex
paper update

@@
 
-\section{Introduction} \label{intro}
+\section{Introduction}
+\label{sec:intro}
 
@@
 
-We already addressed the need for different interfaces for different audiences in Section~\ref{intro}. The challenge we
+We already addressed the need for different interfaces for different audiences in Section~\ref{sec:intro}. The challenge we
 face, and for which LaMachine offers a solution, is one of software maintainability and distribution. How do we maintain
@@
 At the start of the development cycle for LaMachine v2, we investigated whether Anaconda and its ecosystem would provide
-a sufficient foundation for us to build upon. We decided this was not the case for a number of reasons: Anaconda
+a sufficient foundation for us to build upon. We concluded this was not the case for a number of reasons: Anaconda
 introduces more overhead than we desired for our purposes and conflicted with certain technology choices we made. We
@@
 interpreters (e.g. python).  Anaconda, on the other hand, choses to provide its own packages and builds on those. It
-does so using its own package manager and package repositories. This would require repackaging certain software for the
-anaconda ecosystem, especially handling of non-Python software was not trivial.  Whereas LaMachine consists of a series
-of Ansible scripts for the installation and configuration of software, Anaconda has its own package manager (conda)
-that draw from its own package repositories (conda-forge).
+does so using its own package manager (conda) and package repositories (conda-forge). This would require repackaging certain software for the
+anaconda ecosystem, especially handling of non-Python software was not trivial, as Anaconda was originally conveived as a
+Python distribution.
 
@@
 
-\section{Case study}\label{sec-case}
+\section{Case studies}
+\label{sec:case}
+
+\subsection{Text Mining for Health Inspection}
 
@@
 
+\subsection{Workshop: Cataloguing of Textual Cultural Heritage Objects}
+
+%TODO
+
 \section{Conclusion \& Future work}
 
-The recent release of LaMachine v2, which constituted a full rewrite, has opened up LaMachine to outside
-contribution. Contributor documentation has been written, and at this stage, we greatly welcome external participants
-to join in. Use cases as the example in section \ref{sec-case} contribute to thorough testing and running of LaMachine in less ideal circumstances such as nested VM constructions and offline usage.
+LaMachine provides a flexible solution for the installation of a variety of NLP software, resulting in a kind
+of virtual laboratory that, through various interfaces, can be employed by a variety of people, from developers and data
+scientists to the wider research community that CLARIN explicitly addresses.
+
+Use cases as the examples in section \ref{sec:case} contribute to thorough testing and running of LaMachine in less ideal
+circumstances such as nested VM constructions and offline usage.
 
@@
 
+LaMachine v2 is open to  outside contribution. Contributor documentation has been written, and at this stage, we greatly
+welcome external participants to join in.
+
 \section*{Acknowledgement}

### 2019-01-17 docs/lamachine_clarin2018.tex
paper update

@@
 \usepackage{graphicx}
+\usepackage{placeins}
 %\usepackage{latexsym}
@@
 
-The two native options, local and global installation, support a variety of major GNU/Linux distributions:
-
-\begin{itemize}
-    \item Debian
-    \item Ubuntu
-    \item Arch Linux
-    \item CentOS
-    \item Fedora
-\end{itemize}
-
-Distributions that are derivates of these \emph{may} also work, such as RedHat Enterprise Linux (CentOS), Linux Mint (Ubuntu),
+The two native options, local and global installation, support a variety of major GNU/Linux distributions: Debian,
+Ubuntu, Arch Linux, CentOS, Fedora and, to a more limited degree, we also support macOS (the local installation flavour), powered in part by the homebrew package
+manager\footnote{Alternatives such as macports are not supported}. Support for macOS is limited because not all participating
+software supports it. Linux Distributions that are derivates of these \emph{may} also work, such as RedHat Enterprise Linux (CentOS), Linux Mint (Ubuntu),
 and Manjaro (Arch Linux).
-
-In addition, we have limited support for macOS (the local installation flavour), powered in part by the homebrew package
-manager\footnote{Alternatives such as macports are not supported}. Support for macOS is limited because not all participating
-software supports it.
 
@@
 host configuration)}, e.g. the flavour of LaMachine, as well as the set of software to install, \emph{the installation
-manifest}. This set is never static but can be customized by the user. The bootstrap procedure detects and installs the
+manifest}. This set is never static but can be customized by the user. The bootstrap procedure, shown in
+Figure~\ref{fig:bootstrap}, detects and installs the
 necessary prerequisites automatically and eventually invokes Ansible to perform the bulk of the work, unless a
@@
 
-%After a successful build, the user may interact with LaMachine either through the command line, which offers a standard
-%shell and enables access to all lower-level tools and programming languages; or through his or her webbrowser, which
-%when enabled offers a simple portal page towards all installed web-capable tools. Amongst these services is also a
-%Jupyter Notebook environment\footnote{\url{https://jupyter.org/}}, which offers a more graphical scripting environment
-%for e.g. Python and R that has gained great popularity in the scientific community.
-%
-%The current version of LaMachine comes with some simple data sharing facilities only: we provide a
-%single shared dataspace between host and VM/container (where applicable), or allow simple upload. Extensive data
-%search and management functions are deliberately beyond the scope of LaMachine, and left to more high-level tooling.
+\begin{figure}[htb] \begin{center}
+\includegraphics[width=110.0mm]{screenshot_bootstrap.jpg}
+\end{center}
+\caption{\footnotesize{A screenshot of the bootstrap procedure}}
+\label{fig:bootstrap}
+\end{figure}
 
@@
 directly. This in turn enables other tools to do proper service discovery and provenance logging.
+
+\FloatBarrier
 

### 2019-01-17 docs/lamachine_clarin2018.tex
paper update

@@
 settings.}, there is a natural bias towards the former as lower-level interfaces are often a prerequisite to build
-higher-level interfaces on. Depending on the \emph{flavour} of LaMachine chosen, it makes a good virtual research
+higher-level interfaces on. Depending on the \emph{flavour} (more on this later) of LaMachine chosen, it makes a good virtual research
 environment for a data scientist, whether on a personal computer or on a computing cluster, a good development
@@
     \item The source
-        code must be hosted in a public version controlled repository\footnote{e.g. Github, Gitlab, Bitbucket, provided
+        code must be hosted in a public version controlled repository. \footnote{e.g. Github, Gitlab, Bitbucket, provided
         the respository is public}
-    \item There must be build process to compile the source, if applicable, and install the program
-    \item There must be some kind of release protocol (with semantic versioning) using the proper
+    \item There must be a build process to compile the source, if applicable, and install the program or library.
+    \item There must be some kind of release protocol (adhering to semantic versioning) that publishes software using the proper
 technology-specific channels.
@@
         environment from the host system, only the Linux kernel is shared. It is a lighter option than full
-        virtualisation.
+        virtualisation. \footnote{Note that Docker on other platforms such as Windows and macOS does do use full
+        virtualisation.}
     \item \textbf{Singularity container} - Another form of containerisation with more focus on security\footnote{reduced
-        privilege escalation}, making it more suitable than Docker for deployment on computing clusters.
+        privilege escalation}, making it more suitable than Docker for deployment on computing clusters. This flavour is
+        still under development.
     \item \textbf{Virtual Machine} - Installs LaMachine as a virtual machine, i.e. through full virtualisation. This allows
@@
 Ubuntu, Arch Linux, CentOS, Fedora and, to a more limited degree, we also support macOS (the local installation flavour), powered in part by the homebrew package
-manager\footnote{Alternatives such as macports are not supported}. Support for macOS is limited because not all participating
-software supports it. Linux Distributions that are derivates of these \emph{may} also work, such as RedHat Enterprise Linux (CentOS), Linux Mint (Ubuntu),
-and Manjaro (Arch Linux).
+manager\footnote{\url{https://brew.sh}; alternatives such as macports are not supported}. Support for macOS is limited because not all participating
+software supports it. Certain Linux Distributions that are derivatives of these \emph{may} also work, such as RedHat
+Enterprise Linux (CentOS) and Linux Mint (Ubuntu).
 
@@
 
-The nomenclature is admittedly a bit confusing, but the notion of version discussed in the section refers to the
+The nomenclature is admittedly a bit confusing, but the notion of version discussed in this section refers to the
 versions of the various software packages inside LaMachine. It should not be
@@
 Installation of LaMachine begins with a single bootstrap command\footnote{See \url{https://proycon.github.io/LaMachine}}
-executed on the command line.  It can interactively query the users for their software preferences \emph{(stored as the
+executed on the command line.  It can interactively query the user for her software preferences \emph{(stored as the
 host configuration)}, e.g. the flavour of LaMachine, as well as the set of software to install, \emph{the installation
-manifest}. This set is never static but can be customized by the user. The bootstrap procedure, shown in
+manifest}. This set is never static but can be customised by the user. The bootstrap procedure, a screenshot of which is shown in
 Figure~\ref{fig:bootstrap}, detects and installs the
@@
 
-Once LaMachine is installed, in any of its flavours, you can update it from inside by running
+Once LaMachine is installed, in any of its flavours, it can be updated from inside by running
 \texttt{lamachine-update}. This updates all of the software managed by LaMachine.
@@
 
-LaMachine aims to harmonise the metadata of all installed software, by converting metadata from upstream repositories,
+LaMachine aims to harmonise the metadata of all installed software. This is accomplished by converting metadata from upstream repositories,
 i.e. the repositories where tool providers deposit their software, to a common yet simple standard called CodeMeta
@@
 
+We list a selection of the most important software included in LaMachine, grouped by research institute:
+
 \begin{itemize}
@@
 \url{https://github.com/proycon/valkuil-gecco},
-\url{https://github.com/LanguageMachines/PICCL}, \url{https://githu.com/LanguageMachines/ticcltools},
+\url{https://github.com/LanguageMachines/PICCL}, \url{https://github.com/LanguageMachines/ticcltools},
 \url{https://github.com/proycon/labirinto},
@@
          \textbf{Toad}, Trainer Of All Data, training tools for Frog.
-     \item \textbf{Wopr}- Memory-based Word Predictor.
+     \item \textbf{Wopr} - Memory-based Word Predictor.
      \item \textbf{CLAM} - Quickly build RESTful webservices, powers many webservices offered by LaMachine.
@@
 most notably a mature Python environment with renowned scientific modules. The following list is not
-exhaustive: \footnote{\url{https://jupyterlab.readthedocs.io/en/stable/}, \url{https://pytorch.org}, \url{http://www.nltk.org},
+exhaustive: \footnote{Links: \url{https://jupyterlab.readthedocs.io/en/stable/}, \url{https://pytorch.org}, \url{http://www.nltk.org},
     \url{https://spacy.io},  \url{http://www.nextflow.io}, \url{https://stanfordnlp.github.io/CoreNLP/},  \url{https://github.com/tesseract-ocr/tesseract}, \url{https://tensorflow.org},
@@
     \end{itemize}
-    \item R
-    \item Java
+    \item \textbf{R}
+    \item \textbf{Java}
     \begin{itemize}
-        \item NextFlow - A system and language for writing parallel and scalable pipelines in a portable manner.
-        \item Stanford CoreNLP](https://stanfordnlp.github.io/CoreNLP/) - Various types of linguistic enrichment
+        \item \textbf{NextFlow} - A system and language for writing parallel and scalable pipelines in a portable manner.
+        \item \textbf{Stanford CoreNLP} - Various types of linguistic enrichment
     \end{itemize}
@@
 In any of its flavours, LaMachine offers low-level shell access, i.e. a command-line interface accessed through a
-terminal. In flavours that are separated from the host system over the network, this is accomplished over \texttt{ssh}.
-At that point, as shown in Figure~\ref{fig:venv}, the user has the liberty to do whatever she wants and can invoke any
+terminal. In flavours that are separated from the host system by a network, this is accomplished over \texttt{ssh}.
+After accessing the LaMachine environment through a terminal, as shown in Figure~\ref{fig:venv}, the user has the liberty to do whatever she wants and can invoke any
 of the tools that offer a text interface, including text editors such as vim, emacs or nano, version control systems
@@
 LaMachine comes with a webserver\footnote{In situations where web interfaces are not needed or desired, the
-user has the ability to opt-out of this}. This webserver serves various web-capable tools that are incorporated in LaMachine. One
-of these tools is a portal website that provides an overview of all installed software in
-LaMachine, and acts as a point of access for web services and web applications. This portal, as shown in
+user has the ability to opt-out of this.}. This webserver serves various web-capable tools that are incorporated in LaMachine. One
+of these tools is a portal website that provides an overview of all installed software, and acts as a point of access to
+all its web services and web applications. This portal, as shown in
 Figure~\ref{fig:portal}, leverages the metadata registry compiled in each LaMachine installation (see
@@
 Environment (IDE) for scripting in Python and R, web-based terminal access, and so-called \emph{notebooks} which mix
-text, code and data output and have gained great popularity in data science community. This is shown in
-Figure~\ref{fig:lab}. This type of virtual laboratory provides a powerful interface that can form an alternative to
+text, code and data output and have gained great popularity in the data science community. This is shown in
+Figure~\ref{fig:lab}. This type of virtual laboratory provides a powerful interface that provides an alternative to
 regular shell access and may have a larger appeal for those parts of the audience that do not feel completely
@@
 addition also provides a generic web-interface for human end-users. The portal website in LaMachine links to these. This
-again address an audience that does not require a high degree of technical expertise, but is suitable for a wider
+again addresses an audience that does not require a high degree of technical expertise, but is suitable for a wider
 audience of researchers, including those in the Humanities.
@@
 would not be appropriate as we characterise LaMachine more as a meta-distribution which simply builds on the foundation
-of several existing major Linux distributions. As a meta-distribution that runs in a variety of flavours and on a verity
-of platforms, it has a great deal of flexibility that normal distributions don't have. At the same time, the scope of
+of several existing major Linux distributions. As a meta-distribution that runs in a variety of flavours and on a
+variety of platforms, it offers a great deal of flexibility that normal distributions do not have. At the same time, the scope of
 LaMachine is more narrowly defined than that of a generic Linux distribution, including even specialised Linux
@@
 wanted our native flavours (local environment and global environment) to be as close to the underlying distribution as possible and to reuse existing
-technologies and packages from the distribution (or from homebrew for macOS), such as compilers (e.g. gcc) and
-interpreters (e.g. python).  Anaconda, on the other hand, choses to provide its own packages and builds on those. It
+technologies and packages, such as compilers (e.g. gcc) and
+interpreters (e.g. python), from the distribution (or from homebrew for macOS).  Anaconda, on the other hand, choses to provide its own packages and builds on those. It
 does so using its own package manager (conda) and package repositories (conda-forge). This would require repackaging certain software for the
-anaconda ecosystem, especially handling of non-Python software was not trivial, as Anaconda was originally conveived as a
-Python distribution.
+anaconda ecosystem.
 

### 2019-01-25 docs/lamachine_clarin2018.tex
paper updates

@@
 \usepackage{placeins}
+\usepackage{enumitem}
 %\usepackage{latexsym}
@@
 %\setlength\titlebox{5cm}
+\setlist{noitemsep}
 
@@
 Software is a key deliverable and a vital component for research in projects such as those under the CLARIN umbrella. It
-is software that provides researchers the instruments to yield for their research; It is CLARIN's core mission to make
+is software that provides researchers the instruments to yield for their research. It is CLARIN's core mission to make
 digital language resources, including software, available to the wider research community.
@@
 scientist will feel at home at the command line and in scripting environments and uses these fairly low-level
-interfaces, whereas researchers in the humanities demand higher-level interfaces. There is often a power trade-off
+interfaces, whereas researchers in the humanities demand higher-level graphical interfaces. There is often a power trade-off
 between lower and higher-level interfaces, with the former providing maximum flexibility at the cost of a steeper
@@
 
-LaMachine incorporates software providing different types of interfaces\footnote{Command line interfaces, programming
-interfaces, web-user interfaces, webservices.} which, as seen above, typically address different audiences. Whilst we
+LaMachine incorporates software providing different types of interfaces which, as seen above, typically address different audiences. Whilst we
 attempt to accommodate both technical\footnote{Data scientists, DevOps, system administrators, developers.} and
@@
     \item The software included in LaMachine must under a recognised (i.e. OSI-approved) open-source license.
+        Proprietary (closed-source) software is explicitly excluded.
     \item The source
@@
     \item There must be some kind of release protocol (adhering to semantic versioning) that publishes software using the proper
-technology-specific channels.
+technology-specific channels. We will elaborate on this later.
     \item All software that is incorporated in LaMachine must bear at least some relevance to the field of Natural Language Processing.
@@
         Cygwin\footnote{A unix environment on Windows} is not tested or
-        supported.  However, virtualisation technology enables deployment on a wider range of platforms, including Windows.
+        supported either.  However, virtualisation technology enables deployment on a wider range of platforms, including Windows.
 \end{enumerate}
 
-LaMachine is a \emph{meta distribution} as it can be installed in various contexts. At its core, LaMachine consists of a
+We classify LaMachine as a \emph{meta distribution} as it can be installed in various contexts. The heart of LaMachine consists of a
 set of machine-parsable instructions on how to obtain, build (e.g. compile from source), install and configure software.
-These are implemented using Ansible\footnote{\url{https://www.ansible.com}}.  This is notably different from the more
+These are implemented using Ansible\footnote{\url{https://www.ansible.com}; a popular open-source software provisioning,
+configuration and deployment tool by Red Hat}.  This is notably different from the more
 classical notion of Linux distributions, which generally provide their own repositories with (often binary) software
-packages. LaMachine builds on this already established infrastructure by taking these repositories as a given and only
-needs to know which repositories to use.  Similarly, there are different programming-language-specific ecosystems
-providing their own repositories, such as the Python Package Index\footnote{\url{https://pypi.org}} for Python, CRAN\footnote{\url{https://cran.r-project.org/}} for R, CPAN\footnote{\url{https://www.cpan.org}} for Perl, Maven Central\footnote{\url{https://search.maven.org}} for Java.
-As said, LaMachine relies on those to pull and install software from and never forks, archives, or modifies the software in any way.
-In doing so, we compel participating software projects to adhere to well-established distribution standards and ensure the software is more sustainable towards the future \cite{softwarequality}.
-Moreover, we ensure that LaMachine never becomes a prerequisite for the software but merely a courtesy or convenience.
+packages. LaMachine builds on this already established infrastructure by taking these repositories as a foundation where
+possible. Similarly, as implied in point five above, there are different programming-language-specific ecosystems
+providing their own repositories, such as the Python Package Index\footnote{\url{https://pypi.org}} for Python,
+CRAN\footnote{\url{https://cran.r-project.org/}} for R, CPAN\footnote{\url{https://www.cpan.org}} for Perl, Maven
+Central\footnote{\url{https://search.maven.org}} for Java.  LaMachine relies on those to pull and install software from
+and never forks, archives, or modifies the software in any way. In doing so, we compel participating software projects
+to adhere to well-established distribution standards and ensure the software is more sustainable towards the future
+\cite{softwarequality}. Moreover, we ensure that LaMachine never becomes a prerequisite for the software but merely a
+courtesy or convenience, and attempt to limit any amount of duplication in packaging and distribution efforts.
 
@@
     \item \textbf{Local installation} - This installs the bulk of LaMachine in a separate local virtual environment (a seperate
-        directory) that has to be explicitly activated to be used. This also allows for multiple different installations
+        directory\footnote{Python users should know we just use \texttt{virtualenv} for this, with some
+        additions of our own}) that has to be explicitly activated to be used. This also allows for multiple different installations
         of LaMachine on the same host system (e.g. for different users or with different software configurations).
@@
         installed globally so there is only one installation possible, multiple users will all make use of the same
-        installation.
+        installation. Unlike the local installation, we do not make use of a Python Virtual Environment in this flavour.
     \item \textbf{Docker container} - Installs LaMachine in a container. Containerisation separates the entire run-time
@@
     \item \textbf{Singularity container} - Another form of containerisation with more focus on security\footnote{reduced
-        privilege escalation}, making it more suitable than Docker for deployment on computing clusters. This flavour is
-        still under development.
+        privilege escalation}, making it more suitable than Docker for deployment on computing clusters. This flavour is still under development.
     \item \textbf{Virtual Machine} - Installs LaMachine as a virtual machine, i.e. through full virtualisation. This allows
@@
     \item \textbf{Remote provisioning} - Installs LaMachine on a remote dedicated server. This option is most suited for
-        hosting centres.
+        hosting centres and directly uses Ansible's remote provisioning abilities.
 \end{enumerate}
@@
 completely virtualised, containers still share the kernel with the host OS, and the two native installation flavours,
-local \& global, actually compile against the machine's distribution itself and thus offers the least amount of
+local and global, actually compile against the machine's distribution itself and thus offer the least amount of
 overhead.
 
-The two native options, local and global installation, support a variety of major GNU/Linux distributions: Debian,
-Ubuntu, Arch Linux, CentOS, Fedora and, to a more limited degree, we also support macOS (the local installation flavour), powered in part by the homebrew package
+The two native options support a variety of major GNU/Linux distributions: Debian, Ubuntu, Arch Linux, CentOS, Fedora
+and, to a more limited degree, we also support macOS, powered in part by the homebrew package
 manager\footnote{\url{https://brew.sh}; alternatives such as macports are not supported}. Support for macOS is limited because not all participating
-software supports it. Certain Linux Distributions that are derivatives of these \emph{may} also work, such as RedHat
+software supports it. Certain Linux Distributions that are derivatives of the aforementioned distributions \emph{may} also work, such as RedHat
 Enterprise Linux (CentOS) and Linux Mint (Ubuntu).
@@
         versions of all included software.
-    \item \textbf{development} - This installs the latest development versions of the included software.
-In practise, this usually means that software is pulled directly from the version controlled repository and compiled and installed from source.
-Due to the experimental nature, the development version of
+    \item \textbf{development} - This installs the latest development versions of the included software. In practise,
+        this usually means that software is pulled directly from the version-controlled repository is and compiled and installed from source. Due to the experimental nature, the development version of
         LaMachine may at times break and not install successfully.
@@
 
-Installation of LaMachine begins with a single bootstrap command\footnote{See \url{https://proycon.github.io/LaMachine}}
+Installation of LaMachine begins with a single \emph{bootstrap} command\footnote{See \url{https://proycon.github.io/LaMachine}}
 executed on the command line.  It can interactively query the user for her software preferences \emph{(stored as the
@@
 encouraging software developers to provide their codemeta metadata inside their source code repositories and using that
-directly. This in turn enables other tools to do proper service discovery and provenance logging.
+directly. CodeMeta is a linked data initiative that provides a mapping from/to various commonly used software metadata
+standards\footnote{such as DOAP, Github API, Debian packages, Python distutils, R packages, Ruby gems, Maven metadata, DataCite, WikiData}.
+All this metadata in LaMachine in turn enables other tools to do proper service discovery and provenance logging.
+
+The software metadata discussed here is of a generic and practical nature, it most importantly contains information regarding the version, source,
+licensing, and authors of the software. The underlying principle is to obtain sofware metadata from as close to the
+source as possible. This can be contrasted with other metadata efforts, also within CLARIN, to establish a more
+centralised metadata store\footnote{such as CLARIN's CMDI Component Registry;
+\url{https://www.clarin.eu/content/component-metadata}} with more specialised and
+research-oriented metadata. There is merit to both approaches as they serve distinct functions. But in order to best
+serve everybody's different metadata needs, these bottom-up and top-down processes need to come to some kind of
+synthesis, e.g. by having component registries do automatic harvesting of CodeMeta (or other) metadata, similar to what LaMachine
+does.
 
@@
 
-LaMachine exists since May 2015 and has been used extensively ever since by numerous users, in early 2018 version 2 was
-released which was a significant rewrite, powered by Ansible.
+LaMachine exists since May 2015 and has been used extensively ever since by numerous users. In early 2018 version 2 was
+released which was a significant redesign, powered by Ansible.
 
@@
 installation manifest lists all packages that will be installed, at the user's discretion. After the initial
-installation, the user can always add more packages using \texttt{lamachine-add}.
+installation, the user can always add more packages using \texttt{lamachine-add} or editing the installation manifest directly.
 
@@
 Language and Speech Technology, Radboud University Nijmegen. The majority of this sofware was either fully or partially developed under the auspices of CLARIN-NL or successor
-CLARIAH. Some software by other CLARIN-NL/CLARIAH partners is also included. This project is not limited to one research
+CLARIAH. Some software by other CLARIN-NL/CLARIAH partners is also included. LaMachine is not limited to one research
 group and is explicitly open to participation by other software providers, especially those
@@
 \begin{itemize}
- \item by the Language Machines Research Group and the Centre of Language and Speech Technology, Radboud University
+ \item by the Language Machines Research Group and the Centre of Language and Speech Technology, Radboud University,
      Nijmegen\footnote{Links: \url{https://languagemachines.github.io/timbl}, \url{https://languagemachines.github.io/mbt},
@@
  \begin{itemize}
-     \item \textbf{Timbl} - A memory-based machine learning
-         toolkit, and \textbf{Mbt}, a memory-based tagger based on timbl.
-     \item \textbf{Ucto} - A tokeniser
-     \item \textbf{Frog} - Frog is an integration of various memory-based natural language processing (NLP) modules
+     \item \textbf{Timbl} -- A memory-based machine learning
+         toolkit, and \textbf{Mbt}, a memory-based tagger based on timbl. Python bindings included as well.
+     \item \textbf{Ucto} -- A multilingual rule-based tokeniser. Python binding included as well.
+     \item \textbf{Frog} -- An integration of various memory-based natural language processing (NLP) modules
          developed for Dutch. It can do Part-of-Speech tagging, lemmatisation, named entity recogniton, shallow parsing,
          dependency parsing and morphological analysis. Also included in LaMachine;
-         \textbf{Toad}, Trainer Of All Data, training tools for Frog.
-     \item \textbf{Wopr} - Memory-based Word Predictor.
-     \item \textbf{CLAM} - Quickly build RESTful webservices, powers many webservices offered by LaMachine.
-     \item \textbf{FoLiA} - Format for Linguistic Annotation, with tools and libraries in/for Python and $C++$.
-     \item \textbf{FLAT} - FoLiA Linguistic Annotation Tool: a web-based linguistic annotation tool.
-     \item \textbf{PyNLPl} - Python Natural Language Processing Library.
+         Python bindings for Frog and \textbf{Toad}, Trainer Of All Data, training tools for Frog.
+     \item \textbf{Wopr} -- Memory-based Word Predictor.
+     \item \textbf{CLAM} -- Quickly build RESTful webservices, powers many webservices offered by LaMachine.
+     \item \textbf{FoLiA} -- Format for Linguistic Annotation \cite{FOLIAPAPER}, with tools and libraries in/for Python and $C++$.
+     \item \textbf{FLAT} -- FoLiA Linguistic Annotation Tool: a web-based linguistic annotation tool.
+     \item \textbf{PyNLPl} -- Python Natural Language Processing Library.
      \item \textbf{Colibri Core} -- Colibri core is an NLP tool as well as a C++ and Python library for working with
          basic linguistic constructions such as n-grams and skipgrams (i.e patterns with one or more gaps, either of
-         fixed or dynamic size) in a quick and memory-efficient way. -
-     \item \textbf{Gecco} - Generic Environment for Context-Aware Correction of Orthography - and \textbf{Valkuil.net},
-         an instantiation thereof for Dutch.
-     \item \textbf{PICCL} - PICCL: A set of workflows for corpus building through OCR, post-correction (through \textbf{TICCL}) and Natural Language Processing.
-     \item \textbf{Labirinto}- A web-based portal listing all
+         fixed or dynamic size) in a quick and memory-efficient way.
+     \item \textbf{Gecco} -- Generic Environment for Context-Aware Correction of Orthography, an NLP pipeline for
+         spelling correction,  and \textbf{Valkuil.net}, an instantiation thereof for Dutch.
+     \item \textbf{PICCL} -- A set of workflows (NLP pipeline) for corpus building through OCR, post-correction (through \textbf{TICCL}) and Natural Language Processing.
+     \item \textbf{Labirinto} -- A web-based portal listing all
          available tools in LaMachine, an ideal starting point for LaMachine.
-     \item \textbf{Oersetter} - A Frisian-Dutch Machine Translation system in collaboration with the Fryske Akademy.
+     \item \textbf{Oersetter} -- A Frisian-Dutch Machine Translation system in collaboration with the Fryske Akademy.
  \end{itemize}
@@
  \begin{itemize}
-     \item \textbf{Alpino}\footnote{\url{http://www.let.rug.nl/vannoord/alp/Alpino/}} - a dependency parser and tagger for Dutch
+     \item \textbf{Alpino}\footnote{\url{http://www.let.rug.nl/vannoord/alp/Alpino/}} - A dependency parser and tagger for Dutch.
  \end{itemize}
@@
  \begin{itemize}
-     \item \textbf{KafNafParserPy}\footnote{\url{https://github.com/cltl/KafNafParserPy}} - A python module to parse NAF files
+     \item \textbf{KafNafParserPy}\footnote{\url{https://github.com/cltl/KafNafParserPy}} - A python module to parse NAF files.
  \end{itemize}
@@
  \end{itemize}
- \item by Meertens Instituut
+ \item by the Meertens Instituut
  \begin{itemize}
      \item \textbf{Python Course for the Humanities}\footnote{\url{http://www.karsdorp.io/python-course/}} - Interactive
-         tutorial and introduction into programming with Python for the humanities
+         tutorial and introduction into programming with Python for the humanities.
  \end{itemize}
@@
 In addition to the above listed specific software, LaMachine also incorporates a large number of renowned tools by external international parties, offering
-most notably a mature Python environment with renowned scientific modules. The following list is not
+most notably a mature Python environment with scientific modules. The following list gives an impression and is not
 exhaustive: \footnote{Links: \url{https://jupyterlab.readthedocs.io/en/stable/}, \url{https://pytorch.org}, \url{http://www.nltk.org},
@@
 We already addressed the need for different interfaces for different audiences in Section~\ref{sec:intro}. The challenge we
-face, and for which LaMachine offers a solution, is one of software maintainability and distribution. How do we maintain
-and distribute software that is often highly complex and consists of multiple interconnected components, considering
+face, and for which LaMachine offers a solution, is one of software maintainability, distribution and deployment. How do
+we maintain, distribute, and deploy software that is often highly complex and consists of multiple interconnected components, considering
 that this software is used differently by different audiences? A key aspect here is the reusability and accessibility of individual

### 2019-01-25 docs/lamachine_clarin2018.tex
paper updates + made a start on the 2nd use case

@@
 of the tools that offer a text interface, including text editors such as vim, emacs or nano, version control systems
-such as git and interpreters such as Python or R (or compilers for that matter). The latter in turn allow users to use
-one of the many specialised programming libraries included in LaMachine to build their own tools. All this makes
+such as git and interpreters such as Python or R (or compilers for that matter). This allow users to use
+one of the many specialised programming libraries included in LaMachine to build their own tools and . All this makes
 LaMachine ideally suited as a development environment.
@@
 only interested in a single Python package that LaMachine offers? Then she can just as well install it directly using
-\texttt{pip install}. Is the user only interested in a single package that underlying Linux distribution (or macOS' homebrew)
+\texttt{pip install}. Is the user only interested in a single package that  the underlying Linux distribution (or
+macOS's homebrew)
 already provides? Then, by all means, she should go for that.
@@
 
-The best comparison can be drawn with Anaconda\footnote{\url{https://www.anaconda.com/}}, which we could also qualify as
+The best comparison can perhaps be drawn with Anaconda\footnote{\url{https://www.anaconda.com/}}, which we could also qualify as
 a meta-distribution under our definition, and which enjoys great popularity in the data science community. Anaconda has
 a focus on Python and R. It is much wider in scope than LaMachine, which has a strong NLP and CLARIN/CLARIAH focus. When
-it comes to Python, there is considerable overlap in the modules that LaMachine and Anaconda offer. Unlike LaMachine,
+it comes to Python, which Anaconda is initially geared at, there is considerable overlap in the modules that LaMachine and Anaconda offer. Unlike LaMachine,
 Anaconda does not focus on providing different flavours such as a VM or a Docker Container\footnote{This is not a
 technical limitation but simply a matter of different objectives, one could easily create a VM or Docker Container and
-install Anaconda in it.}.
+install Anaconda in it.}, nor separate stable and development versions of the included software.
 
@@
 containers, i.e. containing different software of having different configurations, based on the user's needs. It is
-quite feasible to use instantiate a variety of LaMachine containers and use a container orchestration system such
+quite feasible to instantiate a variety of LaMachine containers and use a container orchestration system such
 Kubernetes or Docker Swarm for automated deployment, scaling and management thereof. Such orchestration, however, is
@@
 
-%TODO
+(TODO) %TODO: some intro
+
+For this workshop, computing power was purchased at SURFsara, a collaborative organization for ICT in Dutch education
+and research. Subsequently, we had the ability to create Virtual Machines on their hosting platform. For this workshop
+we had twelve participants and decided to create a single multi-core and hih memory VM to share amongst the
+participants.
+
+As recommended by SURFsara, we opted for one of their default Linux images based on Ubuntu 18.04, and we bootstrapped
+the local installation flavor of LaMachine inside that VM.
+
+(TODO) %TODO: some intro
 

### 2019-01-28 docs/lamachine_clarin2018.tex
paper update

@@
 \section{Case studies}
-\label{sec:case}
 
 \subsection{Text Mining for Health Inspection}
+\label{sec:case}
 
@@
 \subsection{Workshop: Cataloguing of Textual Cultural Heritage Objects}
+\label{sec:case2}
 
@@
 and research. Subsequently, we had the ability to create Virtual Machines on their hosting platform. For this workshop
-we had twelve participants and decided to create a single multi-core and hih memory VM to share amongst the
+we had twelve participants and decided to create a single multi-core and high memory VM to share amongst the
 participants.
 
-As recommended by SURFsara, we opted for one of their default Linux images based on Ubuntu 18.04, and we bootstrapped
-the local installation flavor of LaMachine inside that VM.
+As recommended by SURFsara, we opted for one of their default Linux images as a basis for the VM, based on Ubuntu 18.04,
+instead of providing a LaMachine VM image directly like we did in \section{sec:case}. This was recommended because their image
+was already preconfigured to integrate nicely with their cloud environment. Inside this VM we simply bootstrapped the local
+installation flavour of LaMachine.  Here we benefit from the flexibility LaMachine offers because of its various
+flavours, and regardless of the flavour, the resulting installations are always functionally equivalent. The local
+environment flavour had the added bonus of not being in anyone's way in case any of the twelve workshop participants did
+not want to use LaMachine, considering it needs to be explicitly activated prior to usage.
 

### 2019-01-28 docs/lamachine_clarin2018.tex
fix

@@
 As recommended by SURFsara, we opted for one of their default Linux images as a basis for the VM, based on Ubuntu 18.04,
-instead of providing a LaMachine VM image directly like we did in \section{sec:case}. This was recommended because their image
+instead of providing a LaMachine VM image directly like we did in~\ref{sec:case}. This was recommended because their image
 was already preconfigured to integrate nicely with their cloud environment. Inside this VM we simply bootstrapped the local

### 2019-01-28 docs/lamachine_clarin2018.tex
minor edits in intro and enhancement of abstract

@@
 community.
-We illustrate the usage of LaMachine in an exploratory text mining project at the Dutch Health Inspectorate where LaMachine was applied to create a research environment for automatic text analysis for health care quality monitoring.
+We explain what Lamachine is, how it cn used and the technical details.
+We also compare Lamachine to alternative software distributions and discuss its advantages and limitations.
+We illustrate how LaMachine can be used in two case studies, one in an exploratory text mining project at the Dutch Health Inspectorate where LaMachine was applied to create a research environment for automatic text analysis for health care quality monitoring, and a second case where Lamachine was used to create a workspace for a one-week, intense collaboration by a diverse research team.
 \end{abstract}
@@
 
-Software is a key deliverable and a vital component for research in projects such as those under the CLARIN umbrella. It
-is software that provides researchers the instruments to yield for their research. It is CLARIN's core mission to make
+Software is a key deliverable and a vital component for research in projects such as those under the CLARIN umbrella.
+Software provides researchers the instruments to yield for their research.
+It is CLARIN's core mission to make
 digital language resources, including software, available to the wider research community.
@@
 environment for a data scientist, whether on a personal computer or on a computing cluster, a good development
-environment for a developer or a good deployment method for production servers in for example CLARIN centres.  We
-demonstrate how LaMachine can create a fully functioning and standalone research environment for text mining  and NLP
-for Dutch texts in a use case project at the Healthcare Inspectorate.
+environment for a developer or a good deployment method for production servers in for example CLARIN centres.
+
+ This paper provides a detailed description of Lamachine, its purpose and functionality and the technical details.
+We discuss the advantages and limitations of Lamachine to alternative software distributions in Section \ref{sec:comparison}.
+We demonstrate how LaMachine can create a fully functioning and standalone research environment for text mining  and NLP for Dutch texts in two use cases in Section \ref{sec:case}.
 
@@
 
-\section{Comparisons and alternatives}
+\section{Comparison to alternative distributions}
 \label{sec:comparisons}
@@
 
-\section{Case studies}
+\section{Case studies}\label{sec:case}
 
 \subsection{Text Mining for Health Inspection}
-\label{sec:case}
 

### 2019-01-28 docs/lamachine_clarin2018.tex
major edits in the description - attempt to move the technical details further behind

@@
 
-\section{Architecture}
+\section{What is Lamachine}
 
-LaMachine is an open-source NLP software distribution. This implies the following in our case:
-
-\begin{enumerate}
-    \item LaMachine facilitates the installation, distribution and configuration of software. It does not fork, modify
-        or appropriate the participating software in any way, nor does it provide a hosting place or repository for
-        software.
-    \item The software included in LaMachine must under a recognised (i.e. OSI-approved) open-source license.
-        Proprietary (closed-source) software is explicitly excluded.
-    \item The source
-        code must be hosted in a public version controlled repository. \footnote{e.g. Github, Gitlab, Bitbucket, provided
-        the respository is public}
-    \item There must be a build process to compile the source, if applicable, and install the program or library.
-    \item There must be some kind of release protocol (adhering to semantic versioning) that publishes software using the proper
-technology-specific channels. We will elaborate on this later.
-    \item All software that is incorporated in LaMachine must bear at least some relevance to the field of Natural Language Processing.
-    \item LaMachine lives in an open-source ecosystem and therefore builds on Unix-like platforms; this primarily
-        means Linux, as well as BSD and, with some restrictions, macOS. This by definition excludes certain software for different platforms, such as
-        mobile platforms (Android/iOS/etc), native Windows/mac desktop applications, or certain interface types in
-        general such as classical desktop GUI applications or mobile `apps', all of which fall beyond our scope.
-        Cygwin\footnote{A unix environment on Windows} is not tested or
-        supported either.  However, virtualisation technology enables deployment on a wider range of platforms, including Windows.
-\end{enumerate}
-
+LaMachine is an open-source NLP software distribution.
+LaMachine facilitates the installation, distribution and configuration of software.
+It does not fork, modify or appropriate the participating software in any way, nor does it provide a hosting place or repository for software.
 We classify LaMachine as a \emph{meta distribution} as it can be installed in various contexts. The heart of LaMachine consists of a
 set of machine-parsable instructions on how to obtain, build (e.g. compile from source), install and configure software.
-These are implemented using Ansible\footnote{\url{https://www.ansible.com}; a popular open-source software provisioning,
-configuration and deployment tool by Red Hat}.  This is notably different from the more
+%iris:  eerst algemene uitleg over lamachine- dit is veel te gedetaileerd op dit punt-
+%These are implemented using Ansible\footnote{\url{https://www.ansible.com}; a popular open-source software provisioning,
+%configuration and deployment tool by Red Hat}.  
+This is notably different from the more
 classical notion of Linux distributions, which generally provide their own repositories with (often binary) software
@@
 courtesy or convenience, and attempt to limit any amount of duplication in packaging and distribution efforts.
+
+LaMachine lives in an open-source ecosystem and therefore builds on Unix-like platforms; this primarily means Linux, as well as BSD and, with some restrictions, macOS.
+This by definition excludes certain software for different platforms, such as mobile platforms (Android/iOS/etc), native Windows/mac desktop applications, or certain interface types in general such as classical desktop GUI applications or mobile `apps', all of which fall beyond our scope.
+Cygwin\footnote{A unix environment on Windows} is not tested or
+ supported either.  However, virtualisation technology enables deployment on a wider range of platforms, including Windows.
+
+The software included in LaMachine has to live up (iris:ik zoek synoniem hier voor live up) to the following prerequisites:
+\begin{enumerate}
+
+    \item The software must have a recognised (i.e. OSI-approved) open-source license.
+        Proprietary (closed-source) software is explicitly excluded.
+    \item The source
+        code must be hosted in a public version controlled repository. \footnote{e.g. Github, Gitlab, Bitbucket, provided
+        the respository is public}
+    \item There must be a build process to compile the source, if applicable, and install the program or library.
+    \item There must be some kind of release protocol (adhering to semantic versioning) that publishes software using the proper
+technology-specific channels. We will elaborate on this later.
+    \item All software that is incorporated in LaMachine must bear at least some relevance to the field of Natural Language Processing.
+    \item {\bf maarten help even met formuleren hier, ik bedoel dat zoiets als Pattern dus helaas niet mee kan doen omdat het niet meer onderhouden wordt }only maintained/uptodate date software that can compile using the current version of compilers and dependencies/libraries
+\end{enumerate}
+
+\section{ Included Software}
+
+LaMachine exists since May 2015 and has been used extensively ever since by numerous users. In early 2018 version 2 was
+released which was a significant redesign, powered by Ansible.
+
+In LaMachine, software is grouped into various ``packages'', each package\footnote{For those familiar with
+Ansible, a package in LaMachine is an Ansible role} groups one or multiple programs that have some kind of relation. The
+installation manifest lists all packages that will be installed, at the user's discretion. After the initial
+installation, the user can always add more packages using \texttt{lamachine-add} or editing the installation manifest directly.
+
+LaMachine was initially conceived as the primary means of distribution of the software stack developed at the Language
+Machines Research Group and the Centre of
+Language and Speech Technology, Radboud University Nijmegen. The majority of this sofware was either fully or partially developed under the auspices of CLARIN-NL or successor
+CLARIAH. Some software by other CLARIN-NL/CLARIAH partners is also included. LaMachine is not limited to one research
+group and is explicitly open to participation by other software providers, especially those
+also in CLARIAH.
+
+We list a selection of the most important software included in LaMachine, grouped by research institute:
+
+\begin{itemize}
+ \item by the Language Machines Research Group and the Centre of Language and Speech Technology, Radboud University,
+     Nijmegen\footnote{Links: \url{https://languagemachines.github.io/timbl}, \url{https://languagemachines.github.io/mbt},
+    \url{https://languagemachines.github.io/ucto}, \url{https://languagemachines.github.io/frog},
+    \url{http://ilk.uvt.nl/wopr}, \url{https://proycon.github.io/folia}, \url{https://proycon.github.io/clam},
+\url{https://github.com/proycon/flat}, \url{https://proycon.anaproy.nl/pynlpl},
+\url{https://proycon.github.io/colibri-core/},
+\url{https://github.com/proycon/gecco},
+\url{https://github.com/proycon/valkuil-gecco},
+\url{https://github.com/LanguageMachines/PICCL}, \url{https://github.com/LanguageMachines/ticcltools},
+\url{https://github.com/proycon/labirinto},
+\url{https://github.com/proycon/oersetter-webservice}
+}
+ \begin{itemize}
+     \item \textbf{Timbl} -- A memory-based machine learning
+         toolkit, and \textbf{Mbt}, a memory-based tagger based on timbl. Python bindings included as well.
+     \item \textbf{Ucto} -- A multilingual rule-based tokeniser. Python binding included as well.
+     \item \textbf{Frog} -- An integration of various memory-based natural language processing (NLP) modules
+         developed for Dutch. It can do Part-of-Speech tagging, lemmatisation, named entity recogniton, shallow parsing,
+         dependency parsing and morphological analysis. Also included in LaMachine;
+         Python bindings for Frog and \textbf{Toad}, Trainer Of All Data, training tools for Frog.
+     \item \textbf{Wopr} -- Memory-based Word Predictor.
+     \item \textbf{CLAM} -- Quickly build RESTful webservices, powers many webservices offered by LaMachine.
+     \item \textbf{FoLiA} -- Format for Linguistic Annotation \cite{FOLIAPAPER}, with tools and libraries in/for Python and $C++$.
+     \item \textbf{FLAT} -- FoLiA Linguistic Annotation Tool: a web-based linguistic annotation tool.
+     \item \textbf{PyNLPl} -- Python Natural Language Processing Library.
+     \item \textbf{Colibri Core} -- Colibri core is an NLP tool as well as a C++ and Python library for working with
+         basic linguistic constructions such as n-grams and skipgrams (i.e patterns with one or more gaps, either of
+         fixed or dynamic size) in a quick and memory-efficient way.
+     \item \textbf{Gecco} -- Generic Environment for Context-Aware Correction of Orthography, an NLP pipeline for
+         spelling correction,  and \textbf{Valkuil.net}, an instantiation thereof for Dutch.
+     \item \textbf{PICCL} -- A set of workflows (NLP pipeline) for corpus building through OCR, post-correction (through \textbf{TICCL}) and Natural Language Processing.
+     \item \textbf{Labirinto} -- A web-based portal listing all
+         available tools in LaMachine, an ideal starting point for LaMachine.
+     \item \textbf{Oersetter} -- A Frisian-Dutch Machine Translation system in collaboration with the Fryske Akademy.
+ \end{itemize}
+ \item by the University of Groningen
+ \begin{itemize}
+     \item \textbf{Alpino}\footnote{\url{http://www.let.rug.nl/vannoord/alp/Alpino/}} - A dependency parser and tagger for Dutch.
+ \end{itemize}
+ \item by the Vrije Universiteit Amsterdam
+ \begin{itemize}
+     \item \textbf{KafNafParserPy}\footnote{\url{https://github.com/cltl/KafNafParserPy}} - A python module to parse NAF files.
+ \end{itemize}
+ \item by Utrecht University
+ \begin{itemize}
+     \item  \textbf{T-scan}\footnote{\url{https://github.com/proycon/tscan}} - T-scan is a Dutch text analytics tool for readability prediction (initially developed at TiCC, Tilburg University, and in collaboration with Radboud University, Nijmegen)
+ \end{itemize}
+ \item by the Meertens Instituut
+ \begin{itemize}
+     \item \textbf{Python Course for the Humanities}\footnote{\url{http://www.karsdorp.io/python-course/}} - Interactive
+         tutorial and introduction into programming with Python for the humanities.
+ \end{itemize}
+\end{itemize}
+
+In addition to the above listed specific software, LaMachine also incorporates a large number of renowned tools by external international parties, offering
+most notably a mature Python environment with scientific modules. The following list gives an impression and is not
+exhaustive: \footnote{Links: \url{https://jupyterlab.readthedocs.io/en/stable/}, \url{https://pytorch.org}, \url{http://www.nltk.org},
+    \url{https://spacy.io},  \url{http://www.nextflow.io}, \url{https://stanfordnlp.github.io/CoreNLP/},  \url{https://github.com/tesseract-ocr/tesseract}, \url{https://tensorflow.org},
+\url{http://kaldi-asr.org}, \url{http://www.statmt.org/moses}}
+
+\begin{itemize}
+    \item \textbf{Python}: Numpy, Scipy, Matplotlib, Scikit-Learn, IPython, Jupyter, ...
+    \begin{itemize}
+        \item \textbf{Jupyter Lab} The successor of the popular Jupyter Notebooks, offers notebooks, a web-based IDE, terminals. An ideal entry point to get started with LaMachine and all it contains!
+        \item \textbf{PyTorch} - Deep-learning library for Python
+        \item \textbf{NLTK} - Natural Language Toolkit for Python
+        \item \textbf{Spacy} - Industrial-Strength NLP in Python
+    \end{itemize}
+    \item \textbf{R}
+    \item \textbf{Java}
+    \begin{itemize}
+        \item \textbf{NextFlow} - A system and language for writing parallel and scalable pipelines in a portable manner.
+        \item \textbf{Stanford CoreNLP} - Various types of linguistic enrichment
+    \end{itemize}
+    \item \textbf{Tesseract} - Open Source Optical Character Recognition (OCR)
+    \item \textbf{Tensorflow} - Open-source machine learning framework
+    \item \textbf{Kaldi} - Speech Recognition Framework (ASR)
+    \item \textbf{Moses} - Statistical Machine Translation system
+\end{itemize}
+
+\section{Architecture}
+
+In this section we present the technical design choices that were made.
+%iris: Maarten je zou hier ook expliciet de motivatie voor de keuzen kunnen toevoegen
 
@@
 \FloatBarrier
-
-\section{Software}
-
-LaMachine exists since May 2015 and has been used extensively ever since by numerous users. In early 2018 version 2 was
-released which was a significant redesign, powered by Ansible.
-
-In LaMachine, software is grouped into various ``packages'', each package\footnote{For those familiar with
-Ansible, a package in LaMachine is an Ansible role} groups one or multiple programs that have some kind of relation. The
-installation manifest lists all packages that will be installed, at the user's discretion. After the initial
-installation, the user can always add more packages using \texttt{lamachine-add} or editing the installation manifest directly.
-
-LaMachine was initially conceived as the primary means of distribution of the software stack developed at the Language
-Machines Research Group and the Centre of
-Language and Speech Technology, Radboud University Nijmegen. The majority of this sofware was either fully or partially developed under the auspices of CLARIN-NL or successor
-CLARIAH. Some software by other CLARIN-NL/CLARIAH partners is also included. LaMachine is not limited to one research
-group and is explicitly open to participation by other software providers, especially those
-also in CLARIAH.
-
-We list a selection of the most important software included in LaMachine, grouped by research institute:
-
-\begin{itemize}
- \item by the Language Machines Research Group and the Centre of Language and Speech Technology, Radboud University,
-     Nijmegen\footnote{Links: \url{https://languagemachines.github.io/timbl}, \url{https://languagemachines.github.io/mbt},
-    \url{https://languagemachines.github.io/ucto}, \url{https://languagemachines.github.io/frog},
-    \url{http://ilk.uvt.nl/wopr}, \url{https://proycon.github.io/folia}, \url{https://proycon.github.io/clam},
-\url{https://github.com/proycon/flat}, \url{https://proycon.anaproy.nl/pynlpl},
-\url{https://proycon.github.io/colibri-core/},
-\url{https://github.com/proycon/gecco},
-\url{https://github.com/proycon/valkuil-gecco},
-\url{https://github.com/LanguageMachines/PICCL}, \url{https://github.com/LanguageMachines/ticcltools},
-\url{https://github.com/proycon/labirinto},
-\url{https://github.com/proycon/oersetter-webservice}
-}
- \begin{itemize}
-     \item \textbf{Timbl} -- A memory-based machine learning
-         toolkit, and \textbf{Mbt}, a memory-based tagger based on timbl. Python bindings included as well.
-     \item \textbf{Ucto} -- A multilingual rule-based tokeniser. Python binding included as well.
-     \item \textbf{Frog} -- An integration of various memory-based natural language processing (NLP) modules
-         developed for Dutch. It can do Part-of-Speech tagging, lemmatisation, named entity recogniton, shallow parsing,
-         dependency parsing and morphological analysis. Also included in LaMachine;
-         Python bindings for Frog and \textbf{Toad}, Trainer Of All Data, training tools for Frog.
-     \item \textbf{Wopr} -- Memory-based Word Predictor.
-     \item \textbf{CLAM} -- Quickly build RESTful webservices, powers many webservices offered by LaMachine.
-     \item \textbf{FoLiA} -- Format for Linguistic Annotation \cite{FOLIAPAPER}, with tools and libraries in/for Python and $C++$.
-     \item \textbf{FLAT} -- FoLiA Linguistic Annotation Tool: a web-based linguistic annotation tool.
-     \item \textbf{PyNLPl} -- Python Natural Language Processing Library.
-     \item \textbf{Colibri Core} -- Colibri core is an NLP tool as well as a C++ and Python library for working with
-         basic linguistic constructions such as n-grams and skipgrams (i.e patterns with one or more gaps, either of
-         fixed or dynamic size) in a quick and memory-efficient way.
-     \item \textbf{Gecco} -- Generic Environment for Context-Aware Correction of Orthography, an NLP pipeline for
-         spelling correction,  and \textbf{Valkuil.net}, an instantiation thereof for Dutch.
-     \item \textbf{PICCL} -- A set of workflows (NLP pipeline) for corpus building through OCR, post-correction (through \textbf{TICCL}) and Natural Language Processing.
-     \item \textbf{Labirinto} -- A web-based portal listing all
-         available tools in LaMachine, an ideal starting point for LaMachine.
-     \item \textbf{Oersetter} -- A Frisian-Dutch Machine Translation system in collaboration with the Fryske Akademy.
- \end{itemize}
- \item by the University of Groningen
- \begin{itemize}
-     \item \textbf{Alpino}\footnote{\url{http://www.let.rug.nl/vannoord/alp/Alpino/}} - A dependency parser and tagger for Dutch.
- \end{itemize}
- \item by the Vrije Universiteit Amsterdam
- \begin{itemize}
-     \item \textbf{KafNafParserPy}\footnote{\url{https://github.com/cltl/KafNafParserPy}} - A python module to parse NAF files.
- \end{itemize}
- \item by Utrecht University
- \begin{itemize}
-     \item  \textbf{T-scan}\footnote{\url{https://github.com/proycon/tscan}} - T-scan is a Dutch text analytics tool for readability prediction (initially developed at TiCC, Tilburg University, and in collaboration with Radboud University, Nijmegen)
- \end{itemize}
- \item by the Meertens Instituut
- \begin{itemize}
-     \item \textbf{Python Course for the Humanities}\footnote{\url{http://www.karsdorp.io/python-course/}} - Interactive
-         tutorial and introduction into programming with Python for the humanities.
- \end{itemize}
-\end{itemize}
-
-In addition to the above listed specific software, LaMachine also incorporates a large number of renowned tools by external international parties, offering
-most notably a mature Python environment with scientific modules. The following list gives an impression and is not
-exhaustive: \footnote{Links: \url{https://jupyterlab.readthedocs.io/en/stable/}, \url{https://pytorch.org}, \url{http://www.nltk.org},
-    \url{https://spacy.io},  \url{http://www.nextflow.io}, \url{https://stanfordnlp.github.io/CoreNLP/},  \url{https://github.com/tesseract-ocr/tesseract}, \url{https://tensorflow.org},
-\url{http://kaldi-asr.org}, \url{http://www.statmt.org/moses}}
-
-\begin{itemize}
-    \item \textbf{Python}: Numpy, Scipy, Matplotlib, Scikit-Learn, IPython, Jupyter, ...
-    \begin{itemize}
-        \item \textbf{Jupyter Lab} The successor of the popular Jupyter Notebooks, offers notebooks, a web-based IDE, terminals. An ideal entry point to get started with LaMachine and all it contains!
-        \item \textbf{PyTorch} - Deep-learning library for Python
-        \item \textbf{NLTK} - Natural Language Toolkit for Python
-        \item \textbf{Spacy} - Industrial-Strength NLP in Python
-    \end{itemize}
-    \item \textbf{R}
-    \item \textbf{Java}
-    \begin{itemize}
-        \item \textbf{NextFlow} - A system and language for writing parallel and scalable pipelines in a portable manner.
-        \item \textbf{Stanford CoreNLP} - Various types of linguistic enrichment
-    \end{itemize}
-    \item \textbf{Tesseract} - Open Source Optical Character Recognition (OCR)
-    \item \textbf{Tensorflow} - Open-source machine learning framework
-    \item \textbf{Kaldi} - Speech Recognition Framework (ASR)
-    \item \textbf{Moses} - Statistical Machine Translation system
-\end{itemize}
 
@@
 
-(TODO) %TODO: some intro
+The ICT-Research Platform Netherlands and NWO organize a yearly one-week workshop `ICT with Industry' \footnote{https://ict-research.nl/ict-with-industry/ictwi2019/} to stimulate collaboration between industry and academia.
+The industrical partner provides a problem and a team of researchers from different backgrounds and universities collaborate to come up with solutions.
+We partipated in the 2019 edition on the case study by the Dutch Royal Library who
 
@@
 we had twelve participants and decided to create a single multi-core and high memory VM to share amongst the
-participants.
+participants to create one common digital workspace.
 
@@
 
-(TODO) %TODO: some intro
+Lamachine offered a convenient platform for a range of different explorations and experiments in the area of NLP and text mining. 
+However, for some situations Lamachine was not a good fit for the tasks at hand: For team members who did not have experience with a non-Windows environment, Lamachine was not a suitable or useful tool.
+And also for members who wanted to use a graphical user interface for their text editors as we did not manage to get the x-forwarding working (note that this also depends on the overall VM).
+After a few attempts the team gave up on resolving this issue die to time pressure but it also demonstrated that finetuning the configuration of Lamachine on certain aspects is beyond the reach of a data scientist without system administration skills.
+ %iris it moet nog wat beter geformuleerd worden maar je snaprt wat ik bedoel
 

### 2019-01-28 docs/lamachine_clarin2018.tex
[paper] minor edits after feedback from iris

@@
 community.
-We explain what Lamachine is, how it cn used and the technical details.
+We explain what Lamachine is, how it can be used and the technical details.
 We also compare Lamachine to alternative software distributions and discuss its advantages and limitations.
@@
 
- This paper provides a detailed description of Lamachine, its purpose and functionality and the technical details.
-We discuss the advantages and limitations of Lamachine to alternative software distributions in Section \ref{sec:comparison}.
+ This paper provides a detailed description of Lamachine, its purpose and functionality and the technical details. We
+ discuss the advantages and limitations of LaMachine compared to alternative software distributions in Section \ref{sec:comparison}.
 We demonstrate how LaMachine can create a fully functioning and standalone research environment for text mining  and NLP for Dutch texts in two use cases in Section \ref{sec:case}.
 
-\section{What is Lamachine}
+ \section{What is Lamachine?}
 
@@
 CRAN\footnote{\url{https://cran.r-project.org/}} for R, CPAN\footnote{\url{https://www.cpan.org}} for Perl, Maven
-Central\footnote{\url{https://search.maven.org}} for Java.  LaMachine relies on those to pull and install software from
-and never forks, archives, or modifies the software in any way. In doing so, we compel participating software projects
+Central\footnote{\url{https://search.maven.org}} for Java.  LaMachine relies on those to obtain and install software. In doing so, we compel participating software projects
 to adhere to well-established distribution standards and ensure the software is more sustainable towards the future
@@
 
-The software included in LaMachine has to live up (iris:ik zoek synoniem hier voor live up) to the following prerequisites:
+The software included in LaMachine has to adhere to the following prerequisites:
+
 \begin{enumerate}
@@
     \item All software that is incorporated in LaMachine must bear at least some relevance to the field of Natural Language Processing.
-    \item {\bf maarten help even met formuleren hier, ik bedoel dat zoiets als Pattern dus helaas niet mee kan doen omdat het niet meer onderhouden wordt }only maintained/uptodate date software that can compile using the current version of compilers and dependencies/libraries
+    \item Participating software must be actively maintained (i.e. not outdated or abandoned) and not place any demands on outdated
+        dependencies.
+%verwerkt <-- maarten help even met formuleren hier, ik bedoel dat zoiets als Pattern dus helaas niet mee kan doen omdat het niet meer onderhouden wordt }only maintained/uptodate date software that can compile using the current version of compilers and dependencies/libraries
 \end{enumerate}
@@
 
-In this section we present the technical design choices that were made.
+In this section we present the technical design choices that were made and lay out the architecture of LaMachine.
 %iris: Maarten je zou hier ook expliciet de motivatie voor de keuzen kunnen toevoegen
@@
 \begin{enumerate}
-    \item \textbf{Local installation} - This installs the bulk of LaMachine in a separate local virtual environment (a seperate
+    \item \textbf{Local installation} -- This installs the bulk of LaMachine in a separate local virtual environment (a seperate
         directory\footnote{Python users should know we just use \texttt{virtualenv} for this, with some
@@
         package manager of your distribution.
-    \item \textbf{Global installation} - This flavour is used for a host that is fully dedicated to LaMachine. Everything will be
+    \item \textbf{Global installation} -- This flavour is used for a host that is fully dedicated to LaMachine. Everything will be
         installed globally so there is only one installation possible, multiple users will all make use of the same
         installation. Unlike the local installation, we do not make use of a Python Virtual Environment in this flavour.
-    \item \textbf{Docker container} - Installs LaMachine in a container. Containerisation separates the entire run-time
+    \item \textbf{Docker container} -- Installs LaMachine in a container. Containerisation separates the entire run-time
         environment from the host system, only the Linux kernel is shared. It is a lighter option than full
@@
         virtualisation.}
-    \item \textbf{Singularity container} - Another form of containerisation with more focus on security\footnote{reduced
+    \item \textbf{Singularity container} -- Another form of containerisation with more focus on security\footnote{reduced
         privilege escalation}, making it more suitable than Docker for deployment on computing clusters. This flavour is still under development.
-    \item \textbf{Virtual Machine} - Installs LaMachine as a virtual machine, i.e. through full virtualisation. This allows
+    \item \textbf{Virtual Machine} --- Installs LaMachine as a virtual machine, i.e. through full virtualisation. This allows
         deployment of LaMachine on hosts which would otherwise not support it, such as Windows. Virtualisation for
         LaMachine is achieved through Vagrant and VirtualBox\footnote{\url{https://vagrant.org}, \url{https://www.virtualbox.org}}.
-    \item \textbf{Remote provisioning} - Installs LaMachine on a remote dedicated server. This option is most suited for
+    \item \textbf{Remote provisioning} -- Installs LaMachine on a remote dedicated server. This option is most suited for
         hosting centres and directly uses Ansible's remote provisioning abilities.
@@
 
-The ICT-Research Platform Netherlands and NWO organize a yearly one-week workshop `ICT with Industry' \footnote{https://ict-research.nl/ict-with-industry/ictwi2019/} to stimulate collaboration between industry and academia.
-The industrical partner provides a problem and a team of researchers from different backgrounds and universities collaborate to come up with solutions.
-We partipated in the 2019 edition on the case study by the Dutch Royal Library who
+The ICT-Research Platform Netherlands and NWO organise a yearly one-week workshop `ICT with Industry'
+\footnote{https://ict-research.nl/ict-with-industry/ictwi2019/} to stimulate collaboration between industry and
+academia. The industrial partner provides a problem and a team of researchers from different backgrounds and
+universities collaborate to come up with solutions. We participated in the 2019 edition on the case study by the Dutch
+Royal Library who %@Iris: zin niet afgemaakt?
 
@@
 
-Lamachine offered a convenient platform for a range of different explorations and experiments in the area of NLP and text mining. 
-However, for some situations Lamachine was not a good fit for the tasks at hand: For team members who did not have experience with a non-Windows environment, Lamachine was not a suitable or useful tool.
-And also for members who wanted to use a graphical user interface for their text editors as we did not manage to get the x-forwarding working (note that this also depends on the overall VM).
-After a few attempts the team gave up on resolving this issue die to time pressure but it also demonstrated that finetuning the configuration of Lamachine on certain aspects is beyond the reach of a data scientist without system administration skills.
+LaMachine offered a convenient platform for a range of different explorations and experiments in the area of NLP and text mining.
+However, for some situations LaMachine was not a good fit for the tasks at hand: For team members who did not have
+experience with a non-Windows environment, LaMachine was not a suitable or useful tool. And also for members who wanted
+to use a graphical user interface for their text editors as we did not manage to get the x-forwarding working (note that
+this also depends on the overall VM). After a few attempts the team gave up on resolving this issue due to time pressure
+but it also demonstrated that fine-tuning the configuration of LaMachine on certain aspects is beyond the reach of a data scientist without system administration skills.
  %iris it moet nog wat beter geformuleerd worden maar je snaprt wat ik bedoel

### 2019-01-28 docs/lamachine_clarin2018.tex
[paper] more major edits after feedback by iris; edited case study 2 and added a paragraph on maintainability of LaMachine

@@
 
+The flexibility LaMachine offers, with various flavours and versions, makes it more accessible and deployable in
+multiple contexts, but this does come at a cost. Maintaining LaMachine itself is a non-trivial task that requires
+constant maintenance and testing\footnote{We run automated tested on a continuous integration platform to this end};
+software exists in an ever moving ecosystem. Any of the distributions we target, or software that participates, \emph{might}
+at any time introduce a change that requires us to adapt LaMachine accordingly. Similarly, underlying Linux
+distribution releases that are currently up-to-date and supported, will eventually be deprecated and replaced by newer releases, a development
+LaMachine is obliged to follow by design.
+
 \section{Case studies}\label{sec:case}
@@
 
-LaMachine offered a convenient platform for a range of different explorations and experiments in the area of NLP and text mining.
-However, for some situations LaMachine was not a good fit for the tasks at hand: For team members who did not have
-experience with a non-Windows environment, LaMachine was not a suitable or useful tool. And also for members who wanted
-to use a graphical user interface for their text editors as we did not manage to get the x-forwarding working (note that
-this also depends on the overall VM). After a few attempts the team gave up on resolving this issue due to time pressure
-but it also demonstrated that fine-tuning the configuration of LaMachine on certain aspects is beyond the reach of a data scientist without system administration skills.
- %iris it moet nog wat beter geformuleerd worden maar je snaprt wat ik bedoel
+LaMachine offered a convenient platform for a range of different explorations and experiments in the area of NLP and
+text mining.  However, for some situations LaMachine, or rather Linux in general, was not a good fit for the audience of
+the workshop: For team members who did not have experience with a non-Windows environment, LaMachine was not a suitable
+or useful tool. The limit of LaMachine was also reached for members who wanted to use desktop text editors with a
+graphical user interface as this is not offered by LaMachine and we did not manage to get X-forwarding working in the
+Ubuntu Linux VM. After a few attempts the team gave up on resolving this issue due to time pressure. This, also
+demonstrates that fine-tuning the configuration of LaMachine on certain aspects is beyond the reach of a data scientist
+without system administration skills. But this certainly also applies also to the installation as a whole in the
+SURFsara context, which involved things like the partitioning, formatting and mounting of (virtual) drives and setting
+up user accounts on the shared VM, all of which require some system administration skills.
+ %verwerkt <-- <iris> it moet nog wat beter geformuleerd worden maar je snaprt wat ik bedoel
+%@iris: hebben jullie verder geen gebruik proberen te maken van de Jupyter Lab omgeving? Dat biedt in ieder geval iets
+%grafisch wat via web te benaderen is?
 

### 2019-01-29 docs/lamachine_clarin2018.tex
fixed the broken sentence, added note to case study 2 and extended the acknowlegdements.

@@
     \item The source
-        code must be hosted in a public version controlled repository. \footnote{e.g. Github, Gitlab, Bitbucket, provided
+        code must be hosted in a public version controlled repository.\footnote{e.g. Github, Gitlab, Bitbucket, provided
         the respository is public}
@@
 LaMachine exists since May 2015 and has been used extensively ever since by numerous users. In early 2018 version 2 was
-released which was a significant redesign, powered by Ansible.
+released which was a significant redesign, powered by Ansible\footnote{\url{https://www.ansible.com}}.
 
@@
 academia. The industrial partner provides a problem and a team of researchers from different backgrounds and
-universities collaborate to come up with solutions. We participated in the 2019 edition on the case study by the Dutch
-Royal Library who %@Iris: zin niet afgemaakt?
+universities collaborate to come up with solutions.
+We participated in the 2019 edition on the case study by the Dutch Royal Library who wanted to investigate automatic methods for cataloguing of textual cultural heritage objects, in this particular case a large collection of digital dissertations.
 
@@
 up user accounts on the shared VM, all of which require some system administration skills.
+Lamachine was convenient and speeded up writing code as the most common scientific data-related packages are already present in Lamachine.
+
  %verwerkt <-- <iris> it moet nog wat beter geformuleerd worden maar je snaprt wat ik bedoel
@@
 %grafisch wat via web te benaderen is?
+%@maarten helaas lukte dat ook niet, dat liep bij mij steeds vast. - ik heb opmerking van alex over tijdswinst toegevoefd
 
@@
 LaMachine provides a flexible solution for the installation of a variety of NLP software, resulting in a kind
-of virtual laboratory that, through various interfaces, can be employed by a variety of people, from developers and data
-scientists to the wider research community that CLARIN explicitly addresses.
+of virtual laboratory that, through various interfaces, can be employed by a variety of people, from developers and data scientists to the wider research community that CLARIN explicitly addresses.
 
@@
 This research was funded by NWO CLARIN-NL, CLARIAH and the ZonMw project {\it Tekstmining in het toezicht: een exploratieve studie naar de automatische verwerking van klachten ingediend bij het Landelijk Meldpunt Zorg}, project number 516004614.
-We thank all project partners: the Dutch Health Inspectorate, IQhealthcare, and Tim Voets for their valuable contributions and help in the ZonMw project.
+We thank the Dutch Health Inspectorate, IQhealthcare, and Tim Voets for their valuable contributions and help in the ZonMw project, and NWO, the ICT Research Platform Nederland, the Lorentz Centre, the Royal Library and all our team members for making the ICT with Industry workshop a success.
 

### 2019-01-29 docs/lamachine_clarin2018.tex
some rephrasing

--- 2019-01-28 docs/lamachine_clarin2018.tex
@@
 or useful tool. The limit of LaMachine was also reached for members who wanted to use desktop text editors with a
-graphical user interface as this is not offered by LaMachine and we did not manage to get X-forwarding working in the
-Ubuntu Linux VM. After a few attempts the team gave up on resolving this issue due to time pressure. This, also
-demonstrates that fine-tuning the configuration of LaMachine on certain aspects is beyond the reach of a data scientist
-without system administration skills. But this certainly also applies also to the installation as a whole in the
+graphical user interface as this is not offered by LaMachine. Moreover, we did not manage to get X-forwarding working in the
+Ubuntu Linux VM and after a few attempts the team gave up on resolving this issue due to time pressure. This, also
+demonstrates that fine-tuning the configuration of certain aspects of LaMachine, but especially beyond LaMachine, is beyond the reach of a data scientist
+without system administration skills. This certainly also applies also to the installation as a whole in the
 SURFsara context, which involved things like the partitioning, formatting and mounting of (virtual) drives and setting
-up user accounts on the shared VM, all of which require some system administration skills.
+up user accounts on the shared VM, all of which require some system administration skills and are too context-specific
+to be within the scope of LaMachine.
  %verwerkt <-- <iris> it moet nog wat beter geformuleerd worden maar je snaprt wat ik bedoel

### 2019-01-29 docs/lamachine_clarin2018.tex

@@
 to be within the scope of LaMachine.
+Lamachine was convenient and speeded up writing code as the most common scientific data-related packages are already present in Lamachine.
  %verwerkt <-- <iris> it moet nog wat beter geformuleerd worden maar je snaprt wat ik bedoel

### 2019-01-29 docs/lamachine_clarin2018.tex
[paper] fix casing

@@
 community.
-We explain what Lamachine is, how it can be used and the technical details.
-We also compare Lamachine to alternative software distributions and discuss its advantages and limitations.
-We illustrate how LaMachine can be used in two case studies, one in an exploratory text mining project at the Dutch Health Inspectorate where LaMachine was applied to create a research environment for automatic text analysis for health care quality monitoring, and a second case where Lamachine was used to create a workspace for a one-week, intense collaboration by a diverse research team.
+We explain what LaMachine is, how it can be used and the technical details.
+We also compare LaMachine to alternative software distributions and discuss its advantages and limitations.
+We illustrate how LaMachine can be used in two case studies, one in an exploratory text mining project at the Dutch Health Inspectorate where LaMachine was applied to create a research environment for automatic text analysis for health care quality monitoring, and a second case where LaMachine was used to create a workspace for a one-week, intense collaboration by a diverse research team.
 \end{abstract}
@@
 
- This paper provides a detailed description of Lamachine, its purpose and functionality and the technical details. We
+ This paper provides a detailed description of LaMachine, its purpose and functionality and the technical details. We
  discuss the advantages and limitations of LaMachine compared to alternative software distributions in Section \ref{sec:comparison}.
@@
 
- \section{What is Lamachine?}
+ \section{What is LaMachine?}
 
@@
 versions of the various software packages inside LaMachine. It should not be
-confused with the actual \emph{version release number} of Lamachine as a whole, which is the version number assigned to
+confused with the actual \emph{version release number} of LaMachine as a whole, which is the version number assigned to
 the collective of installation scripts LaMachine provides, and which marks LaMachine releases.
@@
 to be within the scope of LaMachine.
-Lamachine was convenient and speeded up writing code as the most common scientific data-related packages are already present in Lamachine.
+LaMachine was convenient and speeded up writing code as the most common scientific data-related packages are already present in LaMachine.
  %verwerkt <-- <iris> it moet nog wat beter geformuleerd worden maar je snaprt wat ik bedoel

### 2019-01-30 bootstrap.sh
making a start on implement LXD/LXC support #134

@@ -20,7 +20,7 @@ boldgreen=${bold}$(tput setaf 2) #  green
 boldblue=${bold}$(tput setaf 4) #  blue
 normal=$(tput sgr0)
 
-export LM_VERSION="v2.4.9" #NOTE FOR DEVELOPER: also change version number in codemeta.json *AND* roles/lamachine-core/defaults/main.yml -> lamachine_version!
+export LM_VERSION="v2.4.10" #NOTE FOR DEVELOPER: also change version number in codemeta.json *AND* roles/lamachine-core/defaults/main.yml -> lamachine_version!
 echo "${bold}=====================================================================${normal}"
 echo "           ,              ${bold}LaMachine $LM_VERSION${normal} - NLP Software distribution"
 echo "          ~)                     (http://proycon.github.io/LaMachine)"
@@ -40,6 +40,8 @@ usage () {
     echo "       (uses Docker and Ansible)"
     echo "  singularity = in a Singularity container"
     echo "       (uses Singularity and Ansible)"
+    echo "  lxc = in an LXC container"
+    echo "       (uses LXD, LXC and Ansible)"
     echo "  local = in a local user environment"
     echo "       installs as much as possible in a separate directory"
     echo "       for a particular user, can exists alongside existing"
@@ -412,6 +414,9 @@ if [ -z "$FLAVOUR" ]; then
         if [ $WINDOWS -eq 0 ]; then
         echo "  6) in a Singularity container"
         echo "       (uses Singularity and Ansible)"
+        echo "  7) in an LXC container"
+        echo "       Provides a more persistent and VM-like container experience than Docker"
+        echo "       (uses LXD, LXC and Ansible)"
         fi
         echo -n "${bold}Your choice?${normal} [12345] "
         read choice
@@ -422,6 +427,7 @@ if [ -z "$FLAVOUR" ]; then
             [4]* ) FLAVOUR="global";  break;;
             [5]* ) FLAVOUR="remote"; break;;
             [6]* ) FLAVOUR="singularity"; break;;
+            [7]* ) FLAVOUR="lxc"; break;;
             * ) echo "Please answer with the corresponding number of your preference..";;
         esac
     done
@@ -597,8 +603,8 @@ else
     if ! which git; then
         NEED+=("git")
[…]

### 2019-02-01 docs/lamachine_clarin2018.tex
[paper] removed singularity bullet point and merged it into a remark, added LXD

@@
         environment from the host system, only the Linux kernel is shared. It is a lighter option than full
-        virtualisation. \footnote{Note that Docker on other platforms such as Windows and macOS does do use full
-        virtualisation.}
-    \item \textbf{Singularity container} -- Another form of containerisation with more focus on security\footnote{reduced
-        privilege escalation}, making it more suitable than Docker for deployment on computing clusters. This flavour is still under development.
+        virtualisation\footnote{Note that Docker on other platforms such as Windows and macOS does do use full
+        virtualisation.}.  Support for two other forms of containerisation, i.e. Singularity\footnote{\url{https://www.sylabs.io/singularity/}} and
+            LXC/LXD\footnote{\url{https://linuxcontainers.org}}, are currently under development.
     \item \textbf{Virtual Machine} --- Installs LaMachine as a virtual machine, i.e. through full virtualisation. This allows
@@
 
-\section{Comparison to alternative distributions}
-\label{sec:comparisons}
+\section{Comparison to alternative distributions}\label{sec:comparison}
 

### 2019-02-02 docs/lamachine_clarin2018.tex
last paper fixes

@@
 multiple components, the individual installation (sometimes involving compilation from source) and configuration of
-which would be not be trivial. Moreover, LaMachine shines when the user wants higher-level interfaces on top of that,
+which would not be trivial. Moreover, LaMachine shines when the user wants higher-level interfaces on top of that,
 something she can use right out of the box. It also fulfills a role in case the user simply does not know what she wants
@@
 
-Comparisons could be drawn between LaMachine as a kind of Linux distribution and other Linux distributions. But that
-would not be appropriate as we characterise LaMachine more as a meta-distribution which simply builds on the foundation
+Comparisons could be drawn between LaMachine as a kind of Linux distribution and other Linux distributions. That,
+however, would not be appropriate as we characterise LaMachine more as a meta-distribution which simply builds on the foundation
 of several existing major Linux distributions. As a meta-distribution that runs in a variety of flavours and on a
@@
 a focus on Python and R. It is much wider in scope than LaMachine, which has a strong NLP and CLARIN/CLARIAH focus. When
-it comes to Python, which Anaconda is initially geared at, there is considerable overlap in the modules that LaMachine and Anaconda offer. Unlike LaMachine,
+it comes to Python, which Anaconda was initially aimed at, there is considerable overlap in the modules that LaMachine and Anaconda offer. Unlike LaMachine,
 Anaconda does not focus on providing different flavours such as a VM or a Docker Container\footnote{This is not a
-technical limitation but simply a matter of different objectives, one could easily create a VM or Docker Container and
+technical limitation but simply a matter of different objectives, one could easily create a VM or a container and
 install Anaconda in it.}, nor separate stable and development versions of the included software.
@@
 that builds a single kind of container, LaMachine offers a high degree of flexibility for the construction of different
-containers, i.e. containing different software of having different configurations, based on the user's needs. It is
+containers, i.e. containing different software or having different configurations, based on the user's needs. It is
 quite feasible to instantiate a variety of LaMachine containers and use a container orchestration system such
@@
 
-\section*{Acknowledgement}
+\section*{Acknowledgements}
 

### 2019-03-21 docs/lamachine_clarin2018.tex
[paper] rephrased 2nd sentence of introduction (was ungrammatical/not understandable)
[paper] added a paragraph to clarify that LaMachine is in widespread use and has been installed by numerous users (rather than just by the author as one reviewer thought)

@@
 Software is a key deliverable and a vital component for research in projects such as those under the CLARIN umbrella.
-Software provides researchers the instruments to yield for their research.
-It is CLARIN's core mission to make
+Software can be seen as an invaluable instrument that enable researchers to perform their studies. It is CLARIN's core mission to make
 digital language resources, including software, available to the wider research community.
@@
 
+LaMachine exists since since May 2015 and has been installed extensively ever since by numerous users. In early 2018 version 2 was
+released which was a significant redesign, powered by Ansible\footnote{\url{https://www.ansible.com}}. We ourselves use
+LaMachine as a means by to make our full software stack available to end-users, whether those be individual researchers, research
+groups, or hosting providers like CLARIN centres. LaMachine has in fact been adopted in all of those areas. Quantifying its
+actual use is not trivial despite having some tracking solutions in place, but can roughly be put at around one or two
+installations or upgrades \emph{every day}. We also see a regular influx of questions on our issue tracker, indicating
+community interest.
+
 The software included in LaMachine has to adhere to the following prerequisites:
@@
 \section{ Included Software}
-
-LaMachine exists since May 2015 and has been used extensively ever since by numerous users. In early 2018 version 2 was
-released which was a significant redesign, powered by Ansible\footnote{\url{https://www.ansible.com}}.
 

### 2019-03-21 docs/lamachine_clarin2018.tex
[paper] processed mostly language remarks by reviewer 1

@@
 hand, expose certain functionality of the software in an easy and accessible way, but in doing so often can not expose
-the full power of the software. The cost trade-off is also apparant in the construction of the interfaces, where
+the full power of the software. The cost trade-off is also apparent in the construction of the interfaces, where
 high-level interfaces are typically far more costly to build.
@@
 Central\footnote{\url{https://search.maven.org}} for Java.  LaMachine relies on those to obtain and install software. In doing so, we compel participating software projects
-to adhere to well-established distribution standards and ensure the software is more sustainable towards the future
+to adhere to well-established distribution standards and ensure the software is more sustainable
 \cite{softwarequality}. Moreover, we ensure that LaMachine never becomes a prerequisite for the software but merely a
@@
         code must be hosted in a public version controlled repository.\footnote{e.g. Github, Gitlab, Bitbucket, provided
-        the respository is public}
+        the repository is public.}
     \item There must be a build process to compile the source, if applicable, and install the program or library.
@@
  \begin{itemize}
-     \item \textbf{KafNafParserPy}\footnote{\url{https://github.com/cltl/KafNafParserPy}} - A python module to parse NAF files.
+     \item \textbf{KafNafParserPy}\footnote{\url{https://github.com/cltl/KafNafParserPy}} - A python module to parse NAF
+         files, another format for linguistic annotation.
+     \item \textbf{NafFoLiApy}\footnote{\url{https://github.com/cltl/NafFoLiApy}} - Converters between FoLiA and NAF.
  \end{itemize}
@@
 source as possible. This can be contrasted with other metadata efforts, also within CLARIN, to establish a more
-centralised metadata store\footnote{such as CLARIN's CMDI Component Registry;
+centralised metadata store\footnote{Such as CLARIN's CMDI Component Registry;
 \url{https://www.clarin.eu/content/component-metadata}} with more specialised and

### 2019-03-21 docs/clarin_reviewer_response.md
[paper] added reviewer response

## REVIEW 1

> The second point is that the impression is that the only person that would in fact be able to (or, at least, has done it so far) install
> LaMachine is the author.

On the contrary! The whole idea behind LaMachine is to facilitate installation by other parties, and there are many
users who have installed and are using LaMachine. If only we as developers succeeded in installation then that would indeed
be a major failure on our part. I added a paragraph to prevent confusion in this regard.

> Also, and at least in further work: currenlty, the already ingested software seems to cater (almost) only for Dutch.

A lot of the software is generically applicable actually. It is only CLARIN-NL/CLARIAH software like Frog, Alpino, Valkuil
and Oersetter that is very strongly focussed on Dutch. But even in those situations the underlying software often allows
for multiple languages if somebody trains models for it.

> I can see the objection that everyone can install whetever they wish,

That not an objection from our side, that's in fact deliberately what we aim for  ;)

## REVIEW 2

> The justification for such a system is well argued for but how such facility relates to other functions that CLARIN centres are
> supposed to serve is not entirely clear.

## REVIEW 4

This reviewer poses some very good and interesting questions about the relation between LaMachine and the larger CLARIN infrastructure:

> Is LaMachine actively distributed and maintained at a CLARIN centre or by CLARIN ERIC centrally?

It's distributed by the Centre of Language and Speech Technology, who are in the process of becoming a certified CLARIN
centre but are not yet one at this moment, no.

> Is there a global installation of LaMachine run and maintained by CLARIN and to which all CLARIN users can get access?

There is our installation at https://webservices-lst.science.ru.nl .

> If so, is this installation accessible through login with the CLARIN ID federation?

Not yet, we keep getting contradicting signals whether this is desired and are running into the issue that CLARIN(-NL) (to my
limited understainding) does not yet seem to have the relevant OAuth infrastructure in place that allows single-sign-on
also with RESTful webservices (the delegation problem).

> The paper states "Support for macOS is limited because not all participating software supports it." Could it be a solution to containerize such
> software?

One of LaMachine's flavours provides a container solution (n.b: a single container for all selected software). But on
macOS containerisation often implies virtualisation, which in fact is also already provided by LaMachine as another
flavour. That indeed offers a viable way to use software in LaMachine that is not otherwise possible on macOS (or
Windows for that matter). The statement about limited support, however, applied to compiling software for macOS
natively. I added a small footnote to hopefully clarify this.

> Is LaMachine catalogued in the VLO, with a PID to the metadata?

No, nobody has ever requested this desire yet. I'll look into it. It may be a bit difficult as LaMachine is such a
meta-distribution that it is hard to pin down for any single scenario.

> Does LaMachine aim at compatibility with the CLARIN Language Resource Switchboard? The paper states that LaMachine "transcends" the
> ambitions of the Switchboard, but more relevant is the question whether the two are compatible and can be integrated.

That remark in our paper applies to the VRE project rather than to LaMachine, I clarified it to prevent confusion. LaMachine is not directly comparable with the
switchboard as it offers no switchboard functionality. Now, the switchboard itself possibly be included in the LaMachine
distribution. I briefly considered this but decided to include a simpler portal instead as the switchboard did not meet
all our demands and other initiatives were started to compete with it (CLARIAH WP3 VRE). Tighter compatibility between
the switchboard and services offered by a LaMachine installation would be possible, but this is more a question for the
switchboard developers than for us.

In fact, the current switchboard already points to various webservices that are hosted in our LaMachine installation at
Radboud University, though this involved manual duplication of metadata rather than leveraging the codemeta we offer.

### 2019-03-21 docs/lamachine_clarin2018.tex
[paper] added reviewer response

@@
 manager\footnote{\url{https://brew.sh}; alternatives such as macports are not supported}. Support for macOS is limited because not all participating
-software supports it. Certain Linux Distributions that are derivatives of the aforementioned distributions \emph{may} also work, such as RedHat
+software supports it natively\footnote{One can always still fall back to the virtualised flavour of LaMachine}.
+Certain Linux Distributions that are derivatives of the aforementioned distributions \emph{may} also work, such as RedHat
 Enterprise Linux (CentOS) and Linux Mint (Ubuntu).
@@
 Development of LaMachine presently takes place in collaboration with the CLARIAH WP3 Virtual Research Environment (VRE)
-project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}}, which has higher ambitions in
+project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}}. The VRE project has higher ambitions in
 accommodating the researcher and connectivity of data and services, and transcends also those of the CLARIN Language

### 2019-03-21 docs/clarin_reviewer_response.md
[paper] update

@@
 No, nobody has ever requested this desire yet. I'll look into it. It may be a bit difficult as LaMachine is such a
-meta-distribution that it is hard to pin down for any single scenario.
+meta-distribution that it is hard to pin down for any single scenario, but it would be interesting for the services
+inside a LaMachine installation.  I opened issue https://github.com/proycon/LaMachine/issues/138 to this end.
 

### 2019-03-23 bootstrap.sh
added spacy2folia

@@ -20,7 +20,7 @@ boldgreen=${bold}$(tput setaf 2) #  green
 boldblue=${bold}$(tput setaf 4) #  blue
 normal=$(tput sgr0)
 
-export LM_VERSION="v2.5.0" #NOTE FOR DEVELOPER: also change version number in codemeta.json *AND* roles/lamachine-core/defaults/main.yml -> lamachine_version!
+export LM_VERSION="v2.5.1" #NOTE FOR DEVELOPER: also change version number in codemeta.json *AND* roles/lamachine-core/defaults/main.yml -> lamachine_version!
 echo "${bold}=====================================================================${normal}"
 echo "           ,              ${bold}LaMachine $LM_VERSION${normal} - NLP Software distribution"
 echo "          ~)                     (http://proycon.github.io/LaMachine)"

### 2019-03-28 docs/lamachine_clarin2018.tex
added first comments to clarin paper

@@
 
-LaMachine exists since since May 2015 and has been installed extensively ever since by numerous users. In early 2018 version 2 was
+LaMachine exists since May 2015 and has been installed extensively ever since by numerous users.
+Our statistics show that at the minimum Lamachine was downloaded over a thousand times from over 25 different countries \footnote{https://applejack.science.ru.nl/lamastats/lamachinestats.html}.
+In early 2018 version 2 was
 released which was a significant redesign, powered by Ansible\footnote{\url{https://www.ansible.com}}. We ourselves use
@@
 
+Lamachine is most often used by a researcher or data scientist who installs Lamachine (locally, as a Docker or VM) on his own computer.
+Here we discuss two case studies where Lamachine was used under circumstances that go beyond this most common usage of Lamachine: using Lamachine on a secure server without internet connection, and using Lamachine to provide a work environment for a group of researchers.
+
 \subsection{Text Mining for Health Inspection}

### 2019-03-29 docs/lamachine_clarin2018.tex
[paper] reordered/rephrased some things

--- 2019-03-21 docs/lamachine_clarin2018.tex
@@
 
-LaMachine exists since since May 2015 and has been installed extensively ever since by numerous users. In early 2018 version 2 was
-released which was a significant redesign, powered by Ansible\footnote{\url{https://www.ansible.com}}. We ourselves use
-LaMachine as a means by to make our full software stack available to end-users, whether those be individual researchers, research
-groups, or hosting providers like CLARIN centres. LaMachine has in fact been adopted in all of those areas. Quantifying its
-actual use is not trivial despite having some tracking solutions in place, but can roughly be put at around one or two
-installations or upgrades \emph{every day}. We also see a regular influx of questions on our issue tracker, indicating
-community interest.
+LaMachine exists since May 2015 and has been installed extensively ever since by numerous users. Quantifying its actual
+use is not trivial despite having sme tracking solutions in place, but our statistics show that LaMachine was installed at least over a thousand times, on
+machines in over 25 different countries since its inception\footnote{https://applejack.science.ru.nl/lamastats/lamachinestats.html}.
+In early 2018 version 2 was released which was a significant redesign, powered
+by Ansible\footnote{\url{https://www.ansible.com}}. We ourselves use LaMachine as a means by to make our full software
+stack available to end-users, whether those be individual researchers, research groups, or hosting providers like CLARIN
+centres. LaMachine has in fact been adopted in all of those areas. We also see a regular influx of questions on our issue tracker, indicating community interest.
 
@@
 
+LaMachine is most often used by a researcher or data scientist who installs LaMachine on his own computer (locally, in a
+local environment or as a Docker container or VM). Here we discuss two case studies where LaMachine was used under
+circumstances that go beyond this most common usage of LaMachine: using LaMachine on a secure server without internet
+connection, and using LaMachine to provide a work environment for a group of researchers.
+
 \subsection{Text Mining for Health Inspection}

### 2019-03-29 docs/lamachine_clarin2018.tex
added  to clarin paper

@@
 LaMachine exists since May 2015 and has been installed extensively ever since by numerous users. Quantifying its actual
-use is not trivial despite having sme tracking solutions in place, but our statistics show that LaMachine was installed at least over a thousand times, on
+use is not trivial despite having some tracking solutions in place, but our statistics show that LaMachine was installed at least over a thousand times, on
 machines in over 25 different countries since its inception\footnote{https://applejack.science.ru.nl/lamastats/lamachinestats.html}.
@@
 
+The CLARIN Switchboard \cite{zinn2016clarin} also offers access to NLP tools at a different level as it provides a webinterface where users can upload their texts for NLP processing, proving very easy acces.
+Lamachine is most often used by data scientists and reseachers who do have programming experience and want to work with complex NLP software.
+
 Comparisons could be drawn between LaMachine as a kind of Linux distribution and other Linux distributions. That,
@@
 
+Virtual Research Environments (VRE) such as the Dutch CLARIAH WP3 VRE project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}} and the Spanish VRE project \cite{gonzalez2017evi} aim to develop more high level tool environments with graphical interfaces to accomodate the reseacher.
+Lamachine is currently developed in close collaboration with the Dutch CLARIAH WP3 VRE project.
+
 \section{Case studies}\label{sec:case}
@@
 circumstances that go beyond this most common usage of LaMachine: using LaMachine on a secure server without internet
-connection, and using LaMachine to provide a work environment for a group of researchers.
+connection, and using LaMachine to provide a work environment for a group of researchers. These case studies illustrate the advantages and limitations of Lamachine in practice.
 
@@
 VM with Windows Server 2012, provided to us by the health inspectorate for this project, in which we did have administrative
-rights but no internet access. In terms of hardware we ran on a machine with 8 cores and  32GB internal memory available.
+rights but no internet access.
+In terms of hardware we ran on a machine with 8 cores and  32GB internal memory available.
+Note that as we had no internet access available on the working server, we prepared a custom-made VM image of Lamachine that included an additional editor and some extra Python libraries.
 %using VirtualBox and Vagrnt on the Windows VM
@@
 text mining. However, for some situations LaMachine, or rather Linux in general, was not a good fit for the audience of
-the workshop: For team members who did not have experience with a non-Windows environment, LaMachine was not a suitable
+the workshop: for team members who did not have experience with a non-Windows environment, LaMachine was not a suitable
 or useful tool. The limit of LaMachine was also reached for members who wanted to use desktop text editors with a
@@
 
-Use cases as the examples in section \ref{sec:case} contribute to thorough testing and running of LaMachine in less ideal
-circumstances such as nested VM constructions and offline usage.
+Most users of Lamachine are installing the software distribution in their own machine, either locally, via Docker or VM.
+However, we described to use cases in section \ref{sec:case} where we explore the possibilities under more constrained and demanding circumstances.
+Such case studies contribute to thorough testing and running of LaMachine in less ideal circumstances such as nested VM constructions and offline usage.
 

### 2019-03-29 docs/lamachine_clarin2018.tex
further work on paper

@@
 
-The CLARIN Switchboard \cite{zinn2016clarin} also offers access to NLP tools at a different level as it provides a webinterface where users can upload their texts for NLP processing, proving very easy acces.
-Lamachine is most often used by data scientists and reseachers who do have programming experience and want to work with complex NLP software.
+Within the scope of CLARIN, one might be inclined to draw comparisons with the CLARIN Language Resource Switchboard
+\cite{zinn2016clarin}, which also offers access to NLP tools through a web interface where users can upload their texts
+for NLP processing, providing easy access and automated tool discovery.  This, however, is notably different from what
+LaMachine is. The CLARIN switchboard is not a software distribution, it links to externally hosted applications; and
+LaMachine does currently not provide any switchboard functionality, but merely offers a portal interface. Moreover,
+LaMachine is used by data scientists and reseachers who do have programming experience and want to work with complex NLP
+software, on their own machines and unencumbered by high-level interfaces if they so desire.
 
-Comparisons could be drawn between LaMachine as a kind of Linux distribution and other Linux distributions. That,
+There are also CLARIN initiatives that provide not just a switchboard, but a more elaborate high-level virtual research
+environment (VRE) with graphical interfaces to accommodate the researcher and allow execution and visualisation of
+various NLP pipeline. The Dutch CLARIAH WP3 project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}}
+is such an initiative. LaMachine is currently developed in
+close collaboration with the Dutch CLARIAH WP3 VRE project, in which we act as a technology supplier.
+
+Comparisons could be drawn between LaMachine as a kind of Linux distribution and other Linux distributions. That too,
 however, would not be appropriate as we characterise LaMachine more as a meta-distribution which simply builds on the foundation
@@
 
-Virtual Research Environments (VRE) such as the Dutch CLARIAH WP3 VRE project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}} and the Spanish VRE project \cite{gonzalez2017evi} aim to develop more high level tool environments with graphical interfaces to accomodate the reseacher.
-Lamachine is currently developed in close collaboration with the Dutch CLARIAH WP3 VRE project.
-
 \section{Case studies}\label{sec:case}
@@
 circumstances that go beyond this most common usage of LaMachine: using LaMachine on a secure server without internet
-connection, and using LaMachine to provide a work environment for a group of researchers. These case studies illustrate the advantages and limitations of Lamachine in practice.
+connection, and using LaMachine to provide a work environment for a group of researchers. These case studies illustrate
+the advantages and limitations of LaMachine in practice.
 
@@
 In terms of hardware we ran on a machine with 8 cores and  32GB internal memory available.
-Note that as we had no internet access available on the working server, we prepared a custom-made VM image of Lamachine that included an additional editor and some extra Python libraries.
+Note that as we had no internet access available on the working server, we prepared a custom-made VM image of LaMachine that included an additional editor and some extra Python libraries.
 %using VirtualBox and Vagrnt on the Windows VM
@@
 
-Most users of Lamachine are installing the software distribution in their own machine, either locally, via Docker or VM.
+Most users of LaMachine are installing the software distribution in their own machine, either locally, via Docker or VM.
 However, we described to use cases in section \ref{sec:case} where we explore the possibilities under more constrained and demanding circumstances.

### 2019-03-29 docs/lamachine_clarin2018.tex
minor edits
Various cropping of images, shortening and merging phrases, to get everything within the page limit whilst still offering better readability in the images

@@
 %\setlength\titlebox{5cm}
+\setlist{nosep}
 \setlist{noitemsep}
+\setlength{\belowcaptionskip}{-10pt}
+\setlength{\abovecaptionskip}{0pt}
 
@@
 
-\section{ Included Software}
+\section{Included Software}
 
@@
  \item by the Language Machines Research Group and the Centre of Language and Speech Technology, Radboud University,
-     Nijmegen\footnote{Links: \url{https://languagemachines.github.io/timbl}, \url{https://languagemachines.github.io/mbt},
+     Nijmegen:\footnote{Links: \url{https://languagemachines.github.io/timbl}, \url{https://languagemachines.github.io/mbt},
     \url{https://languagemachines.github.io/ucto}, \url{https://languagemachines.github.io/frog},
@@
  \end{itemize}
- \item by the University of Groningen
- \begin{itemize}
-     \item \textbf{Alpino}\footnote{\url{http://www.let.rug.nl/vannoord/alp/Alpino/}} - A dependency parser and tagger for Dutch.
- \end{itemize}
- \item by the Vrije Universiteit Amsterdam
- \begin{itemize}
-     \item \textbf{KafNafParserPy}\footnote{\url{https://github.com/cltl/KafNafParserPy}} - A python module to parse NAF
-         files, another format for linguistic annotation.
-     \item \textbf{NafFoLiApy}\footnote{\url{https://github.com/cltl/NafFoLiApy}} - Converters between FoLiA and NAF.
- \end{itemize}
- \item by Utrecht University
- \begin{itemize}
-     \item  \textbf{T-scan}\footnote{\url{https://github.com/proycon/tscan}} - T-scan is a Dutch text analytics tool for readability prediction (initially developed at TiCC, Tilburg University, and in collaboration with Radboud University, Nijmegen)
- \end{itemize}
- \item by the Meertens Instituut
- \begin{itemize}
-     \item \textbf{Python Course for the Humanities}\footnote{\url{http://www.karsdorp.io/python-course/}} - Interactive
+ \item by the University of Groningen: \textbf{Alpino}\footnote{\url{http://www.let.rug.nl/vannoord/alp/Alpino/}} -- A dependency parser and tagger for Dutch.
+ \item by the Vrije Universiteit Amsterdam:
+     \textbf{KafNafParserPy}\footnote{\url{https://github.com/cltl/KafNafParserPy}} -- a python module to parse NAF
+         files, another format for linguistic annotation; and
+         \textbf{NafFoLiApy}\footnote{\url{https://github.com/cltl/NafFoLiApy}} -- converters between FoLiA and NAF.
+ \item by Utrecht University: \textbf{T-scan}\footnote{\url{https://github.com/proycon/tscan}} -- a Dutch text analytics tool for readability prediction (initially developed at TiCC, Tilburg University, and in collaboration with Radboud University, Nijmegen)
+ \item by the Meertens Instituut: \textbf{Python Course for the
+     Humanities}\footnote{\url{http://www.karsdorp.io/python-course/}} -- interactive
          tutorial and introduction into programming with Python for the humanities.
- \end{itemize}
 \end{itemize}
@@
     \end{itemize}
-    \item \textbf{R}
-    \item \textbf{Java}
+    \item \textbf{R}, \textbf{Perl}, \textbf{Java}:
     \begin{itemize}
@@
 \begin{figure}[htb] \begin{center}
-\includegraphics[width=110.0mm]{screenshot_bootstrap.jpg}
+\includegraphics[width=120.0mm]{screenshot_bootstrap.jpg}
 \end{center}
@@
 \begin{figure}[htb] \begin{center}
-\includegraphics[width=135.0mm]{screenshot_venv_activate.jpg}
+\includegraphics[width=120.0mm]{screenshot_venv_activate.jpg}
 \end{center}
@@
 \begin{figure}[htb] \begin{center}
-\includegraphics[width=135.0mm]{screenshot_portal.jpg}
+\includegraphics[width=150.0mm]{screenshot_portal.jpg}
 \end{center}
@@
 \begin{figure}[htb] \begin{center}
-\includegraphics[width=135.0mm]{screenshot_lab.jpg}
+\includegraphics[width=150.0mm]{screenshot_lab.jpg}
 \end{center}
@@
 various NLP pipeline. The Dutch CLARIAH WP3 project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}}
-is such an initiative. LaMachine is currently developed in
-close collaboration with the Dutch CLARIAH WP3 VRE project, in which we act as a technology supplier.
+is such an initiative and LaMachine is currently developed in close collaboration with this project, in a role of technology supplier.
 
@@
 
-LaMachine is most often used by a researcher or data scientist who installs LaMachine on his own computer (locally, in a
-local environment or as a Docker container or VM). Here we discuss two case studies where LaMachine was used under
+LaMachine is most often used by a researcher or data scientist who installs LaMachine on his own computer (in a
+local environment, Docker container or VM). Here we discuss two case studies where LaMachine was used under
 circumstances that go beyond this most common usage of LaMachine: using LaMachine on a secure server without internet
@@
 Most users of LaMachine are installing the software distribution in their own machine, either locally, via Docker or VM.
-However, we described to use cases in section \ref{sec:case} where we explore the possibilities under more constrained and demanding circumstances.
-Such case studies contribute to thorough testing and running of LaMachine in less ideal circumstances such as nested VM constructions and offline usage.
+However, we described two case studies in section~\ref{sec:case} where we explore the possibilities under more
+constrained, complex, and demanding circumstances, which contributes to thorough testing and running of LaMachine.
 
@@
 
-Development of LaMachine presently takes place in collaboration with the CLARIAH WP3 Virtual Research Environment (VRE)
-project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}}. The VRE project has higher ambitions in
-accommodating the researcher and connectivity of data and services, and transcends also those of the CLARIN Language
-Resource Switchboard~\cite{switchboard}. An important part of our future focus will therefore be on interoperability
-with the higher-level tools emerging from the VRE efforts, but also with other parts of the CLARIN infrastructure;
-single-sign on authentication being a notable example here.
+An important part of our future focus will be on interoperability with the higher-level tools emerging from
+the CLARIAH WP3 VRE effort, but also with other parts of the CLARIN infrastructure; single-sign on authentication being
+a notable example here.
 

### 2019-03-29 docs/clarin_reviewer_response.md
updated paper and reviewer response

@@
 
+> The screenshots are in general too small to read, pls. make them larger!
+
+Noted, we enlarged them (and with some creative cropping) as far as possible now without getting into trouble with the
+page limit. It will still be on the small side in print, but hopefully the parts that should be readible are now
+readible enough.
+
+> Figure 5 would be a lot more interesting/convincing, if you gave an example from your actualy projects, rather than 101 for programming.
+
+We opted for this image to implicitly convey that LaMachine also has its uses in education. Showing actual development projects
+would not be very meaningful to the uninformed reader.
+
 ## REVIEW 2
@@
 
-This reviewer poses some very good and interesting questions about the relation between LaMachine and the larger CLARIN infrastructure:
+This reviewer poses some very good and interesting questions about the relation between LaMachine and the larger CLARIN infrastructure.
+It's hard to sufficiently address this in the paper in the limited space available, as we focus mostly on providing an
+clear description on what LaMachine is and contains.
 
@@
 
-It's distributed by the Centre of Language and Speech Technology, who are in the process of becoming a certified CLARIN
-centre but are not yet one at this moment, no.
+It's maintained and released by us, the Centre of Language and Speech Technology, distributed through various channels (github, docker hub, vagrant
+cloud). We are in the process of becoming a certified CLARIN centre but I omitted that in the paper because I'm not sure
+of our current CLARIN centre status and things are quite in state of flux so will be outdated quickly.
 
@@
 
-There is our installation at https://webservices-lst.science.ru.nl .
+Yes, good point. There is our installation at https://webservices-lst.science.ru.nl , which exposes at least various
+higher-level interfaces (low-level shell access and Jupyter Lab access would be a security risk of course). I added a
+mention of this in the paper because it was indeed missing.
 
@@
 limited understainding) does not yet seem to have the relevant OAuth infrastructure in place that allows single-sign-on
-also with RESTful webservices (the delegation problem).
+also with RESTful webservices (the delegation problem). We do have a remark in the paper that this is a point of future
+work.
 
@@
 meta-distribution that it is hard to pin down for any single scenario, but it would be interesting for the services
-inside a LaMachine installation.  I opened issue https://github.com/proycon/LaMachine/issues/138 to this end.
+inside a LaMachine installation.  I opened issue https://github.com/proycon/LaMachine/issues/138 to this end. It's not
+in the paper as it does not offer sufficient space for such a discussion.
 
@@
 
-That remark in our paper applies to the VRE project rather than to LaMachine, I clarified it to prevent confusion. LaMachine is not directly comparable with the
-switchboard as it offers no switchboard functionality. Now, the switchboard itself possibly be included in the LaMachine
-distribution. I briefly considered this but decided to include a simpler portal instead as the switchboard did not meet
-all our demands and other initiatives were started to compete with it (CLARIAH WP3 VRE). Tighter compatibility between
-the switchboard and services offered by a LaMachine installation would be possible, but this is more a question for the
-switchboard developers than for us.
+The "transcends" remark in our paper applies to the VRE project rather than to LaMachine, I clarified it to prevent
+confusion. LaMachine is not directly comparable with the switchboard as it offers no switchboard functionality. Now, the
+switchboard itself could possibly be included in the LaMachine distribution. I briefly considered this but decided to
+include a simpler portal instead as the switchboard did not meet all our demands and other initiatives were started to
+compete with it (CLARIAH WP3 VRE). Tighter compatibility between the switchboard and services offered by a LaMachine
+installation would be possible, but this is more a question for the switchboard developers than for us.
 
@@
 Radboud University, though this involved manual duplication of metadata rather than leveraging the codemeta we offer.
+
+> Does LaMachine support and expose CMDI metadata for the data it processes? It seems that only CodeMeta is supported, which I do not
+> believe to be a CLARIN standard. Why not CMDI?
+
+LaMachine itself is a distribution so it doesn't process any data as such; that would be a question for particular tools
+or frameworks and would not be something we as a distribution would enforce either.
+
+As to CodeMeta for software metadata, we tried to clarify our choice for that in the section titled 'Metadata'. We do hope we can come
+to a synthesis with CMDI software descriptions and there are actually some ongoing discussions to that end.

### 2019-03-29 docs/lamachine_clarin2018.tex
updated paper and reviewer response

@@
 Figure~\ref{fig:portal}, leverages the metadata registry compiled in each LaMachine installation (see
-section~\ref{sec:metadata}).
+section~\ref{sec:metadata}). A live example of the portal to our own LaMachine installation on our production
+server is publicly available\footnote{\url{https://webservices-lst.science.ru.nl}, note that we expose only certain
+high-level interfaces so this is just demonstrates one facet of LaMachine rather than the full LaMachine experience}.
 

### 2019-03-29 docs/lamachine_clarin2018.tex
[paper] rephrased switchboard comparison and other parts to make clearer that LaMAchine is a distribution

@@
 Software is a key deliverable and a vital component for research in projects such as those under the CLARIN umbrella.
-Software can be seen as an invaluable instrument that enable researchers to perform their studies. It is CLARIN's core mission to make
+Software can be seen as an invaluable instrument that enables researchers to perform their studies. It is CLARIN's core mission to make
 digital language resources, including software, available to the wider research community.
@@
 suite of interconnected software that is often non-trivial to install, configure, and deploy. This is where LaMachine
-comes in.
+comes in; as a \emph{distribution of software}.
 
@@
 
-LaMachine incorporates software providing different types of interfaces which, as seen above, typically address different audiences. Whilst we
+LaMachine distributes software providing different types of interfaces which, as seen above, typically address different audiences. Whilst we
 attempt to accommodate both technical\footnote{Data scientists, DevOps, system administrators, developers.} and
 less-technical audiences\footnote{The wider researcher community, particularly the Humanities; also educational
-settings.}, there is a natural bias towards the former as lower-level interfaces are often a prerequisite to build
+settings.}. There is a natural bias towards the former as lower-level interfaces are often a prerequisite to build
 higher-level interfaces on. Depending on the \emph{flavour} (more on this later) of LaMachine chosen, it makes a good virtual research
 environment for a data scientist, whether on a personal computer or on a computing cluster, a good development
-environment for a developer or a good deployment method for production servers in for example CLARIN centres.
+environment for a developer or a good deployment method to deliver to production servers such as CLARIN centres.
 
@@
 LaMachine exists since May 2015 and has been installed extensively ever since by numerous users. Quantifying its actual
-use is not trivial despite having some tracking solutions in place, but our statistics show that LaMachine was installed at least over a thousand times, on
-machines in over 25 different countries since its inception\footnote{https://applejack.science.ru.nl/lamastats/lamachinestats.html}.
+use is not trivial despite having some tracking solutions in place; our statistics show that LaMachine was installed at least over a thousand times, on
+machines in over 25 different countries, since its inception\footnote{https://applejack.science.ru.nl/lamastats/lamachinestats.html}.
 In early 2018 version 2 was released which was a significant redesign, powered
@@
 installation manifest lists all packages that will be installed, at the user's discretion. After the initial
-installation, the user can always add more packages using \texttt{lamachine-add} or editing the installation manifest directly.
+installation, the user can always add more packages using \texttt{lamachine-add} or by editing the installation manifest directly.
 
@@
 such as git and interpreters such as Python or R (or compilers for that matter). This allow users to use
-one of the many specialised programming libraries included in LaMachine to build their own tools and . All this makes
+one of the many specialised programming libraries included in LaMachine to build their own tools. All this makes
 LaMachine ideally suited as a development environment.
@@
 
-LaMachine provides various CLAM-based webservices. CLAM is a long-time CLARIN project that allows developers to turn
+LaMachine includes various CLAM-based webservices. CLAM is a long-time CLARIN project that allows developers to turn
 their command-line NLP tool into a RESTful webservice with relatively little effort \cite{CLAMPAPER}, and which in
@@
 Within the scope of CLARIN, one might be inclined to draw comparisons with the CLARIN Language Resource Switchboard
-\cite{zinn2016clarin}, which also offers access to NLP tools through a web interface where users can upload their texts
-for NLP processing, providing easy access and automated tool discovery.  This, however, is notably different from what
-LaMachine is. The CLARIN switchboard is not a software distribution, it links to externally hosted applications; and
-LaMachine does currently not provide any switchboard functionality, but merely offers a portal interface. Moreover,
-LaMachine is used by data scientists and reseachers who do have programming experience and want to work with complex NLP
-software, on their own machines and unencumbered by high-level interfaces if they so desire.
+\cite{zinn2016clarin}, which offers access to NLP tools through a web interface where users can upload their texts for
+NLP processing, providing easy access and automated tool discovery. This, however, is notably different from what
+LaMachine is. LaMachine is a software distribution; the CLARIN switchboard is not. Vice versa, LaMachine does not
+currently provide any switchboard functionality, it merely offers a portal interface. The switchboard allows access to
+tools hosted elsewhere, LaMachine installs and configures tools and only provides access to those it installed. Unlike
+the switchboard, the audience for LaMachine is in the first place the data scientist and researcher who has
+programming experience and wants to work with complex NLP software, on her own machine and unencumbered by high-level
+interfaces if she so desires.
 
-There are also CLARIN initiatives that provide not just a switchboard, but a more elaborate high-level virtual research
+There are also other CLARIN initiatives that provide not just a switchboard, but aim at an eve more ambitious high-level virtual research
 environment (VRE) with graphical interfaces to accommodate the researcher and allow execution and visualisation of
 various NLP pipeline. The Dutch CLARIAH WP3 project\footnote{\url{https://github.com/meertensinstituut/clariah-wp3-vre}}
-is such an initiative and LaMachine is currently developed in close collaboration with this project, in a role of technology supplier.
+is such an initiative: LaMachine is currently developed in close collaboration with this project, in a role of technology supplier.
 

### 2019-03-29 docs/clarin_reviewer_response.md
[paper] rephrased switchboard comparison and other parts to make clearer that LaMAchine is a distribution
updated reviewer response

@@
 
+(see also the comments to reviewer 4 below, who I think has a similar question)
+
 ## REVIEW 4
 
-This reviewer poses some very good and interesting questions about the relation between LaMachine and the larger CLARIN infrastructure.
-It's hard to sufficiently address this in the paper in the limited space available, as we focus mostly on providing an
-clear description on what LaMachine is and contains.
+This reviewer poses some very good and interesting questions about the relation between LaMachine and the larger CLARIN
+infrastructure.  It's hard to sufficiently address this in the paper in the limited space available, as we focus mostly
+on providing an clear description on what LaMachine is (a distribution) and contains. We have now tried to emphasise
+more the fact that LaMachine is a distribution, and how that relates to e.g. the CLARIN switchboard, as some of the
+questions indicate we have not communicated this clearly enough.
+
+I'd like to address each of the questions here:
 
@@
 meta-distribution that it is hard to pin down for any single scenario, but it would be interesting for the services
-inside a LaMachine installation.  I opened issue https://github.com/proycon/LaMachine/issues/138 to this end. It's not
+inside **a** LaMachine installation (our own public one perhaps).  I opened issue https://github.com/proycon/LaMachine/issues/138 to this end. It's not
 in the paper as it does not offer sufficient space for such a discussion.
@@
 
-The "transcends" remark in our paper applies to the VRE project rather than to LaMachine, I clarified it to prevent
+The "transcends" remark in our paper applies to the VRE project rather than to LaMachine, I now clarified it to prevent
 confusion. LaMachine is not directly comparable with the switchboard as it offers no switchboard functionality. Now, the
@@
 include a simpler portal instead as the switchboard did not meet all our demands and other initiatives were started to
-compete with it (CLARIAH WP3 VRE). Tighter compatibility between the switchboard and services offered by a LaMachine
+compete with it (CLARIAH WP3 VRE). Tighter compatibility between the switchboard and services offered by **a** LaMachine
 installation would be possible, but this is more a question for the switchboard developers than for us.
@@
 LaMachine itself is a distribution so it doesn't process any data as such; that would be a question for particular tools
-or frameworks and would not be something we as a distribution would enforce either.
+or frameworks within it, and would not be something we as a distribution would enforce.
 

### 2019-04-01 docs/lamachine_clarin2018.tex
typo and double ref in clarin paper

@@
 Within the scope of CLARIN, one might be inclined to draw comparisons with the CLARIN Language Resource Switchboard
-\cite{zinn2016clarin}, which offers access to NLP tools through a web interface where users can upload their texts for
+\cite{zinn2016}, which offers access to NLP tools through a web interface where users can upload their texts for
 NLP processing, providing easy access and automated tool discovery. This, however, is notably different from what
@@
 
-There are also other CLARIN initiatives that provide not just a switchboard, but aim at an eve more ambitious high-level virtual research
+There are also other CLARIN initiatives that provide not just a switchboard, but aim at an even more ambitious high-level virtual research
 environment (VRE) with graphical interfaces to accommodate the researcher and allow execution and visualisation of
